### 2022-03-24 draft.tex
Deploy to GitHub Pages

%!TEX TS-program = xelatex
\documentclass[11pt]{article}

\usepackage[english]{babel}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{stix}
\usepackage[scaled]{helvet}
\usepackage[scaled]{inconsolata}

\usepackage{lastpage}

\usepackage{setspace}

\usepackage{ccicons}

\usepackage[hang,flushmargin]{footmisc}

\usepackage{geometry}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}

\usepackage{fancyhdr}
\renewcommand{\headrulewidth}{0pt}\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\makeatletter
\newcounter{tableno}
\newenvironment{tablenos:no-prefix-table-caption}{
  \caption@ifcompatibility{}{
    \let\oldthetable\thetable
    \let\oldtheHtable\theHtable
    \renewcommand{\thetable}{tableno:\thetableno}
    \renewcommand{\theHtable}{tableno:\thetableno}
    \stepcounter{tableno}
    \captionsetup{labelformat=empty}
  }
}{
  \caption@ifcompatibility{}{
    \captionsetup{labelformat=default}
    \let\thetable\oldthetable
    \let\theHtable\oldtheHtable
    \addtocounter{table}{-1}
  }
}
\makeatother

\usepackage{array}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\let\PBS=\PreserveBackslash

\usepackage[breaklinks=true]{hyperref}
\hypersetup{colorlinks,%
citecolor=blue,%
filecolor=blue,%
linkcolor=blue,%
urlcolor=blue}
\usepackage{url}

\usepackage{caption}
\setcounter{secnumdepth}{0}
\usepackage{cleveref}

\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\usepackage{longtable}
\usepackage{booktabs}

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for \widthof, \maxof
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=4.5cm}

\usepackage{lineno}
\usepackage[nolists,noheads]{endfloat}

\pagestyle{plain}

\doublespacing

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\thispagestyle{empty}
{\Large\bfseries\sffamily Maximum entropy modelling of food-web
structure}
\vskip 5em

%
\href{https://orcid.org/0000-0001-9051-0597}{Francis\,Banville}%
%
\,\textsuperscript{1,2,3}\quad %
\href{https://orcid.org/0000-0002-4498-7076}{Dominique\,Gravel}%
%
\,\textsuperscript{2,3}\quad %
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,3}

\textsuperscript{1}\,Université de
Montréal\quad \textsuperscript{2}\,Université de
Sherbrooke\quad \textsuperscript{3}\,Quebec Centre for Biodiversity
Science

\textbf{Correspondance to:}\\
Francis Banville --- \texttt{francis.banville@umontreal.ca}\\

\vfill
This work is released by its authors under a CC-BY 4.0 license\hfill\ccby\\
Last revision: \emph{\today}

\clearpage
\thispagestyle{empty}

\vfill
\textbf{\sffamily Abstract: }The principle of maximum entropy is a
rigorous mathematical method of finding constrained probability
distributions that has been proven useful in many ecological modelling
problems. However, despite its broad application in graph and network
theory, it has seldom been used to model ecological networks. Here we
show how the (joint) degree distribution of maximum entropy can be
directly derived using the number of species and the number of links in
terrestrial and aquatic food webs. We also present a heuristic and
flexible approach of finding the network of maximum entropy based on
simulating annealing and SVD entropy. We built two of these
network-level models using constraints given by the connectance and the
joint degree sequence, respectively. All maximum entropy models were
compared against open access food-web data and null and neutral models
commonly used in network ecology. We found that the maximum entropy
network model constrained by the joint degree sequence was a good
predictor of food-web structure, including nestedness and motifs
distribution. Overall, our results suggest that many properties of
ecological networks are mainly driven by the joint degree distribution
and statistical phenomena.
\vfill

\clearpage
\linenumbers
\pagestyle{normal}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Statistical and mathematical models can help fill many gaps in our
knowledge about species interactions. Two complementary types of models
have been developed in network ecology for this purpose. On one hand,
predictive models can partially alleviate the Eltonian shortfall, which
describes our current lack of knowledge on food webs and other
ecological networks (Hortal et al. 2015). A variety of such models have
recently been developed using machine learning and other statistical
tools, most of which are presented in Strydom et al. (2021). On the
other hand, null models help us identify potential ecological mechanisms
that drive species interactions. They do so by comparing empirical data
with an unbiased distribution of measures generated using a set of rules
that exclude the mechanism of interest (M. A. Fortuna and Bascompte
2006; Delmas et al. 2019). Both types of models are frequently
topological, i.e.~they often predict the adjacency matrix or specific
measures of network structure without taking into account species'
identity. According to Strydom et al. (2021), these topological models
could be used to make better predictions of pairwise species
interactions by constraining the space of feasible networks.

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. This mathematical method, briefly presented
in Box 1, has been used in a wide range of disciplines, from
thermodynamics to chemistry and biology (Martyushev and Seleznev 2006).
It has also been proven useful in ecology, e.g.~in species distribution
models (Phillips, Anderson, and Schapire 2006) and macroecological
models (J. Harte et al. 2008; John Harte and Newman 2014). As discussed
in Box 1, maximizing a measure of entropy ensures that the derived
probability distributions are unique and least biased under the set of
constraints used. These constraints are built using state variables,
i.e.~variables that represent the macrostate of the system. The
challenge is to find the set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make rigorous predictions using a minimal amount of
data, as well as helping us describe the most important factors driving
that system.

Despite its extensive use in graph and network theories (e.g., Park and
Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
comparison been little used in network ecology. The very nature of
ecological networks (directed simple graphs frequently having
self-loops) makes the mathematical optimization of maximum entropy graph
models more complicated than with many other types of (non-ecological)
networks. MaxEnt has nevertheless been used to predict the degree
distribution of bipartite ecological networks from the number of species
and the number of interactions (Williams 2011) and to predict
interaction strengths between species pairs using their relative
abundances within an optimal transportation theory regularized with
information entropy (Stock, Poisot, and De Baets 2021). However, to the
best of our knowledge, MaxEnt has never been used to predict food-web
structure directly, even though food webs are among the most documented
and widespread ecological networks.

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of links in ecological networks, in Box 2 we present a method
to predict \(L\) from \(S\), thus allowing the prediction of the joint
degree distribution from \(S\) solely. In turn, the second approach
consists in finding, using different constraints, the adjacency matrix
of maximum entropy from which network properties can be measured. To do
so, we used a flexible and heuristic approach based on simulating
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e., simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e., the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

\hypertarget{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
\section{Box 1 - The principle of maximum entropy: A primer for
ecologists}\label{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}}

The principle of maximum entropy is a mathematical method of finding
probability distributions, strongly rooted in statistical mechanics and
information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
Starting from a set of constraints given by prior knowledge of a system
(i.e., what we call state variables), this method helps us find
least-biased probability distributions subject to the constraints. These
probability distributions are guaranteed to be unique given our prior
knowledge and represent the most we can say about a system without
making more assumptions. For example, if the only thing we know about a
biological community is its average number of individuals per species,
the least-biased inference we could make on its species abundance
distribution is the exponential distribution (Frank and Smith 2011; John
Harte and Newman 2014). However, this does not imply that this
distribution will be the best fit to empirical data. The challenge is to
find the right set of constraints that would best reproduce
distributions found in nature.

Entropy measures the amount of information given by the outcome of a
random variable. Many measures of entropy have been developed in physics
(Beck 2009), but only a fraction of them could be used as an
optimization measure with the principle of maximum entropy. According to
Beck (2009) and Khinchin (2013), a measure of entropy \(H\) should
satisfy four properties in the discrete case: (1) it should be a
function of a probability distribution \(p(n)\) only; (2) it should be
maximized when \(p(n)\) is uniform; (3) it should not be influenced by
outcomes with a null probability; and (4) it should be independent of
the order of information acquisition. The Shannon's entropy (Shannon
1948)

\begin{equation}\protect\hypertarget{eq:shannon}{}{H = -\sum_{n} p(n) \log p(n)}\label{eq:shannon}\end{equation}

satisfies all of these properties. Finding the probability distribution
\(p(n)\) that maximizes \(H\) under a set of \(m\) constraints \(g\) can
be done using the method of Lagrange multipliers. These constraints
could include one or many properties of the probability distribution
(e.g., its mean, variance, and range). However, the normalization
constraint always need to be included in \(g\) in order to make sure
that \(p(n)\) sums to \(1\). The objective is then to find the values of
the Lagrange multipliers \(\lambda_i\) that optimize a function \(F\):

\begin{equation}\protect\hypertarget{eq:F}{}{F = H - \sum_{i=1}^m \lambda_i (g_i-c_i),}\label{eq:F}\end{equation}

where \(g_i\) is the mathematical formulation of the constraint \(i\)
and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
maximized by setting to \(0\) its partial derivative with respect to
\(p(n)\). We will show how this can be done when we derive the joint
degree distribution analytically from the number of species and the
number of links in food webs.

In this contribution, we also use the SVD entropy as a measure of
entropy, which is an application of Shannon's entropy to the relative
non-zero singular values of a truncated singular value decomposition
(t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a food web's Boolean
adjacency matrix. This measure also satisfies all four properties
above-mentioned, while being a proper measure of the internal complexity
of food webs (Strydom, Dalla Riva, and Poisot 2021). We measured SVD
entropy as follows:

\begin{equation}\protect\hypertarget{eq:svd-entropy}{}{J = -\sum_{i=1}^R s_i \log s_i,}\label{eq:svd-entropy}\end{equation}

where \(s_i\) are the relative singular values
(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
the singular values). Following Strydom, Dalla Riva, and Poisot (2021),
we standardized this measure with the rank \(R\) of the matrix (i.e.,
\(J / \ln(R)\)) to account for the difference in dimensions between
networks (Pielou's evenness; Pielou 1975). In a following section, we
will show how SVD entropy can be used to predict a network of maximum
entropy (i.e., of maximum complexity) heuristically.

\hypertarget{testing-maxent-models}{%
\section{Testing MaxEnt models}\label{testing-maxent-models}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

We tested our MaxEnt models (both approaches) against open food-web data
queried from three different sources and integrated into what we call
our \emph{complete dataset}. First, all food webs archived on
\texttt{mangal.io} were directly queried from the database
(\(n = 235\)). Most ecological networks archived on Mangal are
multilayer networks, i.e.~networks that describe different types of
interactions. We kept all networks whose interactions were mainly of
predation and herbivory types, and removed the largest network
(\(S = 714\)) for computational efficiency reasons. Then, to this set we
added food webs from two different sources: the New-Zealand dataset
(\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
\(19\) networks had data on species' relative abundances that were used
in the neutral model presented in a following subsection. These networks
are part of what we call our \emph{abundance dataset}, which is a subset
of our complete dataset.

All code and data to reproduce this article are available at the Open
Science Framework. Data cleaning, simulations and analyses were
conducted in Julia v1.5.4.

\hypertarget{null-models-types-i-and-ii}{%
\subsection{Null models (types I and
II)}\label{null-models-types-i-and-ii}}

Our maximum entropy network models (second approach only) were compared
with two topological null models. The first is the type I null model of
M. A. Fortuna and Bascompte (2006), in which the probability that a
species \(i\) predates on another species \(j\) is given by

\begin{equation}\protect\hypertarget{eq:type1null}{}{p(i \rightarrow j) = \frac{L}{S^2}.}\label{eq:type1null}\end{equation}

The second is the type II null model of Bascompte et al. (2003), in
which the probability of interaction is instead given by

\begin{equation}\protect\hypertarget{eq:type2null}{}{p(i \rightarrow j) = \frac{1}{2} \left(\frac{k_{in}(j)}{S} + \frac{k_{out}(i)}{S}\right),}\label{eq:type2null}\end{equation}

where \(k_{in}\) and \(k_{out}\) are the in and out-degrees,
respectively. The type I null model is based on connectance, whereas the
type II null model is based on the joint degree sequence. Therefore, the
type I and II topological null models correspond with our type I and II
MaxEnt network models, respectively, since they use similar constraints.

We predicted both types of null networks for all empirical networks in
our complete dataset (\(n = 257\)). We converted all probabilistic
networks to Boolean networks by generating \(100\) random Boolean
networks for each of these probabilistic webs. Then, we counted the
number of times each interaction was sampled, and kept the \(L\) entries
that were drawn the most amount of time, with \(L\) given by the number
of links in each food web. This ensured that the resulting null networks
had the same number of interactions as their empirical counterparts.

\hypertarget{neutral-model}{%
\subsection{Neutral model}\label{neutral-model}}

We also compared our MaxEnt network models with a neutral model of
relative abundances, in which the probabilities of interaction are given
by

\begin{equation}\protect\hypertarget{eq:neutralmodel}{}{p(i \rightarrow j) \propto \frac{n_i}{N} \times \frac{n_j}{N},}\label{eq:neutralmodel}\end{equation}

where \(n_i\) and \(n_j\) are the abundances (or biomass) of both
species, and \(N\) is the total abundance (or biomass) of all species in
the network. We predicted neutral abundance matrices for all empirical
networks in our abundance dataset (\(n = 19\)), and converted these
weighted matrices to Boolean networks using an approach analogue to the
one we used for our null models.

\hypertarget{analytical-models-measures-of-maximum-entropy}{%
\section{Analytical models: Measures of maximum
entropy}\label{analytical-models-measures-of-maximum-entropy}}

\hypertarget{joint-degree-distribution}{%
\subsection{Joint degree distribution}\label{joint-degree-distribution}}

The joint degree distribution \(p(k_{in},k_{out})\) is a joint discrete
probability distribution describing the probability that a species has
\(k_{in}\) predators and \(k_{out}\) preys, with \(k_{in}\) and
\(k_{out}\) \(\epsilon\) \([0, S]\). Basal species (e.g., plants) have a
\(k_{out}\) of \(0\), whereas top predators have a \(k_{in}\) of \(0\).
In contrast, the maximum number of preys and predators a species can
have is set by the number of species \(S\) in the food web. Here we show
how the joint degree distribution of maximum entropy can be obtained
given knowledge of \(S\) and \(L\).

We want to maximize Shannon's entropy

\begin{equation}\protect\hypertarget{eq:entropy_jdd}{}{H = -\sum_{k_{in}=0}^S\sum_{k_{out}=0}^S p(k_{in},k_{out}) \log p(k_{in},k_{out})}\label{eq:entropy_jdd}\end{equation}

subject to the following constraints:

\begin{equation}\protect\hypertarget{eq:g1}{}{g_1 = \sum_{k_{in}=0}^S\sum_{k_{out}=0}^S p(k_{in},k_{out}) = 1;}\label{eq:g1}\end{equation}

\begin{equation}\protect\hypertarget{eq:g2}{}{g_2 = \sum_{k_{in}=0}^S\sum_{k_{out}=0}^S k_{in} p(k_{in},k_{out}) = \langle k_{in} \rangle = \frac{L}{S};}\label{eq:g2}\end{equation}

\begin{equation}\protect\hypertarget{eq:g3}{}{g_3 = \sum_{k_{in}=0}^S\sum_{k_{out}=0}^S k_{out} p(k_{in},k_{out}) = \langle k_{out} \rangle = \frac{L}{S}.}\label{eq:g3}\end{equation}

The first constraint \(g_1\) is our normalizing constraint, whereas the
other two (\(g_2\) and \(g_3\)) fix the average of the marginal
distributions of \(k_{in}\) and \(k_{out}\) to the linkage density
\(L/S\). It is important to notice that
\(\langle k_{in} \rangle = \langle k_{out} \rangle\) because every edge
is associated to a predator and a prey. Therefore, without any further
constraints, we expect the joint degree distribution of maximum entropy
to be a symmetric probability distribution with regards to \(k_{in}\)
and \(k_{out}\). However, this does not mean that the joint degree
\emph{sequence} will be symmetric, since the joint degree sequence is
essentially a random realization of its probabilistic counterpart.

The joint probability distribution of maximum entropy given these
constraints is found using the method of Lagrange multipliers. To do so,
we seek to maximize the following expression:

\begin{equation}\protect\hypertarget{eq:F_jdd}{}{F = H - \lambda_1(g_1-1)-\lambda_2\left( g_2-\frac{L}{S}\right) - \lambda_3 \left( g_3-\frac{L}{S}\right),}\label{eq:F_jdd}\end{equation}

where \(\lambda_1\), \(\lambda_2\), and \(\lambda_3\) are the Lagrange
multipliers. The probability distribution that maximizes entropy is
obtained by finding these values. As pointed out in Box 1, \(F\) is just
Shannon's entropy to which we added terms that each sums to zero (our
constraints). \(F\) is maximized by setting to 0 its partial derivative
with respect to \(p(k_{in},k_{out})\). Because the derivative of a
constant is zero, this gives us:

\begin{equation}\protect\hypertarget{eq:lagrange_jdd}{}{\frac{\partial H}{\partial p(k_{in},k_{out})} = \lambda_1 \frac{\partial g_1}{\partial p(k_{in},k_{out})} + \lambda_2 \frac{\partial g_2}{\partial p(k_{in},k_{out})}+ \lambda_3 \frac{\partial g_3}{\partial p(k_{in},k_{out})}.}\label{eq:lagrange_jdd}\end{equation}

Evaluating the partial derivatives with respect to
\(p(k_{in},k_{out})\), we obtain:

\begin{equation}\protect\hypertarget{eq:lagrange2_jdd}{}{-\log p(k_{in},k_{out}) - 1 = \lambda_1 + \lambda_2 k_{in} + \lambda_3 k_{out}.}\label{eq:lagrange2_jdd}\end{equation}

Then, solving eq.~\ref{eq:lagrange2_jdd} for \(p(k_{in},k_{out})\), we
obtain:

\begin{equation}\protect\hypertarget{eq:lagrange3_jdd}{}{p(k_{in},k_{out}) = \frac{e^{-\lambda_2k_{in}-\lambda_3k_{out}}}{Z},}\label{eq:lagrange3_jdd}\end{equation}

where \(Z = e^{1+\lambda_1}\) is called the partition function. The
partition function ensures that probabilities sum to 1 (our
normalization constraint). It can be expressed in terms of \(\lambda_2\)
and \(\lambda_3\) as follows:

\begin{equation}\protect\hypertarget{eq:Z}{}{Z = \sum_{k_{in}=0}^S\sum_{k_{out}=0}^S e^{-\lambda_2k_{in}-\lambda_3k_{out}}.}\label{eq:Z}\end{equation}

After substituting \(p(k_{in},k_{out})\) in eq.~\ref{eq:g2} and
eq.~\ref{eq:g3}, we get a nonlinear system of two equations and two
unknowns:

\begin{equation}\protect\hypertarget{eq:lagrange4_jdd}{}{\frac{1}{Z}\sum_{k_{in}=0}^S\sum_{k_{out}=0}^S k_{in} e^{-\lambda_2k_{in}-\lambda_3k_{out}}  = \frac{L}{S};}\label{eq:lagrange4_jdd}\end{equation}

\begin{equation}\protect\hypertarget{eq:lagrange5_jdd}{}{\frac{1}{Z}\sum_{k_{in}=0}^S\sum_{k_{out}=0}^S k_{out} e^{-\lambda_2k_{in}-\lambda_3k_{out}}  = \frac{L}{S}.}\label{eq:lagrange5_jdd}\end{equation}

We solved eq.~\ref{eq:lagrange4_jdd} and eq.~\ref{eq:lagrange5_jdd}
numerically using the Julia library \texttt{JuMP.jl} v0.21.8 (Dunning,
Huchette, and Lubin 2017). \texttt{JuMP.jl} supports nonlinear
optimization problems by providing exact second derivatives that
increase the accuracy and performance of its solvers. The estimated
values of \(\lambda_2\) and \(\lambda_3\) can be substituted in
eq.~\ref{eq:lagrange3_jdd} to have a more workable expression for the
joint degree distribution.

We predicted the joint degree distribution of maximum entropy for each
food web in our complete dataset, i.e.~using their numbers of species
and numbers of links as state variables. We then sampled one realization
of the degree sequence for each network using the probabilities given by
the joint degree distribution. In fig.~\ref{fig:joint_dd} (left panels),
we show the relationship between \(k_{out}\) and \(k_{in}\) standardized
by the number of species in their networks, for empirical and maximum
entropy joint degree distributions. We see that our model predicts a
greater number of generalist species than empirical food webs (i.e.,
relative values of \(k_{out}\) and \(k_{in}\) closer to \(1\)). However,
plotting the difference between predicted and empirical values for each
species gives a different perspective. The right panel of
fig.~\ref{fig:joint_dd} presents these differences when species are
ordered by their total degree in their network (i.e., by the sum of
their in and out-degrees). Indeed, our predicted joint degree sequences
have the same number of species as their empirical counterparts, but
they are species agnostic; in other words, instead of predicting a value
for each species directly, we predicted the entire joint degree sequence
without taking into account species' identity. When we associate
predictions and empirical data according to their rank in total degrees,
we see that species predicted to be have a higher generality (number of
preys) generally have a lower vulnerability (number of preys) than what
is observed (and conversely). In fig.~S1, we show how these differences
change when species are instead ordered by their out-degrees (left
panel) and in-degrees (right panel), respectively.

\begin{figure}
\hypertarget{fig:joint_dd}{%
\centering
\includegraphics{figures/joint_degree_dist.png}
\caption{Relative number of predators (\(k_{in}\)) as a function of the
relative number of preys (\(k_{out}\)) for each species in (a) empirical
and (b) maximum entropy joint degree sequences. Empirical networks
include all food webs archived on Mangal, as well as the New-Zealand and
Tuesday lake datasets (our complete dataset). The predicted joint degree
sequences were obtained after sampling one realization of the joint
degree distribution of maximum entropy for each network. (c) Difference
between predicted and empirical values when species are ordered
according to their total degrees. In all panels, each dot corresponds to
a single species in one of the network of our complete
dataset.}\label{fig:joint_dd}
}
\end{figure}

We plotted the Kullback--Leibler divergence between in and out-degree
distributions to compare the symmetry of empirical and maximum entropy
joint degree sequences fig.~\ref{fig:kl_diverg}. As we expected, our
model predicted more similar in-degree and out-degree distributions than
empirical data. However, this difference decreased with connectance
right panel of \ref{fig:kl_diverg}. Overall, this suggests that other
ecological constraints might be needed to account for the asymmetry of
the joint degree distribution, especially for networks with a lower
connectance.

\begin{figure}
\hypertarget{fig:kl_diverg}{%
\centering
\includegraphics{figures/kl_divergence.png}
\caption{(a) Probability density of KL divergence between in and
out-degree sequences of empirical and maximum entropy joint degree
sequences. (b) Difference between the KL divergence of empirical and
predicted networks as a function of connectance. In both panels,
empirical networks include all food webs archived on Mangal, as well as
the New-Zealand and Tuesday lake datasets (our complete dataset). The
joint degree sequence of simulated networks was obtained after sampling
one realization of the joint degree distribution of maximum entropy for
each network.}\label{fig:kl_diverg}
}
\end{figure}

\hypertarget{degree-distribution}{%
\subsection{Degree distribution}\label{degree-distribution}}

The degree distribution \(p(k)\) represents the probability that a
species has \(k\) links in a food web, with \(k = k_{in} + k_{out}\). It
can thus be directly obtained from the joint degree distribution:

\[p(k) = \sum_{i=0}^k p(k_{in} = k - i, k_{out} = i).\]

In fig.~S2, we show that the degree distribution of maximum entropy,
given \(S\) and \(L\), predicts very low probabilities that a species
will be isolated in its food web (\emph{i.e.}, having \(k = 0\)). As
MacDonald, Banville, and Poisot (2020) pointed out, the size of food
webs should at least be of \(S-1\) links, since a lower number would
yield isolated species, i.e.~species without any predators or preys. Our
results show that, under our purely information-theoretic model, the
probability that a species is isolated is quite high below this
threshold. The expected proportion of isolated species rapidly declines
by orders of magnitude with increasing numbers of species and links.

The degree distribution could also have been obtained directly using the
principle of maximum entropy, as discussed in Williams (2011). This
gives the following distribution:

\begin{equation}\protect\hypertarget{eq:lagrange_dd}{}{p(k) = \frac{e^{-\lambda_2k}}{Z},}\label{eq:lagrange_dd}\end{equation}

with \(Z = \sum_{k=0}^S e^{-\lambda_2k}.\)

This can be solved numerically using the constraint of the average
degree \(\langle k \rangle = \frac{2L}{S}\) of a species. Note that the
mean degree is twice the value of the linkage density, because every
link must be counted twice when we add in and out-degrees together.

\begin{equation}\protect\hypertarget{eq:lagrange2_dd}{}{\frac{1}{Z}\sum_{k=0}^S k e^{-\lambda_2k} = \frac{2L}{S}}\label{eq:lagrange2_dd}\end{equation}

The numerical solution is identical to the one we obtained using the
joint degree distribution as an intermediate. Ecologists wanting to
model a system without considering isolated species could simply change
the lower limit of \(k\) to \(1\) and solve the resulting equation
numerically.

In this section, we showed how important measures of food-web structure,
namely the degree distribution and the joint degree distribution, could
be derived with the principle of maximum entropy using minimal knowledge
on a biological community. This type of models, although useful to make
least-biased predictions on many network properties, can be hard to
apply for other measures. Indeed, there are dozens of measures of
network structure (Delmas et al. 2019) and many are not calculated with
mathematical equations, but with algorithms. Moreover, the applicability
of this method to empirical systems is limited by the state variables we
can actually measure and use. In the next section, we propose a more
flexible method to predict many measures of network structure
simultaneously, i.e.~by finding networks of maximum entropy
heuristically.

\hypertarget{box-2---working-with-predicted-numbers-of-links}{%
\section{Box 2 - Working with predicted numbers of
links}\label{box-2---working-with-predicted-numbers-of-links}}

Our models need information on the number of species and the number of
links. However, since the later is rarely estimated empirically,
ecologists might need to use predictive methods to estimate the total
number of links in a food web.

We used the flexible links model of MacDonald, Banville, and Poisot
(2020) to predict the number of interactions from the number of species.
The flexible links model, in contrast to other predictive models of the
number of links, incorporates meaningful ecological constraints into the
prediction of \(L\), namely the minimum \(S-1\) and maximum \(S^2\)
numbers of interactions in food webs. It estimates the proportion of the
\(S^2 - (S - 1)\) \emph{flexible links} that are realized. More
precisely, this model states that the number of \emph{realized} flexible
links \(L_{FL}\) in a food web represents the number of realized
interactions above the minimum (i.e., \(L = L_{FL} + S - 1\)) and is
obtained from a beta-binomial distribution with \(S^2 - (S - 1)\) trials
and parameters \(\alpha = \mu e^\phi\) and \(\beta = (1 - \mu) e^\phi\):

\begin{equation}\protect\hypertarget{eq:BB}{}{L_{FL} \sim \mathrm{BB}(S^2 - (S - 1), \mu e^\phi, (1 - \mu) e^\phi),}\label{eq:BB}\end{equation}

where \(\mu\) is the average probability across food webs that a
flexible link is realized, and \(\phi\) is the concentration parameter
around \(\mu\).

We fitted the flexible links model on all food webs in our complete
dataset, and estimated the parameters of eq.~\ref{eq:BB} using a
Hamiltonian Monte Carlo sampler with static trajectory (\(1\) chain and
\(3000\) iterations):

\begin{equation}\protect\hypertarget{eq:BBpost}{}{ [\mu, \phi| \textbf{L}, \textbf{S}] \propto \prod_{i = 1}^{m} \mathrm{BB}(L_i - (S_i - 1) | S_i^2 - (S_i - 1)), \mu e^{\phi}, (1 - \mu) e^\phi) \times \mathrm{B}(\mu| 3 , 7 ) \times \mathcal{N}(\phi | 3, 0.5), }\label{eq:BBpost}\end{equation}

where \(m\) is the number of food webs (\(m = 257\)) and \(\textbf{L}\)
and \(\textbf{S}\) are respectively the vectors of their numbers of
interactions and numbers of species. Our weakly-informative prior
distributions were chosen following MacDonald, Banville, and Poisot
(2020), i.e.~a beta distribution for \(\mu\) and a normal distribution
for \(\phi\). The Monte Carlo sampling of the posterior distribution was
conducted using the Julia library \texttt{Turing} v0.15.12.

The flexible links model is a generative model, i.e.~it can generate
plausible values of the predicted variable. We thus simulated \(1000\)
values of \(L\) for different values of \(S\) using the joint posterior
distribution of our model parameters, and calculated the mean degree for
each simulated values. The resulting distributions are shown in the left
panel of fig.~\ref{fig:degree_dist_fl} for three different values of
species richness. In the right panel of fig.~\ref{fig:degree_dist_fl},
we show how the probability distribution for the mean degree constraints
can be used to generate a distribution of maximum entropy degree
distributions, since each simulated value of mean degree generates a
different maximum entropy degree distribution.

\begin{figure}
\hypertarget{fig:degree_dist_fl}{%
\centering
\includegraphics{figures/maxent_degree_dist_fl.png}
\caption{(a) Probability density of the predicted mean degree for
different values of species richness. The number of links was predicted
using the flexible links model fitted to all empirical networks in our
complete dataset. (b) Degree distributions of maximum entropy for a
network of \(27\) species and different numbers of links. The numbers of
links correspond to the lower and upper bounds of the \(67%
\), \(89%
\), and \(97%
\) percentile intervals (PI), as well as the median, of the
counterfactuals of the flexible links model.}\label{fig:degree_dist_fl}
}
\end{figure}

\hypertarget{heuristical-models-networks-of-maximum-entropy}{%
\section{Heuristical models: Networks of maximum
entropy}\label{heuristical-models-networks-of-maximum-entropy}}

\hypertarget{maxent-network-models-types-i-and-ii}{%
\subsection{MaxEnt network models (types I and
II)}\label{maxent-network-models-types-i-and-ii}}

We define networks of maximum entropy as the configuration of the
adjacency matrix with the highest SVD entropy under a set of
constraints. As mentioned in Box 1, we used the SVD entropy as our
measure of entropy since it has been shown to be a reliable measure of
food-web complexity (Strydom, Dalla Riva, and Poisot 2021), in addition
to having the required properties of a proper measure of information
entropy. We thus seek to find the network with the highest complexity,
or randomness, that exactly reproduces specified constraints on its
structure. Our method is in contrast with maximum entropy graph models
that predict a probability distribution on networks under soft or hard
constraints Cimini et al. (2019). We believe our approach to be more
flexible, easier to compute, while allowing direct comparisons of
empirical food webs with more complex networks with similar structure.

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulating annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains after burn-in. For each network in our complete
and abundance datasets, we estimated their configuration with maximum
entropy using both types of MaxEnt network models.

\hypertarget{structure-of-maxent-networks}{%
\subsection{Structure of MaxEnt
networks}\label{structure-of-maxent-networks}}

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Miguel A. Fortuna et al. 2010). We also measured the maximum
trophic level \emph{maxtl}, network diameter \emph{diam} (i.e., the
longest of the shortest paths between all species pairs; Albert and
Barabasi 2002), the average maximum similarity between species pairs
\emph{MxSim}, the proportion of cannibal species \emph{Cannib} (i.e.,
the proportion of self loops), and the proportion of omnivorous species
\emph{Omniv} (i.e., species whose preys are of different trophic
levels). \emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply
defined in Williams and Martinez (2000).

Overall, we found that models based on the joint degree sequence (type
II null and MaxEnt network models) reproduced the structure of empirical
networks much better than the ones based on connectance and the neutral
model of relative abundances. On average, the type II MaxEnt network
model was better at predicting nestedness (\(0.62 \pm 0.08\)) than its
corresponding null model (\(0.73 \pm 0.05\); empirical networks:
\(0.63 \pm 0.09\)), as well as the proportion of cannibal species.
However, the type II null model was better at predicting network
diameter and average maximum similarity between species pairs.
Predictions were similar between both type II models for the maximum
trophic level and the proportion of omnivorous species.

\hypertarget{tbl:measures_all}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_all}Standardized mean differences of
predicted network measures with empirical data for all networks
(\(n = 257\)). Positive (negative) values indicate that the measure is
overestimated (underestimated) on average. Empirical networks are all
food webs archived on Mangal, as well as the New Zealand and Tuesday
lake food webs (our complete dataset). Null 1: Type I null model based
on connectance. MaxEnt 1: Type I MaxEnt network model based on
connectance. Null 2: Type II null model based on the joint degree
sequence. MaxEnt 2: Type II MaxEnt network model based on the joint
degree sequence. \(\rho\): nestedness measured by the spectral radius of
the adjacency matrix. \emph{maxtl}: maximum trophic level. \emph{diam}:
network diameter. \emph{MxSim}: average maximum similarity between
species pairs. \emph{Cannib}: proportion of cannibal species (self
loops). \emph{Omniv}: proportion of omnivorous species. \emph{entropy}:
SVD entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
null 1 & -0.167 & 0.980 & 1.428 & -0.502 & 2.007 & 1.493 &
0.056\tabularnewline
MaxEnt 1 & -0.226 & 0.831 & 1.274 & -0.524 & 1.982 & 1.863 &
0.106\tabularnewline
null 2 & 0.160 & -0.125 & 0.016 & 0.007 & 1.078 & 0.559 &
-0.023\tabularnewline
MaxEnt 2 & -0.015 & 0.178 & 0.565 & -0.282 & 0.698 & 0.589 &
0.058\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{tbl:measures_abund}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_abund}Standardized mean differences of
predicted network measures with empirical data for networks having
abundance data (\(n = 19\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks are all New Zealand and Tuesday lake food webs with abundance
data (our abundance dataset). Null 1: Type I null model based on
connectance. MaxEnt 1: Type I MaxEnt network model based on connectance.
Null 2: Type II null model based on the joint degree sequence. MaxEnt 2:
Type II MaxEnt network model based on the joint degree sequence.
\(\rho\): nestedness measured by the spectral radius of the adjacency
matrix. \emph{maxtl}: maximum trophic level. \emph{diam}: network
diameter. \emph{MxSim}: average maximum similarity between species
pairs. \emph{Cannib}: proportion of cannibal species (self loops).
\emph{Omniv}: proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
neutral & 0.367 & -0.090 & 0.027 & 0.266 & 6.870 & 0.576 &
-0.083\tabularnewline
null 1 & -0.134 & 0.950 & 1.919 & -0.369 & 2.077 & 0.614 &
0.068\tabularnewline
MaxEnt 1 & -0.229 & 1.020 & 1.946 & -0.355 & 2.215 & 0.801 &
0.121\tabularnewline
null 2 & 0.128 & -0.115 & -0.135 & 0.157 & 1.444 & 0.029 &
-0.021\tabularnewline
MaxEnt 2 & -0.010 & 0.054 & 0.243 & -0.062 & -0.038 & 0.083 &
0.038\tabularnewline
\bottomrule
\end{longtable}

The picture slightly changes when we consider another important property
of ecological networks, i.e.~their motifs profile. We measured the
proportion of three-species motifs, which can be considered as simple
building blocks of ecological networks (Milo et al. 2002; Stouffer et
al. 2007). Motifs are the backbone of complex ecological networks from
which network structure is built upon and play a crucial role in
community dynamics and assembly (Bastolla et al. 2009; Stouffer and
Bascompte 2011). In fig.~\ref{fig:motifs}, we show that the motifs
profile of networks generated using the type II MaxEnt network model was
very close to the one of empirical data, and that they made better
predictions than the type II null model. This is also shown in
fig.~\ref{fig:motifs_rel}, where we see that relationships between
motifs pairs of empirical food webs are very similar to the ones of the
type II MaxEnt networks. Overall, these results suggest that the type II
MaxEnt network model can reproduce many aspects of network structure.
This highlights the importance of the joint degree sequence in shaping
ecological networks.

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include all food webs archived on Mangal, as
well as the New-Zealand and Tuesday lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulating
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include all food webs
archived on Mangal, as well as the New-Zealand and Tuesday lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulating annealing algorithm to find the network of maximum SVD
entropy while maintaining the connectance (type I MaxEnt network model)
or the joint degree sequence (type II MaxEnt network model). The
predictions of the types I and II null models are also plotted.
Regression lines are plotted in each panel. Motifs names are from
Stouffer et al. (2007).}\label{fig:motifs_rel}
}
\end{figure}

Moreover, we found that empirical networks are close to their maximum
entropy considering a fixed joint degree sequence (fig.~S3). Empirical
networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
between the SVD entropy of empirical food webs and their maximum entropy
configuration is plotted in the last panel of fig.~\ref{fig:measures}.
We also found no relationship between the difference in SVD entropy
between empirical and type II MaxEnt networks and species richness, the
number of links, or connectance (fig.~S4). Similarly, we found no
correlation between the difference in SVD entropy and the Jaccard
distance of empirical and type II MaxEnt networks.

Finally, in fig.~\ref{fig:measures} we also show how well empirical
measures are predicted using the type II MaxEnt network model. In
accordance with our previous results, we found that nestedness was very
well predicted by our model. However, the type II MaxEnt network model
overestimated the maximum trophic level and network diameter, especially
for networks with high empirical values. In fig.~S5, we show that the
pairwise relationships between these four measures and species richness
in empirical food webs are similar (in magnitude and sign) to the ones
in type II MaxEnt networks.

\begin{figure}
\hypertarget{fig:measures}{%
\centering
\includegraphics{figures/measures_emp_maxent.png}
\caption{Relationship between the structure of empirical and maximum
entropy food webs. Empirical networks include all food webs archived on
Mangal, as well as the New-Zealand and Tuesday lake datasets (our
complete dataset). Maximum entropy networks were estimated using a
simulating annealing algorithm to find the network of maximum SVD
entropy while maintaining the joint degree sequence (type II MaxEnt
network model). (a) Nestedness (estimated using the spectral radius of
the adjacency matrix), (b) the maximum trophic level, (c) the network
diameter, and (d) the SVD entropy were measured on these empirical and
maximum entropy food webs. The identity line is plotted in each
panel.}\label{fig:measures}
}
\end{figure}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some desired
properties given by prior knowledge on a system. We first used MaxEnt to
predict the joint degree distribution and the degree distribution of
maximum entropy given known numbers of species and of interactions. We
found that the resulting joint degree distributions were more symmetric
than the ones of empirical food webs, which suggests that other
constraints might be needed to improve those predictions. We also used
MaxEnt to predict networks of maximum entropy with a specified
structure. These networks are the most complex, or random, given the
constraints used. Likewise, we found that knowledge of species richness
and of the number of links were not sufficient to reproduce many aspects
of network structure found in nature. However, a model based on the
entire joint degree sequence, i.e.~on the number of preys and predators
for each species, gave more convincing results. Our type II MaxEnt
network model yielded better or similar predictions than the type II
null model, also based on the joint degree sequence, for most measures
considered including the motifs profile. This suggests that the joint
degree sequence drives many aspects of network structure. Indeed,
considering our findings that empirical networks are close to maximum
entropy for a given joint degree sequence, our results suggest that
food-web topology is configured almost entirely randomly around these
marginal numbers of predators and preys.

Our method and results could be used for different purposes. First, they
could be used as first-order approximations of network structure when
only state variables are known. This could prove useful when predicting
network structure at large spatial scales, where few ecological
information is known at that scale. Second, they could be used as
informative priors in Bayesian analyses of the structure of ecological
networks. Third, they could be used to make better predictions of
pairwise species interactions by constraining the space of feasible
networks, as discussed in Strydom et al. (2021). Finally, they could be
used as alternative null models of ecological networks to better
understand ecological mechanisms driving food-web structure. In that
case, our model might need to be slightly adapted to give a probability
distribution of Boolean networks (in contrast with point estimates of
maximum entropy networks).

One of the biggest challenges in using the principle of maximum entropy
is to identify the set of state variables that best reproduce empirical
data. We found that the numbers of preys and predators for each species
are important state variables for the prediction of maximum entropy
networks. However, our predictions overestimated some measures of
network structure, especially the maximum trophic level and network
diameter. Therefore, we should continue playing the ecological detective
to find these other topological constraints that would improve the
predictions of our MaxEnt network models.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

We acknowledge that this study was conducted on land within the
traditional unceded territory of the Saint Lawrence Iroquoian,
Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak nations. This
work was supported by the Institute for Data Valorisation (IVADO) and
the NSERC BIOS\(^2\) CREATE program.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Albert2002StaMec}{}%
Albert, R., and A. L. Barabasi. 2002. {``Statistical Mechanics of
Complex Networks.''} \emph{Reviews of Modern Physics} 74 (1): 47--97.
\url{https://doi.org/10.1103/RevModPhys.74.47}.

\leavevmode\hypertarget{ref-Bascompte2003NesAssa}{}%
Bascompte, J., P. Jordano, C. J. Melian, and J. M. Olesen. 2003. {``The
Nested Assembly of Plant-Animal Mutualistic Networks.''}
\emph{Proceedings of the National Academy of Sciences of the United
States of America} 100 (16): 9383--87.
\url{https://doi.org/10.1073/pnas.1633576100}.

\leavevmode\hypertarget{ref-Bastolla2009ArcMut}{}%
Bastolla, Ugo, Miguel A. Fortuna, Alberto Pascual-Garcia, Antonio
Ferrera, Bartolo Luque, and Jordi Bascompte. 2009. {``The Architecture
of Mutualistic Networks Minimizes Competition and Increases
Biodiversity.''} \emph{Nature} 458 (7241): 1018--U91.
\url{https://doi.org/10.1038/nature07950}.

\leavevmode\hypertarget{ref-Beck2009GenInf}{}%
Beck, Christian. 2009. {``Generalised Information and Entropy Measures
in Physics.''} \emph{Contemporary Physics} 50 (4): 495--510.
\url{https://doi.org/10.1080/00107510902823517}.

\leavevmode\hypertarget{ref-Cimini2019StaPhy}{}%
Cimini, Giulio, Tiziano Squartini, Fabio Saracco, Diego Garlaschelli,
Andrea Gabrielli, and Guido Caldarelli. 2019. {``The Statistical Physics
of Real-World Networks.''} \emph{Nature Reviews Physics} 1 (1): 58--71.
\url{https://doi.org/10.1038/s42254-018-0002-6}.

\leavevmode\hypertarget{ref-Cohen2003EcoComa}{}%
Cohen, Joel E., Tomas Jonsson, and Stephen R. Carpenter. 2003.
{``Ecological Community Description Using the Food Web, Species
Abundance, and Body Size.''} \emph{Proceedings of the National Academy
of Sciences} 100 (4): 1781--86.
\url{https://doi.org/10.1073/pnas.232715699}.

\leavevmode\hypertarget{ref-Delmas2019AnaEco}{}%
Delmas, Eva, Mathilde Besson, Marie Hélène Brice, Laura A. Burkle,
Giulio V. Dalla Riva, Marie Josée Fortin, Dominique Gravel, et al. 2019.
{``Analysing Ecological Networks of Species Interactions.''}
\emph{Biological Reviews}. \url{https://doi.org/10.1111/brv.12433}.

\leavevmode\hypertarget{ref-Dunning2017JumMod}{}%
Dunning, Iain, Joey Huchette, and Miles Lubin. 2017. {``JuMP: A Modeling
Language for Mathematical Optimization.''} \emph{SIAM Review} 59 (2):
295--320. \url{https://doi.org/10.1137/15M1020575}.

\leavevmode\hypertarget{ref-Fortuna2006HabLos}{}%
Fortuna, M. A., and J. Bascompte. 2006. {``Habitat Loss and the
Structure of Plant-Animal Mutualistic Networks.''} \emph{Ecology
Letters} 9 (3): 278--83.
\url{https://doi.org/10.1111/j.1461-0248.2005.00868.x}.

\leavevmode\hypertarget{ref-Fortuna2010NesMod}{}%
Fortuna, Miguel A., Daniel B. Stouffer, Jens M. Olesen, Pedro Jordano,
David Mouillot, Boris R. Krasnov, Robert Poulin, and Jordi Bascompte.
2010. {``Nestedness Versus Modularity in Ecological Networks: Two Sides
of the Same Coin?''} \emph{Journal of Animal Ecology} 79 (4): 811--17.
\url{https://doi.org/10.1111/j.1365-2656.2010.01688.x}.

\leavevmode\hypertarget{ref-Frank2011SimDera}{}%
Frank, S. A., and E. Smith. 2011. {``A Simple Derivation and
Classification of Common Probability Distributions Based on Information
Symmetry and Measurement Scale.''} \emph{Journal of Evolutionary
Biology} 24 (3): 469--84.
\url{https://doi.org/10.1111/j.1420-9101.2010.02204.x}.

\leavevmode\hypertarget{ref-Harremoes2001MaxEnt}{}%
Harremoës, Peter, and Flemming Topsøe. 2001. {``Maximum Entropy
Fundamentals.''} \emph{Entropy} 3 (3): 191--226.
\url{https://doi.org/10.3390/e3030191}.

\leavevmode\hypertarget{ref-Harte2008MaxEnt}{}%
Harte, J., T. Zillio, E. Conlisk, and A. B. Smith. 2008. {``Maximum
Entropy and the State-Variable Approach to Macroecology.''}
\emph{Ecology} 89 (10): 2700--2711.
\url{https://doi.org/10.1890/07-1369.1}.

\leavevmode\hypertarget{ref-Harte2014MaxInf}{}%
Harte, John, and Erica A. Newman. 2014. {``Maximum Information Entropy:
A Foundation for Ecological Theory.''} \emph{Trends in Ecology \&
Evolution} 29 (7): 384--89.
\url{https://doi.org/10.1016/j.tree.2014.04.009}.

\leavevmode\hypertarget{ref-vanderHoorn2018SpaMaxa}{}%
Hoorn, Pim van der, Gabor Lippner, and Dmitri Krioukov. 2018. {``Sparse
Maximum-Entropy Random Graphs with a Given Power-Law Degree
Distribution.''} \emph{Journal of Statistical Physics} 173 (3): 806--44.
\url{https://doi.org/10.1007/s10955-017-1887-7}.

\leavevmode\hypertarget{ref-Hortal2015SevSho}{}%
Hortal, Joaquín, Francesco de Bello, José Alexandre F. Diniz-Filho,
Thomas M. Lewinsohn, Jorge M. Lobo, and Richard J. Ladle. 2015. {``Seven
Shortfalls That Beset Large-Scale Knowledge of Biodiversity.''}
\emph{Annual Review of Ecology, Evolution, and Systematics} 46 (1):
523--49. \url{https://doi.org/10.1146/annurev-ecolsys-112414-054400}.

\leavevmode\hypertarget{ref-Jaynes1957InfThe}{}%
Jaynes, E. T. 1957a. {``Information Theory and Statistical Mechanics.''}
\emph{Physical Review} 106 (4): 620--30.
\url{https://doi.org/10.1103/PhysRev.106.620}.

\leavevmode\hypertarget{ref-Jaynes1957InfThea}{}%
---------. 1957b. {``Information Theory and Statistical Mechanics.
II.''} \emph{Physical Review} 108 (2): 171--90.
\url{https://doi.org/10.1103/PhysRev.108.171}.

\leavevmode\hypertarget{ref-Khinchin2013MatFou}{}%
Khinchin, A. Ya. 2013. \emph{Mathematical Foundations of Information
Theory}. Courier Corporation.

\leavevmode\hypertarget{ref-MacDonald2020RevLin}{}%
MacDonald, Arthur Andrew Meahan, Francis Banville, and Timothée Poisot.
2020. {``Revisiting the Links-Species Scaling Relationship in Food
Webs.''} \emph{Patterns} 0 (0).
\url{https://doi.org/10.1016/j.patter.2020.100079}.

\leavevmode\hypertarget{ref-Martyushev2006MaxEnt}{}%
Martyushev, L. M., and V. D. Seleznev. 2006. {``Maximum Entropy
Production Principle in Physics, Chemistry and Biology.''} \emph{Physics
Reports-Review Section of Physics Letters} 426 (1): 1--45.
\url{https://doi.org/10.1016/j.physrep.2005.12.001}.

\leavevmode\hypertarget{ref-Milo2002NetMot}{}%
Milo, R., S. Shen-Orr, S. Itzkovitz, N. Kashtan, D. Chklovskii, and U.
Alon. 2002. {``Network Motifs: Simple Building Blocks of Complex
Networks.''} \emph{Science} 298 (5594): 824--27.
\url{https://doi.org/10.1126/science.298.5594.824}.

\leavevmode\hypertarget{ref-Park2004StaMeca}{}%
Park, Juyong, and M. E. J. Newman. 2004. {``Statistical Mechanics of
Networks.''} \emph{Physical Review E} 70 (6): 066117.
\url{https://doi.org/10.1103/PhysRevE.70.066117}.

\leavevmode\hypertarget{ref-Phillips2006MaxEnta}{}%
Phillips, Steven J., Robert P. Anderson, and Robert E. Schapire. 2006.
{``Maximum Entropy Modeling of Species Geographic Distributions.''}
\emph{Ecological Modelling} 190 (3): 231--59.
\url{https://doi.org/10.1016/j.ecolmodel.2005.03.026}.

\leavevmode\hypertarget{ref-Pielou1975EcoDiv}{}%
Pielou, Evelyn C. 1975. {``Ecological Diversity.''} In. 574.524018 P5.

\leavevmode\hypertarget{ref-Pomeranz2018DatInf}{}%
Pomeranz, Justin Page, Ross M. Thompson, Timothée Poisot, Jon S.
Harding, and Justin P. F. Pomeranz. 2018. {``Data from: Inferring
Predator-Prey Interactions in Food Webs.''} Dryad.
\url{https://doi.org/10.5061/DRYAD.K59M37F}.

\leavevmode\hypertarget{ref-Shannon1948MatThe}{}%
Shannon, C. E. 1948. {``A Mathematical Theory of Communication.''}
\emph{The Bell System Technical Journal} 27 (3): 379--423.
\url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}.

\leavevmode\hypertarget{ref-Staniczenko2013GhoNes}{}%
Staniczenko, Phillip P. A., Jason C. Kopp, and Stefano Allesina. 2013.
{``The Ghost of Nestedness in Ecological Networks.''} \emph{Nature
Communications} 4 (1): 1391. \url{https://doi.org/10.1038/ncomms2422}.

\leavevmode\hypertarget{ref-Stock2021OptTra}{}%
Stock, Michiel, Timothée Poisot, and Bernard De Baets. 2021. {``Optimal
Transportation Theory for Species Interaction Networks.''} \emph{Ecology
and Evolution} 11 (9): 3841--55.
\url{https://doi.org/10.1002/ece3.7254}.

\leavevmode\hypertarget{ref-Stouffer2011ComInc}{}%
Stouffer, Daniel B., and Jordi Bascompte. 2011. {``Compartmentalization
Increases Food-Web Persistence.''} \emph{Proceedings of the National
Academy of Sciences} 108 (9): 3648--52.
\url{https://doi.org/10.1073/pnas.1014353108}.

\leavevmode\hypertarget{ref-Stouffer2007EviExi}{}%
Stouffer, Daniel B., Juan Camacho, Wenxin Jiang, and Luis A. Nunes
Amaral. 2007. {``Evidence for the Existence of a Robust Pattern of Prey
Selection in Food Webs.''} \emph{Proceedings of the Royal Society
B-Biological Sciences} 274 (1621): 1931--40.
\url{https://doi.org/10.1098/rspb.2007.0571}.

\leavevmode\hypertarget{ref-Strydom2021RoaPre}{}%
Strydom, Tanya, Michael D. Catchen, Francis Banville, Dominique Caron,
Gabriel Dansereau, Philippe Desjardins-Proulx, Norma R. Forero-Muñoz, et
al. 2021. {``A Roadmap Towards Predicting Species Interaction Networks
(across Space and Time).''} \emph{Philosophical Transactions of the
Royal Society B: Biological Sciences} 376 (1837): 20210063.
\url{https://doi.org/10.1098/rstb.2021.0063}.

\leavevmode\hypertarget{ref-Strydom2021SvdEnt}{}%
Strydom, Tanya, Giulio V. Dalla Riva, and Timothee Poisot. 2021. {``SVD
Entropy Reveals the High Complexity of Ecological Networks.''}
\emph{Frontiers in Ecology and Evolution} 9: 623141.
\url{https://doi.org/10.3389/fevo.2021.623141}.

\leavevmode\hypertarget{ref-Williams2011BioMet}{}%
Williams, Richard J. 2011. {``Biology, Methodology or Chance? The Degree
Distributions of Bipartite Ecological Networks.''} \emph{PLOS ONE} 6
(3): e17645. \url{https://doi.org/10.1371/journal.pone.0017645}.

\leavevmode\hypertarget{ref-Williams2000SimRul}{}%
Williams, Richard J., and Neo D. Martinez. 2000. {``Simple Rules Yield
Complex Food Webs.''} \emph{Nature} 404 (6774): 180--83.
\url{https://doi.org/10.1038/35004572}.

\end{CSLReferences}

\end{document}

### 2022-03-24 preprint.tex
Deploy to GitHub Pages

--- 2022-03-24 draft.tex
@@
 %!TEX TS-program = xelatex
-\documentclass[11pt]{article}
+\documentclass[10pt,oneside]{article}
 
@@
 \newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=4.5cm}
-
-\usepackage{lineno}
-\usepackage[nolists,noheads]{endfloat}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}\usepackage[table,dvipsnames]{xcolor}
+
+\geometry{includemp,
+            letterpaper,
+            top=1.2in,
+            bottom=2.510cm,
+            inner=0.5in,
+            outer=0.4in,
+            marginparwidth=1.95in,
+            marginparsep=0.4in}
+
+\usepackage[singlelinecheck=off]{caption}
+\captionsetup{
+  font={small},
+  labelfont={bf},
+  format=plain,
+  labelsep=quad
+}
+\usepackage{floatrow}
+\floatsetup[figure]{margins=hangright,
+              facing=no,
+              capposition=beside,
+              capbesideposition={center,outside},
+              floatwidth=\textwidth}
+\floatsetup[table]{margins=hangoutside,
+             facing=yes,
+             capposition=beside,
+             capbesideposition={center,outside},
+             floatwidth=\textwidth}
 
@@
 
-\doublespacing
+\setcounter{secnumdepth}{5}
+
+\usepackage{titlesec}
+
+\titleformat{\section}[block]
+{\normalfont\large\sffamily}
+{\thesection}{.5em}{\titlerule\\[.8ex]\bfseries}
+
+\titleformat{\subsection}[runin]
+{\normalfont\fontseries{b}\selectfont\filright\sffamily}
+{\thesubsection.}{.5em}{}
+
+\titleformat{\subsubsection}[runin]
+{\normalfont\itshape\rmfamily\bfseries}{\thesubsubsection}{1em}{}
+
+\fancypagestyle{firstpage}
+{
+   \fancyhf{}
+   \renewcommand{\headrulewidth}{0pt}
+   \fancyfoot[R]{\footnotesize\ccby}
+   \fancyfoot[L]{\footnotesize\sffamily\today}
+}
 
@@
 }
+
+\usepackage{tikz}
 \begin{document}
-\thispagestyle{empty}
-{\Large\bfseries\sffamily Maximum entropy modelling of food-web
-structure}
-\vskip 5em
-
+\tikz [remember picture, overlay] %
+\node [shift={(-0.6in,1.1cm)},scale=0.2,opacity=0.4] at (current page.south east)[anchor=south east]{\includegraphics{logo}};%
+\pagestyle{normal}
+\thispagestyle{firstpage}
+
+\newcommand{\colorRule}[3][black]{\textcolor[HTML]{#1}{\rule{#2}{#3}}}
+
+\noindent {\LARGE \textbf{\textsf{Maximum entropy modelling of food-web
+structure}}}
+
+\medskip
+\begin{flushleft}
+{\small
 %
@@
 %
-\,\textsuperscript{1,2,3}\quad %
+\,\textsuperscript{1,2,3}, %
 \href{https://orcid.org/0000-0002-4498-7076}{Dominique\,Gravel}%
 %
-\,\textsuperscript{2,3}\quad %
+\,\textsuperscript{2,3}, %
 \href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
@@
 \,\textsuperscript{1,3}
-
+\vskip 1em
 \textsuperscript{1}\,Université de
-Montréal\quad \textsuperscript{2}\,Université de
-Sherbrooke\quad \textsuperscript{3}\,Quebec Centre for Biodiversity
-Science
-
+Montréal; \textsuperscript{2}\,Université de
+Sherbrooke; \textsuperscript{3}\,Quebec Centre for Biodiversity
+Science\\
+\vskip 1em
 \textbf{Correspondance to:}\\
 Francis Banville --- \texttt{francis.banville@umontreal.ca}\\
-
-\vfill
-This work is released by its authors under a CC-BY 4.0 license\hfill\ccby\\
-Last revision: \emph{\today}
-
-\clearpage
-\thispagestyle{empty}
-
-\vfill
-\textbf{\sffamily Abstract: }The principle of maximum entropy is a
-rigorous mathematical method of finding constrained probability
-distributions that has been proven useful in many ecological modelling
-problems. However, despite its broad application in graph and network
-theory, it has seldom been used to model ecological networks. Here we
-show how the (joint) degree distribution of maximum entropy can be
-directly derived using the number of species and the number of links in
-terrestrial and aquatic food webs. We also present a heuristic and
-flexible approach of finding the network of maximum entropy based on
-simulating annealing and SVD entropy. We built two of these
-network-level models using constraints given by the connectance and the
-joint degree sequence, respectively. All maximum entropy models were
-compared against open access food-web data and null and neutral models
-commonly used in network ecology. We found that the maximum entropy
-network model constrained by the joint degree sequence was a good
-predictor of food-web structure, including nestedness and motifs
-distribution. Overall, our results suggest that many properties of
-ecological networks are mainly driven by the joint degree distribution
-and statistical phenomena.
-\vfill
-
-\clearpage
-\linenumbers
-\pagestyle{normal}
+}
+\end{flushleft}
+
+\vskip 2em
+\makebox[0pt][l]{\colorRule[CCCCCC]{2.0\textwidth}{0.5pt}}
+\vskip 2em
+\noindent
+
+\marginpar{\vskip 1em\flushright
+{\small{\bfseries Keywords}:\par
+ecological modelling\\ecological networks\\food webs\\maximum
+entropy\\null models\\}
+}
+
+\textbf{Abstract}:\,The principle of maximum entropy is a rigorous
+mathematical method of finding constrained probability distributions
+that has been proven useful in many ecological modelling problems.
+However, despite its broad application in graph and network theory, it
+has seldom been used to model ecological networks. Here we show how the
+(joint) degree distribution of maximum entropy can be directly derived
+using the number of species and the number of links in terrestrial and
+aquatic food webs. We also present a heuristic and flexible approach of
+finding the network of maximum entropy based on simulating annealing and
+SVD entropy. We built two of these network-level models using
+constraints given by the connectance and the joint degree sequence,
+respectively. All maximum entropy models were compared against open
+access food-web data and null and neutral models commonly used in
+network ecology. We found that the maximum entropy network model
+constrained by the joint degree sequence was a good predictor of
+food-web structure, including nestedness and motifs distribution.
+Overall, our results suggest that many properties of ecological networks
+are mainly driven by the joint degree distribution and statistical
+phenomena.
+
+\vskip 2em
+\makebox[0pt][l]{\colorRule[CCCCCC]{2.0\textwidth}{0.5pt}}
+\vskip 2em
 

### 2022-03-24 suppmat.tex
Deploy to GitHub Pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage[]{times}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{times}
\renewcommand{\thefigure}{S\arabic{figure}}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{}

\begin{document}

\hypertarget{supplementary-materials}{%
\section{Supplementary materials}\label{supplementary-materials}}

\begin{figure}
\hypertarget{fig:kin_kout_diff}{%
\centering
\includegraphics{figures/kin_kout_difference.png}
\caption{Difference between predicted and empirical values for the
number of predators \(k_{in}\) and the number of preys \(k_{out}\) when
species are ordered according to (a) their out-degree and (b) their
in-degree. Empirical networks include all food webs archived on Mangal,
as well as the New-Zealand and Tuesday lake datasets (our complete
dataset). The joint degree sequences for simulated networks were
obtained after sampling one realization of the joint degree distribution
of maximum entropy for each network. Each dot corresponds to a single
species in one of the network of our dataset.}\label{fig:kin_kout_diff}
}
\end{figure}

\begin{figure}
\hypertarget{fig:heatmap}{%
\centering
\includegraphics{figures/heatmap_disconnected.png}
\caption{Probability that a species is isolated in its food web
according to the degree distribution of maximum entropy. We derived
degree distributions of maximum entropy given a range of values of \(S\)
and \(L\), and plotted the probability that a species has a degree \(k\)
of \(0\) (log-scale color bar). Here species richness varies between
\(5\) and \(100\) species, by increment of \(5\) species. For each level
of species richness, the numbers of links correspond to all 20-quantiles
of the interval between \(0\) and \(S^2\). The black line marks the
\(S-1\) minimum numbers of links required to have no isolated
species.}\label{fig:heatmap}
}
\end{figure}

\begin{figure}
\hypertarget{fig:entropy_dist}{%
\centering
\includegraphics{figures/entropy_distribution.png}
\caption{(a) Distribution of the SVD entropy of MaxEnt food webs (type
II MaxEnt network model) and of empirical food webs. Empirical networks
include all food webs archived on Mangal, as well as the New Zealand and
Tuesday lake datasets (our complete dataset). Maximum entropy networks
were derived using a simulating annealing algorithm to find the network
of maximum SVD entropy while maintaining the joint degree sequence. (b)
Distribution of z-scores of the SVD entropy of all empirical food webs.
Z-scores were computed using the mean and standard deviation of the
distribution of SVD entropy of MaxEnt food webs (type II MaxEnt network
model). The dash line corresponds to the median
z-score.}\label{fig:entropy_dist}
}
\end{figure}

\begin{figure}
\hypertarget{fig:entropy_size}{%
\centering
\includegraphics{figures/difference_entropy.png}
\caption{Difference in SVD entropy between MaxEnt (type II MaxEnt
network model) and empirical food webs as a function of (a) species
richness, (b) the number of links, and (c) connectance. Empirical
networks include all food webs archived on Mangal, as well as the New
Zealand and Tuesday lake datasets (our complete dataset). Maximum
entropy networks were derived using a simulating annealing algorithm to
find the network of maximum SVD entropy while maintaining the joint
degree sequence. Regression lines are plotted in each
panel.}\label{fig:entropy_size}
}
\end{figure}

\begin{figure}
\hypertarget{fig:measures_richness}{%
\centering
\includegraphics{figures/measures_richness.png}
\caption{Structure of empirical and maximum entropy food webs (type II
MaxEnt network model) as a function of species richness. Empirical
networks include all food webs archived on Mangal, as well as the New
Zealand and Tuesday lake datasets (our complete dataset). Maximum
entropy networks were derived using a simulating annealing algorithm to
find the network of maximum SVD entropy while maintaining the joint
degree sequence. (a) Nestedness (estimated with the spectral radius of
the adjacency matrix), (b) the maximum trophic level, (c) the network
diameter (i.e., the longest shortest path between all species pairs),
and (d) the SVD entropy were measured on these empirical and predicted
food webs and plotted against species richness. Regression lines are
plotted in each panel.}\label{fig:measures_richness}
}
\end{figure}

\end{document}

### 2022-07-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1b21aac669c69434820e120640ea1bea88b05e52 🚀

@@
 with an unbiased distribution of measures generated using a set of rules
-that exclude the mechanism of interest (M. A. Fortuna and Bascompte
-2006; Delmas et al. 2019). Both types of models are frequently
-topological, i.e.~they often predict the adjacency matrix or specific
-measures of network structure without taking into account species'
-identity. According to Strydom et al. (2021), these topological models
-could be used to make better predictions of pairwise species
-interactions by constraining the space of feasible networks.
+that exclude the mechanism of interest (Fortuna and Bascompte 2006;
+Delmas et al. 2019). Both types of models are frequently topological,
+i.e.~they often predict the adjacency matrix or specific measures of
+network structure without taking into account species' identity.
+According to Strydom et al. (2021), these topological models could be
+used to make better predictions of pairwise species interactions by
+constraining the space of feasible networks.
 
@@
 models (Phillips, Anderson, and Schapire 2006) and macroecological
-models (J. Harte et al. 2008; John Harte and Newman 2014). As discussed
-in Box 1, maximizing a measure of entropy ensures that the derived
-probability distributions are unique and least biased under the set of
-constraints used. These constraints are built using state variables,
-i.e.~variables that represent the macrostate of the system. The
-challenge is to find the set of state variables that best represent
-natural systems and to translate them into appropriate statistical
-constraints. Having a validated maximum entropy model for the system at
-hand allows us to make rigorous predictions using a minimal amount of
-data, as well as helping us describe the most important factors driving
-that system.
+models (Harte et al. 2008; Harte and Newman 2014). As discussed in Box
+1, maximizing a measure of entropy ensures that the derived probability
+distributions are unique and least biased under the set of constraints
+used. These constraints are built using state variables, i.e.~variables
+that represent the macrostate of the system. The challenge is to find
+the set of state variables that best represent natural systems and to
+translate them into appropriate statistical constraints. Having a
+validated maximum entropy model for the system at hand allows us to make
+rigorous predictions using a minimal amount of data, as well as helping
+us describe the most important factors driving that system.
 
@@
 the least-biased inference we could make on its species abundance
-distribution is the exponential distribution (Frank and Smith 2011; John
+distribution is the exponential distribution (Frank and Smith 2011;
 Harte and Newman 2014). However, this does not imply that this
@@
 with two topological null models. The first is the type I null model of
-M. A. Fortuna and Bascompte (2006), in which the probability that a
-species \(i\) predates on another species \(j\) is given by
+Fortuna and Bascompte (2006), in which the probability that a species
+\(i\) predates on another species \(j\) is given by
 
@@
 of the one of generalists, and is strongly associated with network
-modularity (Miguel A. Fortuna et al. 2010). We also measured the maximum
-trophic level \emph{maxtl}, network diameter \emph{diam} (i.e., the
-longest of the shortest paths between all species pairs; Albert and
-Barabasi 2002), the average maximum similarity between species pairs
-\emph{MxSim}, the proportion of cannibal species \emph{Cannib} (i.e.,
-the proportion of self loops), and the proportion of omnivorous species
-\emph{Omniv} (i.e., species whose preys are of different trophic
-levels). \emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply
-defined in Williams and Martinez (2000).
+modularity (Fortuna et al. 2010). We also measured the maximum trophic
+level \emph{maxtl}, network diameter \emph{diam} (i.e., the longest of
+the shortest paths between all species pairs; Albert and Barabasi 2002),
+the average maximum similarity between species pairs \emph{MxSim}, the
+proportion of cannibal species \emph{Cannib} (i.e., the proportion of
+self loops), and the proportion of omnivorous species \emph{Omniv}
+(i.e., species whose preys are of different trophic levels).
+\emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply defined in
+Williams and Martinez (2000).
 
@@
 \leavevmode\hypertarget{ref-Fortuna2010NesMod}{}%
-Fortuna, Miguel A., Daniel B. Stouffer, Jens M. Olesen, Pedro Jordano,
-David Mouillot, Boris R. Krasnov, Robert Poulin, and Jordi Bascompte.
-2010. {``Nestedness Versus Modularity in Ecological Networks: Two Sides
-of the Same Coin?''} \emph{Journal of Animal Ecology} 79 (4): 811--17.
+Fortuna, M. A., Daniel B. Stouffer, Jens M. Olesen, Pedro Jordano, David
+Mouillot, Boris R. Krasnov, Robert Poulin, and Jordi Bascompte. 2010.
+{``Nestedness Versus Modularity in Ecological Networks: Two Sides of the
+Same Coin?''} \emph{Journal of Animal Ecology} 79 (4): 811--17.
 \url{https://doi.org/10.1111/j.1365-2656.2010.01688.x}.
@@
 
+\leavevmode\hypertarget{ref-Harte2014MaxInf}{}%
+Harte, J., and Erica A. Newman. 2014. {``Maximum Information Entropy: A
+Foundation for Ecological Theory.''} \emph{Trends in Ecology \&
+Evolution} 29 (7): 384--89.
+\url{https://doi.org/10.1016/j.tree.2014.04.009}.
+
 \leavevmode\hypertarget{ref-Harte2008MaxEnt}{}%
@@
 \url{https://doi.org/10.1890/07-1369.1}.
-
-\leavevmode\hypertarget{ref-Harte2014MaxInf}{}%
-Harte, John, and Erica A. Newman. 2014. {``Maximum Information Entropy:
-A Foundation for Ecological Theory.''} \emph{Trends in Ecology \&
-Evolution} 29 (7): 384--89.
-\url{https://doi.org/10.1016/j.tree.2014.04.009}.
 

### 2022-07-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@793e17ad9ae1f8e38ec8aea9df678034c42dd448 🚀

@@
 of maximum entropy from which network properties can be measured. To do
-so, we used a flexible and heuristic approach based on simulating
+so, we used a flexible and heuristic approach based on simulated
 annealing to find networks \emph{close} to maximum entropy. As discussed
@@
 our \emph{complete dataset}. First, all food webs archived on
-\texttt{mangal.io} were directly queried from the database
-(\(n = 235\)). Most ecological networks archived on Mangal are
-multilayer networks, i.e.~networks that describe different types of
-interactions. We kept all networks whose interactions were mainly of
-predation and herbivory types, and removed the largest network
-(\(S = 714\)) for computational efficiency reasons. Then, to this set we
-added food webs from two different sources: the New-Zealand dataset
-(\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
-(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
-\(19\) networks had data on species' relative abundances that were used
-in the neutral model presented in a following subsection. These networks
-are part of what we call our \emph{abundance dataset}, which is a subset
-of our complete dataset.
+\texttt{mangal.io} (Poisot et al. 2016; Banville, Vissault, and Poisot
+2021) were directly queried from the database (\(n = 235\)). Most
+ecological networks archived on Mangal are multilayer networks,
+i.e.~networks that describe different types of interactions. We kept all
+networks whose interactions were mainly of predation and herbivory
+types, and removed the largest network (\(S = 714\)) for computational
+efficiency reasons. Then, to this set we added food webs from two
+different sources: the New-Zealand dataset (\(n = 21\); Pomeranz et al.
+2018) and the Tuesday lake dataset (\(n = 2\); Cohen, Jonsson, and
+Carpenter 2003). Of these two datasets, \(19\) networks had data on
+species' relative abundances that were used in the neutral model
+presented in a following subsection. These networks are part of what we
+call our \emph{abundance dataset}, which is a subset of our complete
+dataset.
 
@@
 constraints as the types I and II null models presented above. For both
-models, we used a simulating annealing algorithm with \(4\) chains,
+models, we used a simulated annealing algorithm with \(4\) chains,
 \(2000\) steps and an initial temperature of \(0.2\). For each chain, we
@@
 well as the New-Zealand and Tuesday lake datasets (our complete
-dataset). Maximum entropy networks were estimated using a simulating
+dataset). Maximum entropy networks were estimated using a simulated
 annealing algorithm to find the network of maximum SVD entropy while
@@
 (our complete dataset). Maximum entropy networks were estimated using a
-simulating annealing algorithm to find the network of maximum SVD
-entropy while maintaining the connectance (type I MaxEnt network model)
-or the joint degree sequence (type II MaxEnt network model). The
-predictions of the types I and II null models are also plotted.
-Regression lines are plotted in each panel. Motifs names are from
-Stouffer et al. (2007).}\label{fig:motifs_rel}
+simulated annealing algorithm to find the network of maximum SVD entropy
+while maintaining the connectance (type I MaxEnt network model) or the
+joint degree sequence (type II MaxEnt network model). The predictions of
+the types I and II null models are also plotted. Regression lines are
+plotted in each panel. Motifs names are from Stouffer et al.
+(2007).}\label{fig:motifs_rel}
 }
@@
 complete dataset). Maximum entropy networks were estimated using a
-simulating annealing algorithm to find the network of maximum SVD
-entropy while maintaining the joint degree sequence (type II MaxEnt
-network model). (a) Nestedness (estimated using the spectral radius of
-the adjacency matrix), (b) the maximum trophic level, (c) the network
+simulated annealing algorithm to find the network of maximum SVD entropy
+while maintaining the joint degree sequence (type II MaxEnt network
+model). (a) Nestedness (estimated using the spectral radius of the
+adjacency matrix), (b) the maximum trophic level, (c) the network
 diameter, and (d) the SVD entropy were measured on these empirical and
@@
 
+\leavevmode\hypertarget{ref-Banville2021ManJla}{}%
+Banville, Francis, Steve Vissault, and Timothée Poisot. 2021.
+{``Mangal.jl and EcologicalNetworks.jl: Two Complementary Packages for
+Analyzing Ecological Networks in Julia.''} \emph{Journal of Open Source
+Software} 6 (61): 2721. \url{https://doi.org/10.21105/joss.02721}.
+
 \leavevmode\hypertarget{ref-Bascompte2003NesAssa}{}%
@@
 Pielou, Evelyn C. 1975. {``Ecological Diversity.''} In. 574.524018 P5.
+
+\leavevmode\hypertarget{ref-Poisot2016ManMakb}{}%
+Poisot, Timothee, Benjamin Baiser, Jennifer A. Dunne, Sonia Kefi,
+Francois Massol, Nicolas Mouquet, Tamara N. Romanuk, Daniel B. Stouffer,
+Spencer A. Wood, and Dominique Gravel. 2016. {``Mangal - Making
+Ecological Network Analysis Simple.''} \emph{Ecography} 39 (4): 384--90.
+\url{https://doi.org/10.1111/ecog.00976}.
 

### 2022-07-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@3d6d340641bce3d032a9391b700ea89c923f4a03 🚀

@@
 Statistical and mathematical models can help fill many gaps in our
-knowledge about species interactions. Two complementary types of models
-have been developed in network ecology for this purpose. On one hand,
-predictive models can partially alleviate the Eltonian shortfall, which
-describes our current lack of knowledge on food webs and other
-ecological networks (Hortal et al. 2015). A variety of such models have
-recently been developed using machine learning and other statistical
-tools, most of which are presented in Strydom et al. (2021). On the
-other hand, null models help us identify potential ecological mechanisms
-that drive species interactions. They do so by comparing empirical data
-with an unbiased distribution of measures generated using a set of rules
-that exclude the mechanism of interest (Fortuna and Bascompte 2006;
-Delmas et al. 2019). Both types of models are frequently topological,
-i.e.~they often predict the adjacency matrix or specific measures of
-network structure without taking into account species' identity.
-According to Strydom et al. (2021), these topological models could be
-used to make better predictions of pairwise species interactions by
-constraining the space of feasible networks.
+knowledge about species interactions. Predictive and null models are two
+complementary types of models that have been developed in network
+ecology for this purpose. On one hand, predictive models can partially
+alleviate the Eltonian shortfall, which describes our current lack of
+knowledge on food webs and other ecological networks (Hortal et al.
+2015). A variety of such models have recently been developed using
+machine learning and other statistical tools, most of which are
+presented in Strydom et al. (2021). On the other hand, null models help
+us identify potential ecological mechanisms that drive species
+interactions. They do so by comparing empirical data with an unbiased
+distribution of measures generated using a set of rules that exclude the
+mechanism of interest (Fortuna and Bascompte 2006; Delmas et al. 2019).
+Both types of models are frequently topological, i.e.~they often predict
+the adjacency matrix or specific measures of network structure without
+taking into account species' identity. According to Strydom et al.
+(2021), these topological models could be used to make better
+predictions of pairwise species interactions by constraining the space
+of feasible networks.
 
@@
 Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
-comparison been little used in network ecology. The very nature of
-ecological networks (directed simple graphs frequently having
-self-loops) makes the mathematical optimization of maximum entropy graph
-models more complicated than with many other types of (non-ecological)
-networks. MaxEnt has nevertheless been used to predict the degree
-distribution of bipartite ecological networks from the number of species
-and the number of interactions (Williams 2011) and to predict
-interaction strengths between species pairs using their relative
-abundances within an optimal transportation theory regularized with
-information entropy (Stock, Poisot, and De Baets 2021). However, to the
-best of our knowledge, MaxEnt has never been used to predict food-web
-structure directly, even though food webs are among the most documented
-and widespread ecological networks.
+comparison been little used in network ecology. Like other real system
+networks, ecological networks are represented mathematically as graphs.
+However, the very nature of ecological networks (directed simple graphs
+frequently having self-loops) makes the mathematical optimization of
+maximum entropy graph models more complicated than with many other types
+of (non-ecological) networks. MaxEnt has nevertheless been used to
+predict the degree distribution of bipartite ecological networks from
+the number of species and the number of interactions (Williams 2011) and
+to predict interaction strengths between species pairs using their
+relative abundances within an optimal transportation theory regularized
+with information entropy (Stock, Poisot, and De Baets 2021). However, to
+the best of our knowledge, MaxEnt has never been used to predict
+food-web structure directly, even though food webs are among the most
+documented and widespread ecological networks.
 
@@
 its in and out-degrees). Because of the scarcity of empirical data on
-the number of links in ecological networks, in Box 2 we present a method
-to predict \(L\) from \(S\), thus allowing the prediction of the joint
-degree distribution from \(S\) solely. In turn, the second approach
-consists in finding, using different constraints, the adjacency matrix
-of maximum entropy from which network properties can be measured. To do
-so, we used a flexible and heuristic approach based on simulated
+the number of interactions in ecological networks, we present a method
+to predict \(L\) from \(S\) (Box 2), thus allowing the prediction of the
+joint degree distribution from \(S\) solely. In turn, the second
+approach consists in finding, using different constraints, the adjacency
+matrix of maximum entropy from which network properties can be measured.
+To do so, we used a flexible and heuristic approach based on simulated
 annealing to find networks \emph{close} to maximum entropy. As discussed
 above, our choice of algorithm stands from the very nature of food webs
-(i.e., simple directed networks allowing self-loops) that makes the
+(i.e.~simple directed networks allowing self-loops) that makes the
 analytical derivation of a maximum entropy graph model difficult. We
 first built our type I MaxEnt network model constrained by the
-connectance of the network (i.e., the ratio \(L/S^2\)). A comparison of
+connectance of the network (i.e.~the ratio \(L/S^2\)). A comparison of
 this model against empirical data indicated that connectance alone was
@@
 Starting from a set of constraints given by prior knowledge of a system
-(i.e., what we call state variables), this method helps us find
+(i.e.~what we call state variables), this method helps us find
 least-biased probability distributions subject to the constraints. These
@@
 degree distribution analytically from the number of species and the
-number of links in food webs.
+number of interactions in food webs.
 
@@
 the singular values). Following Strydom, Dalla Riva, and Poisot (2021),
-we standardized this measure with the rank \(R\) of the matrix (i.e.,
-\(J / \ln(R)\)) to account for the difference in dimensions between
-networks (Pielou's evenness; Pielou 1975). In a following section, we
-will show how SVD entropy can be used to predict a network of maximum
-entropy (i.e., of maximum complexity) heuristically.
+we standardized this measure with the rank \(R\) of the matrix
+(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
+between networks (\emph{sensu} Pielou's evenness; Pielou 1975). In a
+following section, we will show how SVD entropy can be used to predict a
+network of maximum entropy (i.e.~of maximum complexity) heuristically.
 
@@
 that were drawn the most amount of time, with \(L\) given by the number
-of links in each food web. This ensured that the resulting null networks
-had the same number of interactions as their empirical counterparts.
+of interactions in each food web. This ensured that the resulting null
+networks had the same number of interactions as their empirical
+counterparts.
 
@@
 food web in our complete dataset, i.e.~using their numbers of species
-and numbers of links as state variables. We then sampled one realization
-of the degree sequence for each network using the probabilities given by
-the joint degree distribution. In fig.~\ref{fig:joint_dd} (left panels),
-we show the relationship between \(k_{out}\) and \(k_{in}\) standardized
-by the number of species in their networks, for empirical and maximum
-entropy joint degree distributions. We see that our model predicts a
-greater number of generalist species than empirical food webs (i.e.,
-relative values of \(k_{out}\) and \(k_{in}\) closer to \(1\)). However,
-plotting the difference between predicted and empirical values for each
-species gives a different perspective. The right panel of
+and numbers of interactions as state variables. We then sampled one
+realization of the degree sequence for each network using the
+probabilities given by the joint degree distribution. In
+fig.~\ref{fig:joint_dd} (left panels), we show the relationship between
+\(k_{out}\) and \(k_{in}\) standardized by the number of species in
+their networks, for empirical and maximum entropy joint degree
+distributions. We see that our model predicts higher values of
+generality and vulnerability compared to empirical food webs
+(i.e.~relative values of \(k_{out}\) and \(k_{in}\) closer to \(1\)).
+However, plotting the difference between predicted and empirical values
+for each species gives a different perspective. The right panel of
 fig.~\ref{fig:joint_dd} presents these differences when species are
-ordered by their total degree in their network (i.e., by the sum of
-their in and out-degrees). Indeed, our predicted joint degree sequences
-have the same number of species as their empirical counterparts, but
-they are species agnostic; in other words, instead of predicting a value
-for each species directly, we predicted the entire joint degree sequence
-without taking into account species' identity. When we associate
-predictions and empirical data according to their rank in total degrees,
-we see that species predicted to be have a higher generality (number of
-preys) generally have a lower vulnerability (number of preys) than what
-is observed (and conversely). In fig.~S1, we show how these differences
+ordered by their total degree in their network (i.e.~by the sum of their
+in and out-degrees). Indeed, our predicted joint degree sequences have
+the same number of species as their empirical counterparts, but they are
+species agnostic; in other words, instead of predicting a value for each
+species directly, we predicted the entire joint degree sequence without
+taking into account species' identity. When we associate predictions and
+empirical data according to their rank in total degrees, we see that
+species predicted to be have a higher generality (number of preys)
+generally have a lower vulnerability (number of preys) than what is
+observed (and conversely). In fig.~S1, we show how these differences
 change when species are instead ordered by their out-degrees (left
@@
 The degree distribution \(p(k)\) represents the probability that a
-species has \(k\) links in a food web, with \(k = k_{in} + k_{out}\). It
-can thus be directly obtained from the joint degree distribution:
+species has \(k\) interactions in a food web, with
+\(k = k_{in} + k_{out}\). It can thus be directly obtained from the
+joint degree distribution:
 
@@
 given \(S\) and \(L\), predicts very low probabilities that a species
-will be isolated in its food web (\emph{i.e.}, having \(k = 0\)). As
-MacDonald, Banville, and Poisot (2020) pointed out, the size of food
-webs should at least be of \(S-1\) links, since a lower number would
-yield isolated species, i.e.~species without any predators or preys. Our
+will be isolated in its food web (i.e.~having \(k = 0\)). As MacDonald,
+Banville, and Poisot (2020) pointed out, the size of food webs should at
+least be of \(S-1\) interactions, since a lower number would yield
+isolated species, i.e.~species without any predators or preys. Our
 results show that, under our purely information-theoretic model, the
@@
 threshold. The expected proportion of isolated species rapidly declines
-by orders of magnitude with increasing numbers of species and links.
+by orders of magnitude with increasing numbers of species and
+interactions.
 
@@
 
-\hypertarget{box-2---working-with-predicted-numbers-of-links}{%
+\hypertarget{box-2---working-with-predicted-numbers-of-interactions}{%
 \section{Box 2 - Working with predicted numbers of
-links}\label{box-2---working-with-predicted-numbers-of-links}}
+interactions}\label{box-2---working-with-predicted-numbers-of-interactions}}
 
 Our models need information on the number of species and the number of
-links. However, since the later is rarely estimated empirically,
+interactions. However, since the later is rarely estimated empirically,
 ecologists might need to use predictive methods to estimate the total
-number of links in a food web.
+number of interactions in a food web.
 
@@
 The flexible links model, in contrast to other predictive models of the
-number of links, incorporates meaningful ecological constraints into the
-prediction of \(L\), namely the minimum \(S-1\) and maximum \(S^2\)
-numbers of interactions in food webs. It estimates the proportion of the
-\(S^2 - (S - 1)\) \emph{flexible links} that are realized. More
-precisely, this model states that the number of \emph{realized} flexible
-links \(L_{FL}\) in a food web represents the number of realized
-interactions above the minimum (i.e., \(L = L_{FL} + S - 1\)) and is
-obtained from a beta-binomial distribution with \(S^2 - (S - 1)\) trials
-and parameters \(\alpha = \mu e^\phi\) and \(\beta = (1 - \mu) e^\phi\):
+number of interactions, incorporates meaningful ecological constraints
+into the prediction of \(L\), namely the minimum \(S-1\) and maximum
+\(S^2\) numbers of interactions in food webs. It estimates the
+proportion of the \(S^2 - (S - 1)\) \emph{flexible links} that are
+realized. More precisely, this model states that the number of
+\emph{realized} flexible links (or interactions) \(L_{FL}\) in a food
+web represents the number of realized interactions above the minimum
+(i.e.~\(L = L_{FL} + S - 1\)) and is obtained from a beta-binomial
+distribution with \(S^2 - (S - 1)\) trials and parameters
+\(\alpha = \mu e^\phi\) and \(\beta = (1 - \mu) e^\phi\):
 
@@
 \caption{(a) Probability density of the predicted mean degree for
-different values of species richness. The number of links was predicted
-using the flexible links model fitted to all empirical networks in our
-complete dataset. (b) Degree distributions of maximum entropy for a
-network of \(27\) species and different numbers of links. The numbers of
-links correspond to the lower and upper bounds of the \(67%
+different values of species richness. The number of interactions was
+predicted using the flexible links model fitted to all empirical
+networks in our complete dataset. (b) Degree distributions of maximum
+entropy for a network of \(27\) species and different numbers of
+interactions. The numbers of interactions correspond to the lower and
+upper bounds of the \(67%
 \), \(89%
@@
 that predict a probability distribution on networks under soft or hard
-constraints Cimini et al. (2019). We believe our approach to be more
-flexible, easier to compute, while allowing direct comparisons of
-empirical food webs with more complex networks with similar structure.
+constraints (e.g., Park and Newman 2004; Cimini et al. 2019). We believe
+our approach to be more flexible, easier to compute, while allowing
+direct comparisons of empirical food webs with more complex networks
+with similar structure.
 
@@
 modularity (Fortuna et al. 2010). We also measured the maximum trophic
-level \emph{maxtl}, network diameter \emph{diam} (i.e., the longest of
+level \emph{maxtl}, network diameter \emph{diam} (i.e.~the longest of
 the shortest paths between all species pairs; Albert and Barabasi 2002),
 the average maximum similarity between species pairs \emph{MxSim}, the
-proportion of cannibal species \emph{Cannib} (i.e., the proportion of
+proportion of cannibal species \emph{Cannib} (i.e.~the proportion of
 self loops), and the proportion of omnivorous species \emph{Omniv}
-(i.e., species whose preys are of different trophic levels).
+(i.e.~species whose preys are of different trophic levels).
 \emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply defined in
@@
 between empirical and type II MaxEnt networks and species richness, the
-number of links, or connectance (fig.~S4). Similarly, we found no
+number of interactions, or connectance (fig.~S4). Similarly, we found no
 correlation between the difference in SVD entropy and the Jaccard
@@
 constraints used. Likewise, we found that knowledge of species richness
-and of the number of links were not sufficient to reproduce many aspects
-of network structure found in nature. However, a model based on the
-entire joint degree sequence, i.e.~on the number of preys and predators
-for each species, gave more convincing results. Our type II MaxEnt
-network model yielded better or similar predictions than the type II
-null model, also based on the joint degree sequence, for most measures
-considered including the motifs profile. This suggests that the joint
-degree sequence drives many aspects of network structure. Indeed,
+and of the number of interactions were not sufficient to reproduce many
+aspects of network structure found in nature. However, a model based on
+the entire joint degree sequence, i.e.~on the number of preys and
+predators for each species, gave more convincing results. Our type II
+MaxEnt network model yielded better or similar predictions than the type
+II null model, also based on the joint degree sequence, for most
+measures considered including the motifs profile. This suggests that the
+joint degree sequence drives many aspects of network structure. Indeed,
 considering our findings that empirical networks are close to maximum
@@
 informative priors in Bayesian analyses of the structure of ecological
-networks. Third, they could be used to make better predictions of
-pairwise species interactions by constraining the space of feasible
-networks, as discussed in Strydom et al. (2021). Finally, they could be
-used as alternative null models of ecological networks to better
-understand ecological mechanisms driving food-web structure. In that
-case, our model might need to be slightly adapted to give a probability
-distribution of Boolean networks (in contrast with point estimates of
-maximum entropy networks).
+networks (e.g., Cirtwill et al. 2019). Third, they could be used to make
+better predictions of pairwise species interactions by constraining the
+space of feasible networks, as discussed in Strydom et al. (2021).
+Finally, they could be used as alternative null models of ecological
+networks to better understand ecological mechanisms driving food-web
+structure. In that case, our model might need to be slightly adapted to
+give a probability distribution of Boolean networks (in contrast with
+point estimates of maximum entropy networks).
 
@@
 
+\leavevmode\hypertarget{ref-Cirtwill2019QuaFrac}{}%
+Cirtwill, Alyssa R., Anna Eklof, Tomas Roslin, Kate Wootton, and
+Dominique Gravel. 2019. {``A Quantitative Framework for Investigating
+the Reliability of Empirical Network Construction.''} \emph{Methods in
+Ecology and Evolution} 10 (6): 902--11.
+\url{https://doi.org/10.1111/2041-210X.13180}.
+
 \leavevmode\hypertarget{ref-Cohen2003EcoComa}{}%

### 2022-07-19 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@587d1042f4ada9e8e096c69a8f3032b236224fdd 🚀

@@
 \thispagestyle{empty}
-{\Large\bfseries\sffamily Maximum entropy modelling of food-web
-structure}
+{\Large\bfseries\sffamily What constrains food webs? A maximum entropy
+framework for predicting their structure with minimal biases}
 \vskip 5em
@@
 \vfill
-\textbf{\sffamily Abstract: }The principle of maximum entropy is a
-rigorous mathematical method of finding constrained probability
-distributions that has been proven useful in many ecological modelling
-problems. However, despite its broad application in graph and network
-theory, it has seldom been used to model ecological networks. Here we
-show how the (joint) degree distribution of maximum entropy can be
-directly derived using the number of species and the number of links in
-terrestrial and aquatic food webs. We also present a heuristic and
-flexible approach of finding the network of maximum entropy based on
-simulating annealing and SVD entropy. We built two of these
-network-level models using constraints given by the connectance and the
-joint degree sequence, respectively. All maximum entropy models were
-compared against open access food-web data and null and neutral models
-commonly used in network ecology. We found that the maximum entropy
-network model constrained by the joint degree sequence was a good
-predictor of food-web structure, including nestedness and motifs
-distribution. Overall, our results suggest that many properties of
-ecological networks are mainly driven by the joint degree distribution
-and statistical phenomena.
+\textbf{\sffamily Abstract: }Food webs are complex ecological networks
+whose structure is both ecologically and statistically constrained, with
+many network properties being correlated with each other. Despite the
+recognition of these invariable relationships in food webs, the use of
+the principle of maximum entropy (MaxEnt) in network ecology is still
+rare. This is surprising considering that MaxEnt is a renowned and
+rigorous statistical tool precisely designed for understanding and
+predicting many different types of constrained systems. Precisely, this
+principle asserts that the least-biased probability distribution of a
+system's property, constrained by prior knowledge about that system, is
+the one with maximum information entropy. MaxEnt has been proven useful
+in many ecological modelling problems, but its application in food webs
+and other ecological networks is limited. Here we show how MaxEnt can be
+used to derive many food-web properties both analytically and
+heuristically. First, we show how the joint degree distribution (the
+joint probability distribution of the numbers of prey and predators for
+each species in the network) can be derived analytically using the
+number of species and the number of interactions in food webs. Second,
+we present a heuristic and flexible approach of finding a network's
+adjacency matrix (the network's representation in matrix format) based
+on simulated annealing and SVD entropy. We built two heuristic models
+using the connectance and the joint degree sequence as statistical
+constraints, respectively. We compared both models' predictions against
+corresponding null and neutral models commonly used in network ecology
+using open access data of terrestrial and aquatic food webs sampled
+globally (N = 257). We found that the heuristic model constrained by the
+joint degree sequence was a good predictor of many measures of food-web
+structure, especially the nestedness and motifs distribution.
+Specifically, our results suggest that the structure of terrestrial and
+aquatic food webs is mainly driven by their joint degree distribution.
 \vfill
@@
 work was supported by the Institute for Data Valorisation (IVADO) and
-the NSERC BIOS\(^2\) CREATE program.
+the Natural Sciences and Engineering Research Council of Canada (NSERC)
+Collaborative Research and Training Experience (CREATE) program, through
+the Computational Biodiversity Science and Services (BIOS\(^2\))
+program.
 

### 2022-07-19 preprint.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@587d1042f4ada9e8e096c69a8f3032b236224fdd 🚀

@@
 
-\noindent {\LARGE \textbf{\textsf{Maximum entropy modelling of food-web
-structure}}}
+\noindent {\LARGE \textbf{\textsf{What constrains food webs? A maximum
+entropy framework for predicting their structure with minimal biases}}}
 
@@
 
-\textbf{Abstract}:\,The principle of maximum entropy is a rigorous
-mathematical method of finding constrained probability distributions
-that has been proven useful in many ecological modelling problems.
-However, despite its broad application in graph and network theory, it
-has seldom been used to model ecological networks. Here we show how the
-(joint) degree distribution of maximum entropy can be directly derived
-using the number of species and the number of links in terrestrial and
-aquatic food webs. We also present a heuristic and flexible approach of
-finding the network of maximum entropy based on simulating annealing and
-SVD entropy. We built two of these network-level models using
-constraints given by the connectance and the joint degree sequence,
-respectively. All maximum entropy models were compared against open
-access food-web data and null and neutral models commonly used in
-network ecology. We found that the maximum entropy network model
-constrained by the joint degree sequence was a good predictor of
-food-web structure, including nestedness and motifs distribution.
-Overall, our results suggest that many properties of ecological networks
-are mainly driven by the joint degree distribution and statistical
-phenomena.
+\textbf{Abstract}:\,Food webs are complex ecological networks whose
+structure is both ecologically and statistically constrained, with many
+network properties being correlated with each other. Despite the
+recognition of these invariable relationships in food webs, the use of
+the principle of maximum entropy (MaxEnt) in network ecology is still
+rare. This is surprising considering that MaxEnt is a renowned and
+rigorous statistical tool precisely designed for understanding and
+predicting many different types of constrained systems. Precisely, this
+principle asserts that the least-biased probability distribution of a
+system's property, constrained by prior knowledge about that system, is
+the one with maximum information entropy. MaxEnt has been proven useful
+in many ecological modelling problems, but its application in food webs
+and other ecological networks is limited. Here we show how MaxEnt can be
+used to derive many food-web properties both analytically and
+heuristically. First, we show how the joint degree distribution (the
+joint probability distribution of the numbers of prey and predators for
+each species in the network) can be derived analytically using the
+number of species and the number of interactions in food webs. Second,
+we present a heuristic and flexible approach of finding a network's
+adjacency matrix (the network's representation in matrix format) based
+on simulated annealing and SVD entropy. We built two heuristic models
+using the connectance and the joint degree sequence as statistical
+constraints, respectively. We compared both models' predictions against
+corresponding null and neutral models commonly used in network ecology
+using open access data of terrestrial and aquatic food webs sampled
+globally (N = 257). We found that the heuristic model constrained by the
+joint degree sequence was a good predictor of many measures of food-web
+structure, especially the nestedness and motifs distribution.
+Specifically, our results suggest that the structure of terrestrial and
+aquatic food webs is mainly driven by their joint degree distribution.
 

### 2022-07-19 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@23efa8475c06ad8174f1be4d1e9041c517d5daf8 🚀

@@
 make predictions of network structure and to better understand processes
-shaping ecological networks. This mathematical method, briefly presented
-in Box 1, has been used in a wide range of disciplines, from
-thermodynamics to chemistry and biology (Martyushev and Seleznev 2006).
-It has also been proven useful in ecology, e.g.~in species distribution
-models (Phillips, Anderson, and Schapire 2006) and macroecological
-models (Harte et al. 2008; Harte and Newman 2014). As discussed in Box
-1, maximizing a measure of entropy ensures that the derived probability
-distributions are unique and least biased under the set of constraints
-used. These constraints are built using state variables, i.e.~variables
-that represent the macrostate of the system. The challenge is to find
-the set of state variables that best represent natural systems and to
-translate them into appropriate statistical constraints. Having a
-validated maximum entropy model for the system at hand allows us to make
-rigorous predictions using a minimal amount of data, as well as helping
-us describe the most important factors driving that system.
+shaping ecological networks. Food-web properties that can be derived
+using MaxEnt are varied and pertain to different elements of the network
+(i.e.~at the species (node), the interaction (edge) or the community
+(network) levels). For example, at the species level, MaxEnt can be used
+to predict the distribution of trophic levels among species, as well as
+the distribution of species' vulnerability (number of predators) and
+generality (number of preys). By contrast, at the interaction level,
+predictions can be made on the distribution of interaction strengths in
+weighted food webs. At the community level, it can give us a probability
+distribution of potential networks or of specific measures of their
+emerging structure. The applicability and performance of this method
+depend on the ecological information available and on our capacity to
+find the set of state variables that best represent natural systems and
+to translate them into appropriate statistical constraints. MaxEnt has
+been used in a wide range of disciplines, from thermodynamics to
+chemistry and biology (Martyushev and Seleznev 2006). It has also been
+proven useful in ecology, e.g.~in species distribution models (Phillips,
+Anderson, and Schapire 2006) and macroecological models (Harte et al.
+2008; Harte and Newman 2014). Maximizing a measure of entropy ensures
+that the derived probability distributions are unique and least biased
+under the set of constraints used. These constraints are built using
+state variables, i.e.~variables that represent the macrostate of the
+system. Having a validated maximum entropy model for the system at hand
+allows us to make rigorous predictions using a minimal amount of data,
+as well as helping us describe the most important factors driving that
+system.
 
@@
 the number of interactions in ecological networks, we present a method
-to predict \(L\) from \(S\) (Box 2), thus allowing the prediction of the
+to predict \(L\) from \(S\) (Box 1), thus allowing the prediction of the
 joint degree distribution from \(S\) solely. In turn, the second
@@
 
-\hypertarget{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
-\section{Box 1 - The principle of maximum entropy: A primer for
-ecologists}\label{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}}
+\hypertarget{the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
+\subsection{The principle of maximum entropy: A primer for
+ecologists}\label{the-principle-of-maximum-entropy-a-primer-for-ecologists}}
 
@@
 
-In this contribution, we also use the SVD entropy as a measure of
-entropy, which is an application of Shannon's entropy to the relative
-non-zero singular values of a truncated singular value decomposition
-(t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a food web's Boolean
-adjacency matrix. This measure also satisfies all four properties
-above-mentioned, while being a proper measure of the internal complexity
-of food webs (Strydom, Dalla Riva, and Poisot 2021). We measured SVD
+Shannon's entropy can only be calculated on conventional probability
+distributions such as the joint degree distribution. This is an issue
+when working with ecological networks, which are frequently represented
+as adjacency matrices. For this reason, in this contribution, we also
+use the SVD entropy as a measure of entropy, which is an application of
+Shannon's entropy to the relative non-zero singular values of a
+truncated singular value decomposition (t-SVD; Strydom, Dalla Riva, and
+Poisot 2021) of a food web's Boolean adjacency matrix. We measured SVD
 entropy as follows:
@@
 
-where \(s_i\) are the relative singular values
+where \(s_i\) are the relative singular values of the adjacency matrix
 (\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
-the singular values). Following Strydom, Dalla Riva, and Poisot (2021),
-we standardized this measure with the rank \(R\) of the matrix
-(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
-between networks (\emph{sensu} Pielou's evenness; Pielou 1975). In a
-following section, we will show how SVD entropy can be used to predict a
-network of maximum entropy (i.e.~of maximum complexity) heuristically.
-
-\hypertarget{testing-maxent-models}{%
-\section{Testing MaxEnt models}\label{testing-maxent-models}}
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
+the singular values). Note that the distribution of relative singular
+values is analogous to a probability distribution, with \(0 < s_i < 1\)
+and \(\sum s_i = 1\). Thus, this measure also satisfies all four
+properties above-mentioned, while being a proper measure of the internal
+complexity of food webs (Strydom, Dalla Riva, and Poisot 2021).
+Following Strydom, Dalla Riva, and Poisot (2021), we standardized this
+measure with the rank \(R\) of the matrix (i.e.~\(J / \ln(R)\)) to
+account for the difference in dimensions between networks (\emph{sensu}
+Pielou's evenness; Pielou 1975). We will show how SVD entropy can be
+used to predict the network of maximum entropy (i.e.~of maximum
+complexity) heuristically given a predetermined set of constraints.
+
+\hypertarget{data-and-code}{%
+\section{Data and code}\label{data-and-code}}
 
@@
 
-\hypertarget{null-models-types-i-and-ii}{%
-\subsection{Null models (types I and
-II)}\label{null-models-types-i-and-ii}}
-
-Our maximum entropy network models (second approach only) were compared
-with two topological null models. The first is the type I null model of
-Fortuna and Bascompte (2006), in which the probability that a species
-\(i\) predates on another species \(j\) is given by
-
-\begin{equation}\protect\hypertarget{eq:type1null}{}{p(i \rightarrow j) = \frac{L}{S^2}.}\label{eq:type1null}\end{equation}
-
-The second is the type II null model of Bascompte et al. (2003), in
-which the probability of interaction is instead given by
-
-\begin{equation}\protect\hypertarget{eq:type2null}{}{p(i \rightarrow j) = \frac{1}{2} \left(\frac{k_{in}(j)}{S} + \frac{k_{out}(i)}{S}\right),}\label{eq:type2null}\end{equation}
-
-where \(k_{in}\) and \(k_{out}\) are the in and out-degrees,
-respectively. The type I null model is based on connectance, whereas the
-type II null model is based on the joint degree sequence. Therefore, the
-type I and II topological null models correspond with our type I and II
-MaxEnt network models, respectively, since they use similar constraints.
-
-We predicted both types of null networks for all empirical networks in
-our complete dataset (\(n = 257\)). We converted all probabilistic
-networks to Boolean networks by generating \(100\) random Boolean
-networks for each of these probabilistic webs. Then, we counted the
-number of times each interaction was sampled, and kept the \(L\) entries
-that were drawn the most amount of time, with \(L\) given by the number
-of interactions in each food web. This ensured that the resulting null
-networks had the same number of interactions as their empirical
-counterparts.
-
-\hypertarget{neutral-model}{%
-\subsection{Neutral model}\label{neutral-model}}
-
-We also compared our MaxEnt network models with a neutral model of
-relative abundances, in which the probabilities of interaction are given
-by
-
-\begin{equation}\protect\hypertarget{eq:neutralmodel}{}{p(i \rightarrow j) \propto \frac{n_i}{N} \times \frac{n_j}{N},}\label{eq:neutralmodel}\end{equation}
-
-where \(n_i\) and \(n_j\) are the abundances (or biomass) of both
-species, and \(N\) is the total abundance (or biomass) of all species in
-the network. We predicted neutral abundance matrices for all empirical
-networks in our abundance dataset (\(n = 19\)), and converted these
-weighted matrices to Boolean networks using an approach analogue to the
-one we used for our null models.
-
 \hypertarget{analytical-models-measures-of-maximum-entropy}{%
@@
 multipliers. The probability distribution that maximizes entropy is
-obtained by finding these values. As pointed out in Box 1, \(F\) is just
-Shannon's entropy to which we added terms that each sums to zero (our
-constraints). \(F\) is maximized by setting to 0 its partial derivative
-with respect to \(p(k_{in},k_{out})\). Because the derivative of a
-constant is zero, this gives us:
+obtained by finding these values. As pointed out in the introduction,
+\(F\) is just Shannon's entropy to which we added terms that each sums
+to zero (our constraints). \(F\) is maximized by setting to 0 its
+partial derivative with respect to \(p(k_{in},k_{out})\). Because the
+derivative of a constant is zero, this gives us:
 
@@
 results show that, under our purely information-theoretic model, the
-probability that a species is isolated is quite high below this
-threshold. The expected proportion of isolated species rapidly declines
-by orders of magnitude with increasing numbers of species and
-interactions.
+probability that a species is isolated is quite high when the total
+number of interactions is below this threshold. The expected proportion
+of isolated species rapidly declines by orders of magnitude with
+increasing numbers of species and interactions.
 
@@
 
-\hypertarget{box-2---working-with-predicted-numbers-of-interactions}{%
-\section{Box 2 - Working with predicted numbers of
-interactions}\label{box-2---working-with-predicted-numbers-of-interactions}}
+\hypertarget{box-1---working-with-predicted-numbers-of-interactions}{%
+\section{Box 1 - Working with predicted numbers of
+interactions}\label{box-1---working-with-predicted-numbers-of-interactions}}
 
@@
 
+\hypertarget{box-2---corresponding-null-and-neutral-models}{%
+\section{Box 2 - Corresponding null and neutral
+models}\label{box-2---corresponding-null-and-neutral-models}}
+
+\hypertarget{null-models-types-i-and-ii}{%
+\subsection{Null models (types I and
+II)}\label{null-models-types-i-and-ii}}
+
+Our maximum entropy network models (second approach only) were compared
+with two topological null models. The first is the type I null model of
+Fortuna and Bascompte (2006), in which the probability that a species
+\(i\) predates on another species \(j\) is given by
+
+\begin{equation}\protect\hypertarget{eq:type1null}{}{p(i \rightarrow j) = \frac{L}{S^2}.}\label{eq:type1null}\end{equation}
+
+The second is the type II null model of Bascompte et al. (2003), in
+which the probability of interaction is instead given by
+
+\begin{equation}\protect\hypertarget{eq:type2null}{}{p(i \rightarrow j) = \frac{1}{2} \left(\frac{k_{in}(j)}{S} + \frac{k_{out}(i)}{S}\right),}\label{eq:type2null}\end{equation}
+
+where \(k_{in}\) and \(k_{out}\) are the in and out-degrees,
+respectively. The type I null model is based on connectance, whereas the
+type II null model is based on the joint degree sequence. Therefore, the
+type I and II topological null models correspond with our type I and II
+MaxEnt network models, respectively, since they use similar constraints.
+
+We predicted both types of null networks for all empirical networks in
+our complete dataset (\(n = 257\)). We converted all probabilistic
+networks to Boolean networks by generating \(100\) random Boolean
+networks for each of these probabilistic webs. Then, we counted the
+number of times each interaction was sampled, and kept the \(L\) entries
+that were drawn the most amount of time, with \(L\) given by the number
+of interactions in each food web. This ensured that the resulting null
+networks had the same number of interactions as their empirical
+counterparts.
+
+\hypertarget{neutral-model}{%
+\subsection{Neutral model}\label{neutral-model}}
+
+We also compared our MaxEnt network models with a neutral model of
+relative abundances, in which the probabilities of interaction are given
+by
+
+\begin{equation}\protect\hypertarget{eq:neutralmodel}{}{p(i \rightarrow j) \propto \frac{n_i}{N} \times \frac{n_j}{N},}\label{eq:neutralmodel}\end{equation}
+
+where \(n_i\) and \(n_j\) are the abundances (or biomass) of both
+species, and \(N\) is the total abundance (or biomass) of all species in
+the network. We predicted neutral abundance matrices for all empirical
+networks in our abundance dataset (\(n = 19\)), and converted these
+weighted matrices to Boolean networks using an approach analogue to the
+one we used for our null models.
+
 \hypertarget{heuristical-models-networks-of-maximum-entropy}{%
@@
 adjacency matrix with the highest SVD entropy under a set of
-constraints. As mentioned in Box 1, we used the SVD entropy as our
-measure of entropy since it has been shown to be a reliable measure of
-food-web complexity (Strydom, Dalla Riva, and Poisot 2021), in addition
-to having the required properties of a proper measure of information
-entropy. We thus seek to find the network with the highest complexity,
-or randomness, that exactly reproduces specified constraints on its
+constraints. As mentioned above, we used the SVD entropy as our measure
+of entropy since it has been shown to be a reliable measure of food-web
+complexity (Strydom, Dalla Riva, and Poisot 2021), in addition to having
+the required properties of a proper measure of information entropy. We
+thus seek to find the network with the highest complexity, or
+randomness, that exactly reproduces specified constraints on its
 structure. Our method is in contrast with maximum entropy graph models

### 2022-07-20 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@e384fa6a51081659e616b4ff8043bde8292e4e30 🚀

@@
 values of \(L\) for different values of \(S\) using the joint posterior
-distribution of our model parameters, and calculated the mean degree for
-each simulated values. The resulting distributions are shown in the left
-panel of fig.~\ref{fig:degree_dist_fl} for three different values of
-species richness. In the right panel of fig.~\ref{fig:degree_dist_fl},
-we show how the probability distribution for the mean degree constraints
-can be used to generate a distribution of maximum entropy degree
-distributions, since each simulated value of mean degree generates a
-different maximum entropy degree distribution.
+distribution of our model parameters eq.~\ref{eq:BBpost}, and calculated
+the mean degree for each simulated values. The resulting distributions
+are shown in the left panel of fig.~\ref{fig:degree_dist_fl} for three
+different values of species richness. In the right panel of
+fig.~\ref{fig:degree_dist_fl}, we show how the probability distribution
+for the mean degree constraints can be used to generate a distribution
+of maximum entropy degree distributions, since each simulated value of
+mean degree generates a different maximum entropy degree distribution
+eq.~\ref{eq:lagrange_dd}.
 
@@
 predicted using the flexible links model fitted to all empirical
-networks in our complete dataset. (b) Degree distributions of maximum
-entropy for a network of \(27\) species and different numbers of
-interactions. The numbers of interactions correspond to the lower and
-upper bounds of the \(67%
+networks in our complete dataset eq.~\ref{eq:BBpost}. (b) Degree
+distributions of maximum entropy for a network of \(S=27\) species and
+different numbers of interactions \(L\). The numbers of interactions
+correspond to the lower and upper bounds of the \(67%
 \), \(89%
 \), and \(97%
-\) percentile intervals (PI), as well as the median, of the
-counterfactuals of the flexible links model.}\label{fig:degree_dist_fl}
+\) percentile intervals (PI), as well as the median (in blue), of the
+counterfactuals of the flexible links model. Each degree distribution
+was obtained using eq.~\ref{eq:lagrange_dd} after solving numerically
+eq.~\ref{eq:lagrange2_dd} using different values for the mean degree
+constraint \(2L/S\).}\label{fig:degree_dist_fl}
 }

### 2022-07-20 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@ce1bc28dd8ef4504c2c64d28c93cd48ccae6af91 🚀

@@
 values of \(L\) for different values of \(S\) using the joint posterior
-distribution of our model parameters eq.~\ref{eq:BBpost}, and calculated
-the mean degree for each simulated values. The resulting distributions
-are shown in the left panel of fig.~\ref{fig:degree_dist_fl} for three
-different values of species richness. In the right panel of
-fig.~\ref{fig:degree_dist_fl}, we show how the probability distribution
-for the mean degree constraints can be used to generate a distribution
-of maximum entropy degree distributions, since each simulated value of
-mean degree generates a different maximum entropy degree distribution
-eq.~\ref{eq:lagrange_dd}.
+distribution of our model parameters (eq.~\ref{eq:BBpost}), and
+calculated the mean degree for each simulated values. The resulting
+distributions are shown in the left panel of
+fig.~\ref{fig:degree_dist_fl} for three different values of species
+richness. In the right panel of fig.~\ref{fig:degree_dist_fl}, we show
+how the probability distribution for the mean degree constraints can be
+used to generate a distribution of maximum entropy degree distributions,
+since each simulated value of mean degree generates a different maximum
+entropy degree distribution (eq.~\ref{eq:lagrange_dd} and
+eq.~\ref{eq:lagrange2_dd}).
 
@@
 predicted using the flexible links model fitted to all empirical
-networks in our complete dataset eq.~\ref{eq:BBpost}. (b) Degree
+networks in our complete dataset (eq.~\ref{eq:BBpost}). (b) Degree
 distributions of maximum entropy for a network of \(S=27\) species and

### 2022-07-20 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@11ba21f9324a2b4bba7a74cddb24c7a98b456959 🚀

@@
 
-Shannon's entropy can only be calculated on conventional probability
-distributions such as the joint degree distribution. This is an issue
-when working with ecological networks, which are frequently represented
-as adjacency matrices. For this reason, in this contribution, we also
-use the SVD entropy as a measure of entropy, which is an application of
-Shannon's entropy to the relative non-zero singular values of a
-truncated singular value decomposition (t-SVD; Strydom, Dalla Riva, and
-Poisot 2021) of a food web's Boolean adjacency matrix. We measured SVD
-entropy as follows:
-
-\begin{equation}\protect\hypertarget{eq:svd-entropy}{}{J = -\sum_{i=1}^R s_i \log s_i,}\label{eq:svd-entropy}\end{equation}
-
-where \(s_i\) are the relative singular values of the adjacency matrix
-(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
-the singular values). Note that the distribution of relative singular
-values is analogous to a probability distribution, with \(0 < s_i < 1\)
-and \(\sum s_i = 1\). Thus, this measure also satisfies all four
-properties above-mentioned, while being a proper measure of the internal
-complexity of food webs (Strydom, Dalla Riva, and Poisot 2021).
-Following Strydom, Dalla Riva, and Poisot (2021), we standardized this
-measure with the rank \(R\) of the matrix (i.e.~\(J / \ln(R)\)) to
-account for the difference in dimensions between networks (\emph{sensu}
-Pielou's evenness; Pielou 1975). We will show how SVD entropy can be
-used to predict the network of maximum entropy (i.e.~of maximum
-complexity) heuristically given a predetermined set of constraints.
-
 \hypertarget{data-and-code}{%
@@
 flexible method to predict many measures of network structure
-simultaneously, i.e.~by finding networks of maximum entropy
-heuristically.
+simultaneously, i.e.~by finding heuristically the configuration of the
+network having maximum entropy given partial knowledge of its emerging
+structure.
 
@@
 
+\hypertarget{from-shannons-to-svd-entropy}{%
+\subsection{From Shannon's to SVD
+entropy}\label{from-shannons-to-svd-entropy}}
+
+The principle of maximum entropy can be applied on the network itself if
+we decompose its adjacency matrix into a non-zero vector of relative
+values. This is a necessary step when working with food webs, which are
+frequently expressed as a matrix \(A = [a_{ij}]\) of Boolean values
+representing the presence (\(a_{ij} = 1\)) or absence (\(a_{ij} = 0\))
+of an interaction between two species \(i\) and \(j\). Knowing one or
+many properties of a food web of interest (e.g., its number of species
+and number of interactions), we can simulate its adjacency matrix
+randomly by using these known ecological information to constrain the
+space of potential networks. The entropy of this hypothetical matrix can
+then be measured after decomposing it into appropriate values.
+Simulating a series of networks until we find the one having the highest
+entropy allows us to search for the most complex food-web configuration
+given the ecological constraints used. This configuration is the least
+biased one considering the information we had. In other words, the most
+we can say about a network's adjacency matrix, without making more
+assumptions than the ones given by our incomplete knowledge of its
+structure, is the one of maximum entropy. Generating the most complex
+network that corresponds to this structure allows us to explore more
+easily other properties of food webs under MaxEnt.
+
+Shannon's entropy can only be calculated on conventional probability
+distributions such as the joint degree distribution. This is an issue
+when working with ecological networks, which are represented as
+adjacency matrices. For this reason, we need to use another measure of
+entropy if we want to predict the network itself. We used the SVD
+entropy as our measure of entropy, which is an application of Shannon's
+entropy to the relative non-zero singular values of a truncated singular
+value decomposition (t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a
+food web's Boolean adjacency matrix. We measured SVD entropy as follows:
+
+\begin{equation}\protect\hypertarget{eq:svd-entropy}{}{J = -\sum_{i=1}^R s_i \log s_i,}\label{eq:svd-entropy}\end{equation}
+
+where \(s_i\) are the relative singular values of the adjacency matrix
+(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
+the singular values). Note that the distribution of relative singular
+values is analogous to a probability distribution, with \(0 < s_i < 1\)
+and \(\sum s_i = 1\). Thus, this measure also satisfies all four
+properties of an appropriate entropy measure (see introduction), while
+being a proper measure of the internal complexity of food webs (Strydom,
+Dalla Riva, and Poisot 2021). Following Strydom, Dalla Riva, and Poisot
+(2021), we standardized this measure with the rank \(R\) of the matrix
+(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
+between networks (\emph{sensu} Pielou's evenness; Pielou 1975). We used
+SVD entropy to predict the network configuration of maximum entropy
+(i.e.~of maximum complexity) heuristically given a predetermined set of
+constraints.
+
 \hypertarget{maxent-network-models-types-i-and-ii}{%
@@
 II)}\label{maxent-network-models-types-i-and-ii}}
-
-We define networks of maximum entropy as the configuration of the
-adjacency matrix with the highest SVD entropy under a set of
-constraints. As mentioned above, we used the SVD entropy as our measure
-of entropy since it has been shown to be a reliable measure of food-web
-complexity (Strydom, Dalla Riva, and Poisot 2021), in addition to having
-the required properties of a proper measure of information entropy. We
-thus seek to find the network with the highest complexity, or
-randomness, that exactly reproduces specified constraints on its
-structure. Our method is in contrast with maximum entropy graph models
-that predict a probability distribution on networks under soft or hard
-constraints (e.g., Park and Newman 2004; Cimini et al. 2019). We believe
-our approach to be more flexible, easier to compute, while allowing
-direct comparisons of empirical food webs with more complex networks
-with similar structure.
 
@@
 entropy using both types of MaxEnt network models.
+
+Our method is in contrast with maximum entropy graph models that predict
+a probability distribution on networks under soft or hard constraints
+(e.g., Park and Newman 2004; Cimini et al. 2019). We believe our
+approach to be more flexible, easier to compute, while allowing direct
+comparisons of empirical food webs with more complex networks with
+similar structure.
 

### 2022-07-20 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@2938f44d869e03e1f18265e29ad281d36f52bd1b 🚀

@@
 (i.e.~at the species (node), the interaction (edge) or the community
-(network) levels). For example, at the species level, MaxEnt can be used
-to predict the distribution of trophic levels among species, as well as
-the distribution of species' vulnerability (number of predators) and
-generality (number of preys). By contrast, at the interaction level,
-predictions can be made on the distribution of interaction strengths in
-weighted food webs. At the community level, it can give us a probability
-distribution of potential networks or of specific measures of their
-emerging structure. The applicability and performance of this method
-depend on the ecological information available and on our capacity to
-find the set of state variables that best represent natural systems and
-to translate them into appropriate statistical constraints. MaxEnt has
-been used in a wide range of disciplines, from thermodynamics to
-chemistry and biology (Martyushev and Seleznev 2006). It has also been
-proven useful in ecology, e.g.~in species distribution models (Phillips,
-Anderson, and Schapire 2006) and macroecological models (Harte et al.
-2008; Harte and Newman 2014). Maximizing a measure of entropy ensures
-that the derived probability distributions are unique and least biased
-under the set of constraints used. These constraints are built using
-state variables, i.e.~variables that represent the macrostate of the
-system. Having a validated maximum entropy model for the system at hand
-allows us to make rigorous predictions using a minimal amount of data,
-as well as helping us describe the most important factors driving that
-system.
+(network) levels). Because MaxEnt is a method of finding probability
+distributions that are least biased given partial knowledge about a
+system, these properties need to be represented probabilistically.
+Otherwise, some data transformation might be needed. For example, at the
+species level, MaxEnt can be used to predict the distribution of trophic
+levels among species, as well as the distribution of species'
+vulnerability (number of predators) and generality (number of preys). By
+contrast, at the interaction level, predictions can be made on the
+distribution of interaction strengths in weighted food webs. At the
+community level, it can give us a probability distribution of potential
+networks or of specific measures of their emerging structure. The
+applicability and performance of this method depend on the ecological
+information available and on our capacity to find the set of state
+variables that best represent natural systems and to translate them into
+appropriate statistical constraints. MaxEnt has been used in a wide
+range of disciplines, from thermodynamics to chemistry and biology
+(Martyushev and Seleznev 2006). It has also been proven useful in
+ecology, e.g.~in species distribution models (Phillips, Anderson, and
+Schapire 2006) and macroecological models (Harte et al. 2008; Harte and
+Newman 2014). The probability distribution of the variable of interest
+with the maximum entropy value is unique and least biased under the set
+of constraints used. These constraints are built using state variables,
+i.e.~variables that represent the macrostate of the system. Having a
+validated maximum entropy model for the system at hand allows us to make
+rigorous predictions using a minimal amount of data, as well as helping
+us describe the most important factors driving that system.
 
@@
 comparison been little used in network ecology. Like other real system
-networks, ecological networks are represented mathematically as graphs.
-However, the very nature of ecological networks (directed simple graphs
-frequently having self-loops) makes the mathematical optimization of
-maximum entropy graph models more complicated than with many other types
-of (non-ecological) networks. MaxEnt has nevertheless been used to
-predict the degree distribution of bipartite ecological networks from
-the number of species and the number of interactions (Williams 2011) and
-to predict interaction strengths between species pairs using their
-relative abundances within an optimal transportation theory regularized
-with information entropy (Stock, Poisot, and De Baets 2021). However, to
-the best of our knowledge, MaxEnt has never been used to predict
-food-web structure directly, even though food webs are among the most
-documented and widespread ecological networks.
+networks, food webs and other ecological networks are represented
+mathematically as graphs. However, the very nature of food webs
+(directed simple graphs frequently having self-loops) makes the
+mathematical optimization of maximum entropy graph models more
+complicated than with many other types of (non-ecological) networks.
+MaxEnt has nevertheless been used to predict the degree distribution of
+bipartite ecological networks from the number of species and the number
+of interactions (Williams 2011) and to predict interaction strengths
+between species pairs using their relative abundances within an optimal
+transportation theory regularized with information entropy (Stock,
+Poisot, and De Baets 2021). However, to the best of our knowledge,
+MaxEnt has never been used to predict food-web structure directly, even
+though food webs are among the most documented and widespread ecological
+networks.
 
@@
 consists in deriving constrained probability distributions of given
-network properties directly. We derived the joint degree distribution (a
-probability distribution) of maximum entropy using only the number of
-species \(S\) and the number of interactions \(L\) as state variables.
-Then, we predicted the degree distribution of maximum entropy directly
-from the joint degree distribution since the first is the sum of the
-marginal distributions of the second (a species' degree is the sum of
-its in and out-degrees). Because of the scarcity of empirical data on
-the number of interactions in ecological networks, we present a method
-to predict \(L\) from \(S\) (Box 1), thus allowing the prediction of the
-joint degree distribution from \(S\) solely. In turn, the second
+network properties directly. We derived the joint degree distribution
+(the joint probability distribution that a species has a given number of
+prey and predators in a food web) of maximum entropy using only the
+number of species \(S\) and the number of interactions \(L\) as state
+variables. Then, we predicted the degree distribution of maximum entropy
+directly from the joint degree distribution since the first is the sum
+of the marginal distributions of the second (a species' degree is the
+sum of its in and out-degrees). Because of the scarcity of empirical
+data on the number of interactions in ecological networks, we present a
+method to predict \(L\) from \(S\) (Box 1), thus allowing the prediction
+of the joint degree distribution from \(S\) solely. In turn, the second
 approach consists in finding, using different constraints, the adjacency
@@
 To do so, we used a flexible and heuristic approach based on simulated
-annealing to find networks \emph{close} to maximum entropy. As discussed
-above, our choice of algorithm stands from the very nature of food webs
-(i.e.~simple directed networks allowing self-loops) that makes the
-analytical derivation of a maximum entropy graph model difficult. We
-first built our type I MaxEnt network model constrained by the
-connectance of the network (i.e.~the ratio \(L/S^2\)). A comparison of
-this model against empirical data indicated that connectance alone was
-not sufficient to predict many aspects of network structure. For this
-reason, we built our type II MaxEnt network model, which instead uses
-the whole joint degree sequence as a constraint. Overall, we found that
-this second model was much better at predicting food-web structure than
-the one based on connectance.
+annealing (an optimization algorithm) to find networks \emph{close} to
+maximum entropy. As discussed above, our choice of algorithm stands from
+the very nature of food webs (i.e.~simple directed networks allowing
+self-loops) that makes the analytical derivation of a maximum entropy
+graph model difficult. We first built our type I MaxEnt network model
+constrained by the connectance of the network (i.e.~the ratio
+\(L/S^2\)). A comparison of this model against empirical data indicated
+that connectance alone was not sufficient to predict many aspects of
+network structure. For this reason, we built our type II MaxEnt network
+model, which instead uses the whole joint degree sequence as a
+constraint. Overall, we found that this second model was much better at
+predicting food-web structure than the one based on connectance.
 
@@
 
-Entropy measures the amount of information given by the outcome of a
-random variable. Many measures of entropy have been developed in physics
-(Beck 2009), but only a fraction of them could be used as an
-optimization measure with the principle of maximum entropy. According to
-Beck (2009) and Khinchin (2013), a measure of entropy \(H\) should
-satisfy four properties in the discrete case: (1) it should be a
-function of a probability distribution \(p(n)\) only; (2) it should be
-maximized when \(p(n)\) is uniform; (3) it should not be influenced by
-outcomes with a null probability; and (4) it should be independent of
-the order of information acquisition. The Shannon's entropy (Shannon
-1948)
+The least-biased probability distribution given the constraints used is
+the one with the highest entropy among all probability distributions
+that satisfy these constraints. Entropy is a measure of the average
+amount of information given by the outcome of a random variable. Many
+measures of entropy have been developed in physics (Beck 2009), but only
+a fraction of them could be used as an optimization measure with the
+principle of maximum entropy. According to Beck (2009) and Khinchin
+(2013), a measure of entropy \(H\) should satisfy four properties in the
+discrete case: (1) it should be a function of a probability distribution
+\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
+it should not be influenced by outcomes with a null probability; and (4)
+it should be independent of the order of information acquisition. The
+Shannon's entropy (Shannon 1948)
 
@@
 maximized by setting to \(0\) its partial derivative with respect to
-\(p(n)\). We will show how this can be done when we derive the joint
-degree distribution analytically from the number of species and the
-number of interactions in food webs.
+\(p(n)\). In this contribution, we show how this can be done by deriving
+the joint degree distribution analytically from the number of species
+and the number of interactions in food webs.
 
@@
 
-We tested our MaxEnt models (both approaches) against open food-web data
-queried from three different sources and integrated into what we call
-our \emph{complete dataset}. First, all food webs archived on
-\texttt{mangal.io} (Poisot et al. 2016; Banville, Vissault, and Poisot
-2021) were directly queried from the database (\(n = 235\)). Most
-ecological networks archived on Mangal are multilayer networks,
-i.e.~networks that describe different types of interactions. We kept all
-networks whose interactions were mainly of predation and herbivory
-types, and removed the largest network (\(S = 714\)) for computational
-efficiency reasons. Then, to this set we added food webs from two
-different sources: the New-Zealand dataset (\(n = 21\); Pomeranz et al.
-2018) and the Tuesday lake dataset (\(n = 2\); Cohen, Jonsson, and
-Carpenter 2003). Of these two datasets, \(19\) networks had data on
-species' relative abundances that were used in the neutral model
-presented in a following subsection. These networks are part of what we
-call our \emph{abundance dataset}, which is a subset of our complete
-dataset.
+We tested our analytical and heuristic MaxEnt models against open
+food-web data queried from three different sources and integrated into
+what we call our \emph{complete dataset}. These sources include (1)
+terrestrial and aquatic food webs sampled globally and archived on the
+ecological interactions database Mangal; (2) freshwater stream food webs
+from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
+United States). First, all food webs archived on \texttt{mangal.io}
+(Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were directly
+queried from the database (\(n = 235\)). Most ecological networks
+archived on Mangal are multilayer networks, i.e.~networks that describe
+different types of interactions. We kept all networks whose interactions
+were mainly of predation and herbivory types, and removed the largest
+network (\(S = 714\)) for computational efficiency reasons. Then, to
+this set we added food webs from two different sources: the New-Zealand
+dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
+(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
+\(19\) networks had data on species' relative abundances that were used
+in the neutral model (Box 2). These \(19\) networks constitute our
+\emph{abundance dataset}, which is a subset of our complete dataset.
 
@@
 
-Our maximum entropy network models (second approach only) were compared
-with two topological null models. The first is the type I null model of
-Fortuna and Bascompte (2006), in which the probability that a species
-\(i\) predates on another species \(j\) is given by
+Our heuristic MaxEnt models, predicting adjacency matrices, were
+compared with two topological null models. The first is the type I null
+model of Fortuna and Bascompte (2006), in which the probability that a
+species \(i\) predates on another species \(j\) is given by
 
@@
 
-\hypertarget{heuristical-models-networks-of-maximum-entropy}{%
-\section{Heuristical models: Networks of maximum
-entropy}\label{heuristical-models-networks-of-maximum-entropy}}
+\hypertarget{heuristic-models-networks-of-maximum-entropy}{%
+\section{Heuristic models: Networks of maximum
+entropy}\label{heuristic-models-networks-of-maximum-entropy}}
 
@@
 have a (non MaxEnt) probability distribution of networks when working
-with the entire chains after burn-in. For each network in our complete
-and abundance datasets, we estimated their configuration with maximum
-entropy using both types of MaxEnt network models.
+with the entire chains (or part of the chains). For each network in our
+complete and abundance datasets, we estimated their configuration with
+maximum entropy using both types of MaxEnt network models.
 

### 2022-07-21 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@a79df87cb25b51991d50fc9341f158dedd48620a 🚀

@@
 
-Our method is in contrast with maximum entropy graph models that predict
-a probability distribution on networks under soft or hard constraints
-(e.g., Park and Newman 2004; Cimini et al. 2019). We believe our
-approach to be more flexible, easier to compute, while allowing direct
-comparisons of empirical food webs with more complex networks with
-similar structure.
-
 \hypertarget{structure-of-maxent-networks}{%
@@
 
+\hypertarget{alternative-maxent-models}{%
+\subsection{Alternative MaxEnt models}\label{alternative-maxent-models}}
+
+There are different ways to generate food webs using MaxEnt. In this
+contribution, we used a method based on simulated annealing to find the
+network configuration with the highest SVD entropy while fixing some
+aspects of its structure. Another technique, also based on simulated
+annealing, could start by generating a food web randomly with fixed
+numbers of species and interactions and calculating its joint degree
+distribution. Pairs of interactions could be swapped sequentially until
+we minimize the divergence between the calculated joint degree
+distribution and the one of maximum entropy obtained analytically. In
+that case, this is the entropy of the joint degree distribution that
+would be maximized, not the one of the network's topology. To a certain
+extent, this method would thus bridge the gap between the two approaches
+presented in this article (i.e.~analytical and heuristic MaxEnt models).
+More research is needed to compare the quality of different methods
+generating adjacency matrices of food webs using MaxEnt.
+
+Another method generating adjacency matrices are maximum entropy graph
+models, which predict a probability distribution of networks under soft
+or hard constraints (e.g., Park and Newman 2004; Cimini et al. 2019).
+Under hard constraints, every network with a non-zero probability
+exactly satisfies the constraints on its structure. This is in contrast
+with soft constraints, which require that networks satisfy them on
+average (i.e.~many networks with a non-zero probability do not have the
+exact structure set by the constraints). Maximum entropy graph models
+are helpful because they can provide probability distributions for many
+network properties, by measuring the structure of all adjacency matrices
+with a non-zero probability. However, we consider that our approach
+based on simulated annealing is more flexible and more computationally
+efficient. Indeed, many measures of food-web structure are hard to
+translate into mathematical constraints. Moreover, because food webs are
+directed networks with self-loops, it makes the mathematical derivation
+of maximum entropy graph models difficult. We believe that identifying
+heuristically what really constrains the topology of food webs is a
+useful first step before attempting to derive the mathematical
+formulation of a maximum entropy graph model adapted to food webs.
+
+\hypertarget{applications}{%
+\subsection{Applications}\label{applications}}
+
 Our method and results could be used for different purposes. First, they

### 2022-07-21 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@a05ab93f84c626d14312205bd08c03a04957647d 🚀

@@
 species directly, we predicted the entire joint degree sequence without
-taking into account species' identity. When we associate predictions and
-empirical data according to their rank in total degrees, we see that
-species predicted to be have a higher generality (number of preys)
-generally have a lower vulnerability (number of preys) than what is
-observed (and conversely). In fig.~S1, we show how these differences
-change when species are instead ordered by their out-degrees (left
-panel) and in-degrees (right panel), respectively.
+taking into account species' identity. The challenge is to associate
+predictions with empirical data. If we associate each node according to
+their rank in total degrees, we see that species predicted to be have a
+higher generality (number of preys) generally have a lower vulnerability
+(number of preys) than what is observed (and conversely). This is
+because the difference in total degree (\(k_{out} + k_{in}\)) is
+minimized when species are ordered by their total degrees in both
+predicted and empirical sets (i.e.~the average sum of relative
+\(k_{out}\) and \(k_{in}\) is close to \(0\) across all species). This
+result shows that the differences between predicted and empirical
+degrees (in and out) are low for most species when ordered by their
+total degrees and that there is no apparent biases towards in or out
+degrees. In fig.~S1, we show how these differences change when species
+are instead ordered by their out-degrees (left panel) and in-degrees
+(right panel), respectively.
 
@@
 
-We plotted the Kullback--Leibler divergence between in and out-degree
-distributions to compare the symmetry of empirical and maximum entropy
-joint degree sequences fig.~\ref{fig:kl_diverg}. As we expected, our
-model predicted more similar in-degree and out-degree distributions than
-empirical data. However, this difference decreased with connectance
-right panel of \ref{fig:kl_diverg}. Overall, this suggests that other
-ecological constraints might be needed to account for the asymmetry of
-the joint degree distribution, especially for networks with a lower
-connectance.
+We can describe the shape of the joint degree distribution by comparing
+its marginal distributions with one another. Doing so allows us to
+evaluate how much the shape of the joint degree distribution of maximum
+entropy differs from empirical data. We plotted the Kullback--Leibler
+(KL) divergence between in and out-degree sequences to compare the
+symmetry of empirical and maximum entropy joint degree sequences
+fig.~\ref{fig:kl_diverg}. As we expected, our model predicted more
+similar in-degree and out-degree distributions than empirical data
+(shown by a lower KL divergence). However, this difference decreased
+with connectance right panel of \ref{fig:kl_diverg}. This suggests that
+other ecological constraints might be needed to account for the
+asymmetry of the joint degree distribution, especially for networks with
+a lower connectance. This might be due to the fact that food webs with
+low connectance are typically harder to predict than food webs with high
+connectance. Indeed, in low connectance systems, what makes two species
+interact might be more important for predictions than in high
+connectance systems, in which what prevents species from interacting
+might be more meaningful. This implies that more ecological information
+might be needed in food webs with fewer numbers of interactions because
+more nuances determine interactions compared to non-interactions.
 
@@
 corresponding null model (\(0.73 \pm 0.05\); empirical networks:
-\(0.63 \pm 0.09\)), as well as the proportion of cannibal species.
-However, the type II null model was better at predicting network
-diameter and average maximum similarity between species pairs.
-Predictions were similar between both type II models for the maximum
-trophic level and the proportion of omnivorous species.
+\(0.63 \pm 0.09\)), as well as the proportion of cannibal species. This
+might in part be due because nestedness was calculated using the
+spectral radius of the adjacency matrix, which leverages information on
+the network itself just like the MaxEnt network models. However, the
+type II null model was better at predicting network diameter and average
+maximum similarity between species pairs. Predictions were similar
+between both type II models for the maximum trophic level and the
+proportion of omnivorous species.
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 
+Our results bring to the forefront the role of the joint degree
+distribution in shaping food-web structure. This echoes the work of
+Williams (2011), which shows using MaxEnt how the degree distribution
+can drive the structure of bipartite networks. Fortuna et al. (2010)
+also found that the degree distribution of ecological networks is
+associated with some of their emerging properties such as nestedness and
+modularity. Network ecologists tend to focus on several emerging
+properties of food webs (Delmas et al. 2019), although we show that
+food-web structure is very well predicted by the joint degree
+distribution. In the light of our results, there is a lot more
+ecological information in the difference between properties of empirical
+networks and MaxEnt predictions than in these properties themselves.
+
 \hypertarget{alternative-maxent-models}{%
@@
 
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
+
 One of the biggest challenges in using the principle of maximum entropy

### 2022-08-08 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@2c9b6ff5150b9f0aa82a66d3a5535a4721e5b462 🚀

@@
 
+\hypertarget{the-constrained-structure-of-ecological-networks}{%
+\subsection{The constrained structure of ecological
+networks}\label{the-constrained-structure-of-ecological-networks}}
+
+A variety of measures of the structure of ecological networks have been
+used to describe the organization of species interactions in a
+biological community (Delmas et al. 2019). These measures provide
+valuable information on the functioning of ecosystems and their
+responses to environmental change (e.g., Pascual and Dunne 2006; Gómez,
+Perfectti, and Jordano 2011). For instance, Bascompte et al. (2003)
+suggest that plant--pollinator and seed-disperser networks have a highly
+nested structure that can promote species persistence. Another example,
+in food webs, shows that a high connectance can promote the robustness
+of the system to species lost (Dunne, Williams, and Martinez 2002).
+However, despite the growing literature on the ecological implications
+of network structure, the association between many of these measures
+impedes our ability to fully understand what drives the structure and
+behavior of ecological networks. In particular, nestedness and
+modularity are strongly associated in ecological networks (Fortuna et
+al. 2010), and network connectance has been shown to be an important
+driver of many other emerging network properties (Timothée Poisot and
+Gravel 2014). In light of these observations, it is difficult to assess
+whether attributed effects of given properties are the artifacts of
+other, perhaps simpler, measures. It might be necessary to factor out
+the effect of other correlated measures before analyzing the structure
+of ecological networks.
+
+One way to tackle this issue is first to recognize that food webs and
+other ecological networks are constrained systems. In other words, the
+space of possible network configurations shrinks as we know more about a
+network structure. Consequently, the range of values that a given
+network property can take depends on the values of other properties.
+Above all, the structure of ecological networks is constrained by the
+number of species, or nodes, present. Food webs with high species
+richness typically have a lower connectance (MacDonald, Banville, and
+Poisot 2020) than smaller networks. This is because the number of
+realized interactions in empirical food webs scales slower than the
+number of possible species pairs (MacDonald, Banville, and Poisot 2020).
+As shown by Timothée Poisot and Gravel (2014), connectance itself can
+constrain different aspects of network structure such as the degree
+distribution. Other measures, such as the maximum trophic level, can
+also constrain the space of feasible networks.
+
 Statistical and mathematical models can help fill many gaps in our
@@
 of feasible networks.
+
+\hypertarget{the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
+\subsection{The principle of maximum entropy: A primer for
+ecologists}\label{the-principle-of-maximum-entropy-a-primer-for-ecologists}}
 
@@
 
+The principle of maximum entropy is a mathematical method of finding
+probability distributions, strongly rooted in statistical mechanics and
+information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
+Starting from a set of constraints given by prior knowledge of a system
+(i.e.~what we call state variables), this method helps us find
+least-biased probability distributions subject to the constraints. These
+probability distributions are guaranteed to be unique given our prior
+knowledge and represent the most we can say about a system without
+making more assumptions. For example, if the only thing we know about a
+biological community is its average number of individuals per species,
+the least-biased inference we could make on its species abundance
+distribution is the exponential distribution (Frank and Smith 2011;
+Harte and Newman 2014). However, this does not imply that this
+distribution will be the best fit to empirical data. The challenge is to
+find the right set of constraints that would best reproduce
+distributions found in nature.
+
+The least-biased probability distribution given the constraints used is
+the one with the highest entropy among all probability distributions
+that satisfy these constraints. Entropy is a measure of the average
+amount of information given by the outcome of a random variable. Many
+measures of entropy have been developed in physics (Beck 2009), but only
+a fraction of them could be used as an optimization measure with the
+principle of maximum entropy. According to Beck (2009) and Khinchin
+(2013), a measure of entropy \(H\) should satisfy four properties in the
+discrete case: (1) it should be a function of a probability distribution
+\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
+it should not be influenced by outcomes with a null probability; and (4)
+it should be independent of the order of information acquisition. The
+Shannon's entropy (Shannon 1948)
+
+\begin{equation}\protect\hypertarget{eq:shannon}{}{H = -\sum_{n} p(n) \log p(n)}\label{eq:shannon}\end{equation}
+
+satisfies all of these properties. Finding the probability distribution
+\(p(n)\) that maximizes \(H\) under a set of \(m\) constraints \(g\) can
+be done using the method of Lagrange multipliers. These constraints
+could include one or many properties of the probability distribution
+(e.g., its mean, variance, and range). However, the normalization
+constraint always need to be included in \(g\) in order to make sure
+that \(p(n)\) sums to \(1\). The objective is then to find the values of
+the Lagrange multipliers \(\lambda_i\) that optimize a function \(F\):
+
+\begin{equation}\protect\hypertarget{eq:F}{}{F = H - \sum_{i=1}^m \lambda_i (g_i-c_i),}\label{eq:F}\end{equation}
+
+where \(g_i\) is the mathematical formulation of the constraint \(i\)
+and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
+which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
+maximized by setting to \(0\) its partial derivative with respect to
+\(p(n)\). In this contribution, we show how this can be done by deriving
+the joint degree distribution analytically from the number of species
+and the number of interactions in food webs.
+
+\hypertarget{objectives}{%
+\subsection{Objectives}\label{objectives}}
+
 Despite its extensive use in graph and network theories (e.g., Park and
@@
 
-\hypertarget{the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
-\subsection{The principle of maximum entropy: A primer for
-ecologists}\label{the-principle-of-maximum-entropy-a-primer-for-ecologists}}
-
-The principle of maximum entropy is a mathematical method of finding
-probability distributions, strongly rooted in statistical mechanics and
-information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
-Starting from a set of constraints given by prior knowledge of a system
-(i.e.~what we call state variables), this method helps us find
-least-biased probability distributions subject to the constraints. These
-probability distributions are guaranteed to be unique given our prior
-knowledge and represent the most we can say about a system without
-making more assumptions. For example, if the only thing we know about a
-biological community is its average number of individuals per species,
-the least-biased inference we could make on its species abundance
-distribution is the exponential distribution (Frank and Smith 2011;
-Harte and Newman 2014). However, this does not imply that this
-distribution will be the best fit to empirical data. The challenge is to
-find the right set of constraints that would best reproduce
-distributions found in nature.
-
-The least-biased probability distribution given the constraints used is
-the one with the highest entropy among all probability distributions
-that satisfy these constraints. Entropy is a measure of the average
-amount of information given by the outcome of a random variable. Many
-measures of entropy have been developed in physics (Beck 2009), but only
-a fraction of them could be used as an optimization measure with the
-principle of maximum entropy. According to Beck (2009) and Khinchin
-(2013), a measure of entropy \(H\) should satisfy four properties in the
-discrete case: (1) it should be a function of a probability distribution
-\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
-it should not be influenced by outcomes with a null probability; and (4)
-it should be independent of the order of information acquisition. The
-Shannon's entropy (Shannon 1948)
-
-\begin{equation}\protect\hypertarget{eq:shannon}{}{H = -\sum_{n} p(n) \log p(n)}\label{eq:shannon}\end{equation}
-
-satisfies all of these properties. Finding the probability distribution
-\(p(n)\) that maximizes \(H\) under a set of \(m\) constraints \(g\) can
-be done using the method of Lagrange multipliers. These constraints
-could include one or many properties of the probability distribution
-(e.g., its mean, variance, and range). However, the normalization
-constraint always need to be included in \(g\) in order to make sure
-that \(p(n)\) sums to \(1\). The objective is then to find the values of
-the Lagrange multipliers \(\lambda_i\) that optimize a function \(F\):
-
-\begin{equation}\protect\hypertarget{eq:F}{}{F = H - \sum_{i=1}^m \lambda_i (g_i-c_i),}\label{eq:F}\end{equation}
-
-where \(g_i\) is the mathematical formulation of the constraint \(i\)
-and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
-which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
-maximized by setting to \(0\) its partial derivative with respect to
-\(p(n)\). In this contribution, we show how this can be done by deriving
-the joint degree distribution analytically from the number of species
-and the number of interactions in food webs.
-
 \hypertarget{data-and-code}{%
@@
 United States). First, all food webs archived on \texttt{mangal.io}
-(Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were directly
-queried from the database (\(n = 235\)). Most ecological networks
-archived on Mangal are multilayer networks, i.e.~networks that describe
-different types of interactions. We kept all networks whose interactions
-were mainly of predation and herbivory types, and removed the largest
-network (\(S = 714\)) for computational efficiency reasons. Then, to
-this set we added food webs from two different sources: the New-Zealand
-dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
-(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
-\(19\) networks had data on species' relative abundances that were used
-in the neutral model (Box 2). These \(19\) networks constitute our
-\emph{abundance dataset}, which is a subset of our complete dataset.
+(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
+directly queried from the database (\(n = 235\)). Most ecological
+networks archived on Mangal are multilayer networks, i.e.~networks that
+describe different types of interactions. We kept all networks whose
+interactions were mainly of predation and herbivory types, and removed
+the largest network (\(S = 714\)) for computational efficiency reasons.
+Then, to this set we added food webs from two different sources: the
+New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
+lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these
+two datasets, \(19\) networks had data on species' relative abundances
+that were used in the neutral model (Box 2). These \(19\) networks
+constitute our \emph{abundance dataset}, which is a subset of our
+complete dataset.
 
@@
 
+\leavevmode\hypertarget{ref-Dunne2002NetStrb}{}%
+Dunne, Jennifer A., Richard J. Williams, and Neo D. Martinez. 2002.
+{``Network Structure and Biodiversity Loss in Food Webs: Robustness
+Increases with Connectance.''} \emph{Ecology Letters} 5 (4): 558--67.
+\url{https://doi.org/10.1046/j.1461-0248.2002.00354.x}.
+
 \leavevmode\hypertarget{ref-Dunning2017JumMod}{}%
@@
 
+\leavevmode\hypertarget{ref-Gomez2011FunConb}{}%
+Gómez, José M., Francisco Perfectti, and Pedro Jordano. 2011. {``The
+Functional Consequences of Mutualistic Network Architecture.''}
+\emph{PLOS ONE} 6 (1): e16143.
+\url{https://doi.org/10.1371/journal.pone.0016143}.
+
 \leavevmode\hypertarget{ref-Harremoes2001MaxEnt}{}%
@@
 
+\leavevmode\hypertarget{ref-Pascual2006EcoNeta}{}%
+Pascual, Department of Ecology and Evolutionary Biology Mercedes, and
+Visiting Professor Jennifer A. Dunne. 2006. \emph{Ecological Networks:
+Linking Structure to Dynamics in Food Webs}. Oxford University Press,
+USA.
+
 \leavevmode\hypertarget{ref-Phillips2006MaxEnta}{}%
@@
 \url{https://doi.org/10.1111/ecog.00976}.
+
+\leavevmode\hypertarget{ref-Poisot2014WheEcoa}{}%
+Poisot, Timothée, and Dominique Gravel. 2014. {``When Is an Ecological
+Network Complex? Connectance Drives Degree Distribution and Emerging
+Network Properties.''} \emph{PeerJ} 2: e251.
+\url{https://doi.org/10.7717/peerj.251}.
 

### 2022-08-09 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@19d988557e6f7228fa28bc84dd766f7a35ae8383 🚀

@@
 whether attributed effects of given properties are the artifacts of
-other, perhaps simpler, measures. It might be necessary to factor out
-the effect of other correlated measures before analyzing the structure
-of ecological networks.
+other, perhaps simpler, measures.
 
@@
 space of possible network configurations shrinks as we know more about a
-network structure. Consequently, the range of values that a given
-network property can take depends on the values of other properties.
-Above all, the structure of ecological networks is constrained by the
-number of species, or nodes, present. Food webs with high species
-richness typically have a lower connectance (MacDonald, Banville, and
-Poisot 2020) than smaller networks. This is because the number of
-realized interactions in empirical food webs scales slower than the
-number of possible species pairs (MacDonald, Banville, and Poisot 2020).
-As shown by Timothée Poisot and Gravel (2014), connectance itself can
-constrain different aspects of network structure such as the degree
-distribution. Other measures, such as the maximum trophic level, can
-also constrain the space of feasible networks.
-
-Statistical and mathematical models can help fill many gaps in our
-knowledge about species interactions. Predictive and null models are two
-complementary types of models that have been developed in network
-ecology for this purpose. On one hand, predictive models can partially
-alleviate the Eltonian shortfall, which describes our current lack of
-knowledge on food webs and other ecological networks (Hortal et al.
-2015). A variety of such models have recently been developed using
-machine learning and other statistical tools, most of which are
-presented in Strydom et al. (2021). On the other hand, null models help
-us identify potential ecological mechanisms that drive species
-interactions. They do so by comparing empirical data with an unbiased
-distribution of measures generated using a set of rules that exclude the
-mechanism of interest (Fortuna and Bascompte 2006; Delmas et al. 2019).
-Both types of models are frequently topological, i.e.~they often predict
-the adjacency matrix or specific measures of network structure without
-taking into account species' identity. According to Strydom et al.
-(2021), these topological models could be used to make better
-predictions of pairwise species interactions by constraining the space
-of feasible networks.
+network structure. For example, there is a finite number of networks
+with specified numbers of nodes and edges. Indeed, the structure of
+ecological networks is first and above all constrained by the number of
+species, or nodes, present. Food webs with high species richness
+typically have a lower connectance (MacDonald, Banville, and Poisot
+2020) than smaller networks. This is because the number of realized
+interactions in empirical food webs scales slower than the number of
+possible species pairs (MacDonald, Banville, and Poisot 2020). As shown
+by Timothée Poisot and Gravel (2014), connectance itself can constrain
+different aspects of network structure such as the degree distribution
+(i.e.~the probability distribution of the number of interspecific
+interactions realized by a species). Other measures, such as the maximum
+trophic level, can also constrain the space of feasible networks.
+
+Prior knowledge on the structure of ecological networks is thus
+especially useful in the current context of data scarcity about species
+interactions (Jordano 2016). Indeed, this Eltonian shortfall (Hortal et
+al. 2015) can be partially alleviated using known information about an
+ecological network. As suggested by Strydom et al. (2021), network
+structure can be used to improve the prediction of pairwise species
+interactions when data is lacking by constraining the space of feasible
+networks. Similarly, partial knowledge on the structure of an ecological
+network can also be used to predict others of its properties by
+constraining their range of possible values. This is important given
+that many aspects of network structure cannot be measured empirically
+without data on pairwise species interactions, a prevailing situation
+worldwide (Timothée Poisot et al. 2021).
+
+Understanding the ecological constraints that shape species interactions
+networks and predicting their emerging structure are thus two
+complementary aims of network ecology. This distinction between
+understanding and predicting is essential when using statistical and
+mathematical models in network ecology and interpreting them. On one
+hand, null models help us identify potential ecological mechanisms that
+drive species interactions and constrain ecological networks. Null
+models generate a distribution for a target measure using a set of rules
+that exclude the mechanism of interest (Fortuna and Bascompte 2006;
+Delmas et al. 2019). The difference between the model and empirical data
+helps us evaluate the effect of this ecological process in nature. On
+the other hand, predictive models can help fill many gaps on species
+interactions data. A variety of such models have recently been developed
+using machine learning and other statistical tools, most of which are
+presented in Strydom et al. (2021). Given the constrained nature of
+ecological networks, it is surprising that the principle of maximum
+entropy, a rigorous mathematical method designed for both the analysis
+and prediction of constrained systems, has been barely used in network
+ecology.
 
@@
 
-The principle of maximum entropy (MaxEnt) is a statistical and
-topological model that can be used for both of these purposes, i.e.~to
-make predictions of network structure and to better understand processes
-shaping ecological networks. Food-web properties that can be derived
-using MaxEnt are varied and pertain to different elements of the network
-(i.e.~at the species (node), the interaction (edge) or the community
-(network) levels). Because MaxEnt is a method of finding probability
-distributions that are least biased given partial knowledge about a
-system, these properties need to be represented probabilistically.
-Otherwise, some data transformation might be needed. For example, at the
-species level, MaxEnt can be used to predict the distribution of trophic
-levels among species, as well as the distribution of species'
-vulnerability (number of predators) and generality (number of preys). By
-contrast, at the interaction level, predictions can be made on the
-distribution of interaction strengths in weighted food webs. At the
-community level, it can give us a probability distribution of potential
-networks or of specific measures of their emerging structure. The
-applicability and performance of this method depend on the ecological
-information available and on our capacity to find the set of state
-variables that best represent natural systems and to translate them into
-appropriate statistical constraints. MaxEnt has been used in a wide
-range of disciplines, from thermodynamics to chemistry and biology
-(Martyushev and Seleznev 2006). It has also been proven useful in
-ecology, e.g.~in species distribution models (Phillips, Anderson, and
-Schapire 2006) and macroecological models (Harte et al. 2008; Harte and
-Newman 2014). The probability distribution of the variable of interest
-with the maximum entropy value is unique and least biased under the set
-of constraints used. These constraints are built using state variables,
-i.e.~variables that represent the macrostate of the system. Having a
-validated maximum entropy model for the system at hand allows us to make
-rigorous predictions using a minimal amount of data, as well as helping
-us describe the most important factors driving that system.
-
-The principle of maximum entropy is a mathematical method of finding
-probability distributions, strongly rooted in statistical mechanics and
-information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
-Starting from a set of constraints given by prior knowledge of a system
-(i.e.~what we call state variables), this method helps us find
-least-biased probability distributions subject to the constraints. These
-probability distributions are guaranteed to be unique given our prior
-knowledge and represent the most we can say about a system without
-making more assumptions. For example, if the only thing we know about a
-biological community is its average number of individuals per species,
-the least-biased inference we could make on its species abundance
-distribution is the exponential distribution (Frank and Smith 2011;
-Harte and Newman 2014). However, this does not imply that this
-distribution will be the best fit to empirical data. The challenge is to
-find the right set of constraints that would best reproduce
-distributions found in nature.
-
-The least-biased probability distribution given the constraints used is
-the one with the highest entropy among all probability distributions
-that satisfy these constraints. Entropy is a measure of the average
-amount of information given by the outcome of a random variable. Many
-measures of entropy have been developed in physics (Beck 2009), but only
-a fraction of them could be used as an optimization measure with the
-principle of maximum entropy. According to Beck (2009) and Khinchin
-(2013), a measure of entropy \(H\) should satisfy four properties in the
-discrete case: (1) it should be a function of a probability distribution
-\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
-it should not be influenced by outcomes with a null probability; and (4)
-it should be independent of the order of information acquisition. The
-Shannon's entropy (Shannon 1948)
+The principle of maximum entropy (MaxEnt) is a mathematical method of
+finding probability distributions, strongly rooted in statistical
+mechanics and information theory (Jaynes 1957a, 1957b; Harremoës and
+Topsøe 2001). Starting from a set of constraints given by prior
+knowledge of a system (i.e.~what we call state variables), this method
+helps us find least-biased probability distributions subject to the
+constraints. These probability distributions are guaranteed to be unique
+given our prior knowledge and represent the most we can say about a
+system without making more assumptions. For example, if the only thing
+we know about a biological community is its average number of
+individuals per species, the least-biased inference we could make on its
+species abundance distribution is the exponential distribution (Frank
+and Smith 2011; Harte and Newman 2014). However, this does not imply
+that this distribution will be the best fit to empirical data. The
+challenge is to find the right set of constraints that would best
+reproduce distributions found in nature.
+
+MaxEnt states that the least-biased probability distribution given the
+constraints used is the one with the highest entropy among all
+probability distributions that satisfy these constraints. Entropy is a
+measure of the average amount of information given by the outcome of a
+random variable. Many measures of entropy have been developed in physics
+(Beck 2009), but only a fraction of them could be used as an
+optimization measure with the principle of maximum entropy. According to
+Beck (2009) and Khinchin (2013), a measure of entropy \(H\) should
+satisfy four properties in the discrete case: (1) it should be a
+function of a probability distribution \(p(n)\) only; (2) it should be
+maximized when \(p(n)\) is uniform; (3) it should not be influenced by
+outcomes with a null probability; and (4) it should be independent of
+the order of information acquisition. The Shannon's entropy (Shannon
+1948)
 
@@
 maximized by setting to \(0\) its partial derivative with respect to
-\(p(n)\). In this contribution, we show how this can be done by deriving
-the joint degree distribution analytically from the number of species
-and the number of interactions in food webs.
-
-\hypertarget{objectives}{%
-\subsection{Objectives}\label{objectives}}
-
-Despite its extensive use in graph and network theories (e.g., Park and
-Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
-comparison been little used in network ecology. Like other real system
-networks, food webs and other ecological networks are represented
-mathematically as graphs. However, the very nature of food webs
-(directed simple graphs frequently having self-loops) makes the
-mathematical optimization of maximum entropy graph models more
-complicated than with many other types of (non-ecological) networks.
-MaxEnt has nevertheless been used to predict the degree distribution of
-bipartite ecological networks from the number of species and the number
-of interactions (Williams 2011) and to predict interaction strengths
-between species pairs using their relative abundances within an optimal
-transportation theory regularized with information entropy (Stock,
-Poisot, and De Baets 2021). However, to the best of our knowledge,
-MaxEnt has never been used to predict food-web structure directly, even
-though food webs are among the most documented and widespread ecological
-networks.
+\(p(n)\).
+
+The principle of maximum entropy has been used in a wide range of
+disciplines, from thermodynamics, chemistry and biology (Martyushev and
+Seleznev 2006) to graph and network theory (e.g., Park and Newman 2004;
+van der Hoorn, Lippner, and Krioukov 2018). It has also been proven
+useful in ecology, e.g.~in species distribution models (Phillips,
+Anderson, and Schapire 2006) and macroecological models (Harte et al.
+2008; Harte and Newman 2014). In network ecology, MaxEnt has been used
+to predict the degree distribution of bipartite networks from the number
+of species and the number of interactions (Williams 2011) and to predict
+interaction strengths between species pairs using their relative
+abundances within an optimal transportation theory regularized with
+information entropy (Stock, Poisot, and De Baets 2021). Thus, to the
+best of our knowledge, MaxEnt has never been used to predict food-web
+structure directly, even though food webs are among the most documented
+and widespread ecological networks (Ings et al. 2009).
+
+Food-web properties that can be derived using MaxEnt are varied and
+pertain to different elements of the network (i.e.~at the species
+(node), the interaction (edge) or the community (network) levels).
+Because MaxEnt is a method of finding least-biased probability
+distributions given partial knowledge about a system, these properties
+need to be represented probabilistically. For example, at the species
+level, MaxEnt can be used to predict the distribution of trophic levels
+among species, as well as the distribution of species' vulnerability
+(number of predators) and generality (number of preys). By contrast, at
+the interaction level, predictions can be made on the distribution of
+interaction strengths in weighted food webs. At the community level, it
+can generate probability distributions of many measures of their
+emerging structure and of networks themselves (i.e.~a probability
+distribution that specific network configurations are realized given the
+model and constraints). Overall, the potential of this method in the
+study of food webs is broad. The applicability and performance of MaxEnt
+mostly depend on the ecological information available and on our
+capacity to find the set of state variables that best represent natural
+systems and to translate them into appropriate statistical constraints.
+Having a validated maximum entropy model for the system at hand allows
+us to make least-biased predictions using a minimal amount of data, as
+well as identify the most important ecological processes shaping that
+system.
+
+\hypertarget{analytical-and-heuristic-approaches}{%
+\subsection{Analytical and heuristic
+approaches}\label{analytical-and-heuristic-approaches}}
 
 In this contribution, we used two complementary approaches to predict
-the structure of food webs using the principle of maximum entropy. We
-then compared our predictions against empirical data and null and
-neutral models commonly used in network ecology. The first approach
-consists in deriving constrained probability distributions of given
-network properties directly. We derived the joint degree distribution
-(the joint probability distribution that a species has a given number of
-prey and predators in a food web) of maximum entropy using only the
-number of species \(S\) and the number of interactions \(L\) as state
-variables. Then, we predicted the degree distribution of maximum entropy
-directly from the joint degree distribution since the first is the sum
-of the marginal distributions of the second (a species' degree is the
-sum of its in and out-degrees). Because of the scarcity of empirical
-data on the number of interactions in ecological networks, we present a
+the structure of food webs using the principle of maximum entropy. The
+first approach consists in deriving constrained probability
+distributions of given network properties analytically, whereas the
+second approach consists in finding the adjacency matrix of maximum
+entropy heuristically, from which network properties can be measured. We
+compared our predictions against empirical data and null and neutral
+models commonly used in network ecology. We focus on deterministic and
+unweighted (Boolean) food webs in both approaches for data availability
+reasons. However, our framework can be applied to all types of
+ecological networks and a wide variety of measures.
+
+For the first approach (analytic), we focus on species level properties.
+Specifically, we derived the joint degree distribution (i.e.~the joint
+probability distribution that a species has a given number of prey and
+predators in its network) of maximum entropy using only the number of
+species \(S\) and the number of interactions \(L\) as state variables.
+Then, we predicted the degree distribution of maximum entropy directly
+from the joint degree distribution since the first is the sum of the
+marginal distributions of the second. Because of the scarcity of
+empirical data on the number of interactions in food webs, we present a
 method to predict \(L\) from \(S\) (Box 1), thus allowing the prediction
-of the joint degree distribution from \(S\) solely. In turn, the second
-approach consists in finding, using different constraints, the adjacency
-matrix of maximum entropy from which network properties can be measured.
-To do so, we used a flexible and heuristic approach based on simulated
-annealing (an optimization algorithm) to find networks \emph{close} to
-maximum entropy. As discussed above, our choice of algorithm stands from
-the very nature of food webs (i.e.~simple directed networks allowing
-self-loops) that makes the analytical derivation of a maximum entropy
-graph model difficult. We first built our type I MaxEnt network model
-constrained by the connectance of the network (i.e.~the ratio
-\(L/S^2\)). A comparison of this model against empirical data indicated
-that connectance alone was not sufficient to predict many aspects of
-network structure. For this reason, we built our type II MaxEnt network
-model, which instead uses the whole joint degree sequence as a
-constraint. Overall, we found that this second model was much better at
-predicting food-web structure than the one based on connectance.
+of the joint degree distribution from \(S\) solely.
+
+For the second approach (heuristic), we focus on network level
+properties. We used a flexible and heuristic model based on simulated
+annealing (an optimization algorithm) to find the network configuration
+\emph{close} to maximum entropy and measured its structure. We developed
+this heuristic model because the analytical derivation of a maximum
+entropy graph model of food webs is difficult, and because this model is
+readily applicable to other types of ecological networks and measures.
+Indeed, the mathematical representation of food webs (i.e.~directed
+simple graphs frequently having self-loops) makes the optimization of
+maximum entropy graph models more complicated than with many other types
+of (non-ecological) networks. In other words, deriving a probability
+distribution on the graphs themselves is difficult when working with
+food webs. We built two types of heuristic MaxEnt models depending on
+the constraint used. Our type I MaxEnt model uses the connectance of the
+network (i.e.~the ratio \(L/S^2\)) as a constraint, whereas our type II
+MaxEnt model uses the whole joint degree sequence as a constraint.
 
@@
 New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
-lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these
-two datasets, \(19\) networks had data on species' relative abundances
-that were used in the neutral model (Box 2). These \(19\) networks
-constitute our \emph{abundance dataset}, which is a subset of our
-complete dataset.
+Lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Our
+complete dataset thus contained a total of \(257\) food webs. Of the
+New-Zealand and Tuesday Lake datasets, \(19\) networks had data on
+species' relative abundances that were used in the neutral model (Box
+2). These \(19\) food webs constitute our \emph{abundance dataset},
+which is a subset of our complete dataset.
 
@@
 Science Framework. Data cleaning, simulations and analyses were
-conducted in Julia v1.5.4.
-
-\hypertarget{analytical-models-measures-of-maximum-entropy}{%
-\section{Analytical models: Measures of maximum
-entropy}\label{analytical-models-measures-of-maximum-entropy}}
+conducted in Julia v1.6.1.
+
+\hypertarget{analytical-maximum-entropy-models}{%
+\section{Analytical maximum entropy
+models}\label{analytical-maximum-entropy-models}}
 
@@
 include all food webs archived on Mangal, as well as the New-Zealand and
-Tuesday lake datasets (our complete dataset). The predicted joint degree
+Tuesday Lake datasets (our complete dataset). The predicted joint degree
 sequences were obtained after sampling one realization of the joint
@@
 empirical networks include all food webs archived on Mangal, as well as
-the New-Zealand and Tuesday lake datasets (our complete dataset). The
+the New-Zealand and Tuesday Lake datasets (our complete dataset). The
 joint degree sequence of simulated networks was obtained after sampling
@@
 
-\hypertarget{heuristic-models-networks-of-maximum-entropy}{%
-\section{Heuristic models: Networks of maximum
-entropy}\label{heuristic-models-networks-of-maximum-entropy}}
+\hypertarget{heuristic-maximum-entropy-models}{%
+\section{Heuristic maximum entropy
+models}\label{heuristic-maximum-entropy-models}}
 
@@
 food webs archived on Mangal, as well as the New Zealand and Tuesday
-lake food webs (our complete dataset). Null 1: Type I null model based
+Lake food webs (our complete dataset). Null 1: Type I null model based
 on connectance. MaxEnt 1: Type I MaxEnt network model based on
@@
 the measure is overestimated (underestimated) on average. Empirical
-networks are all New Zealand and Tuesday lake food webs with abundance
+networks are all New Zealand and Tuesday Lake food webs with abundance
 data (our abundance dataset). Null 1: Type I null model based on
@@
 webs. Empirical networks include all food webs archived on Mangal, as
-well as the New-Zealand and Tuesday lake datasets (our complete
+well as the New-Zealand and Tuesday Lake datasets (our complete
 dataset). Maximum entropy networks were estimated using a simulated
@@
 entropy, and null food webs. Empirical networks include all food webs
-archived on Mangal, as well as the New-Zealand and Tuesday lake datasets
+archived on Mangal, as well as the New-Zealand and Tuesday Lake datasets
 (our complete dataset). Maximum entropy networks were estimated using a
@@
 entropy food webs. Empirical networks include all food webs archived on
-Mangal, as well as the New-Zealand and Tuesday lake datasets (our
+Mangal, as well as the New-Zealand and Tuesday Lake datasets (our
 complete dataset). Maximum entropy networks were estimated using a
@@
 
+\leavevmode\hypertarget{ref-Ings2009RevEco}{}%
+Ings, Thomas C., José M. Montoya, Jordi Bascompte, Nico Blüthgen, Lee
+Brown, Carsten F. Dormann, François Edwards, et al. 2009. {``Review:
+Ecological Networks Beyond Food Webs.''} \emph{Journal of Animal
+Ecology} 78 (1): 253--69.
+\url{https://doi.org/10.1111/j.1365-2656.2008.01460.x}.
+
 \leavevmode\hypertarget{ref-Jaynes1957InfThe}{}%
@@
 \url{https://doi.org/10.1103/PhysRev.108.171}.
+
+\leavevmode\hypertarget{ref-Jordano2016SamNeta}{}%
+Jordano, Pedro. 2016. {``Sampling Networks of Ecological
+Interactions.''} \emph{Functional Ecology} 30 (12): 1883--93.
+\url{https://doi.org/10.1111/1365-2435.12763}.
 
@@
 \url{https://doi.org/10.1111/ecog.00976}.
+
+\leavevmode\hypertarget{ref-Poisot2021GloKno}{}%
+Poisot, Timothée, Gabriel Bergeron, Kevin Cazelles, Tad Dallas,
+Dominique Gravel, Andrew MacDonald, Benjamin Mercier, Clément Violet,
+and Steve Vissault. 2021. {``Global Knowledge Gaps in Species
+Interaction Networks Data.''} \emph{Journal of Biogeography} 48 (7):
+1552--63. \url{https://doi.org/10.1111/jbi.14127}.
 

### 2022-08-10 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1924f65396fc26278b64f69f504949fdb8302210 🚀

@@
 using machine learning and other statistical tools, most of which are
-presented in Strydom et al. (2021). Given the constrained nature of
-ecological networks, it is surprising that the principle of maximum
-entropy, a rigorous mathematical method designed for both the analysis
-and prediction of constrained systems, has been barely used in network
-ecology.
+presented in Strydom et al. (2021). However, given the constrained
+nature of ecological networks, it is surprising that the principle of
+maximum entropy, a rigorous mathematical method designed for both the
+analysis and prediction of constrained systems, has been barely used in
+network ecology.
 
@@
 abundances within an optimal transportation theory regularized with
-information entropy (Stock, Poisot, and De Baets 2021). Thus, to the
+information entropy (Stock, Poisot, and De Baets 2021). However, to the
 best of our knowledge, MaxEnt has never been used to predict food-web
@@
 mostly depend on the ecological information available and on our
-capacity to find the set of state variables that best represent natural
-systems and to translate them into appropriate statistical constraints.
-Having a validated maximum entropy model for the system at hand allows
-us to make least-biased predictions using a minimal amount of data, as
-well as identify the most important ecological processes shaping that
-system.
+capacity to find the right set of state variables that best represent
+natural systems and to translate them into appropriate statistical
+constraints. Having a validated maximum entropy model for the system at
+hand allows us to make least-biased predictions using a minimal amount
+of data, as well as identify the most important ecological processes
+shaping that system. In other words, MaxEnt is a valuable and
+underexploited tool in network ecology that allows us to better
+understand and predict the structure of ecological networks.
 
@@
 (Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
-directly queried from the database (\(n = 235\)). Most ecological
+directly queried from the database (\(N = 235\)). Most ecological
 networks archived on Mangal are multilayer networks, i.e.~networks that
@@
 Then, to this set we added food webs from two different sources: the
-New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
-Lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Our
+New-Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday
+Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
 complete dataset thus contained a total of \(257\) food webs. Of the

### 2022-08-11 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@481219c9ee03a62380abfc0b350bf453c57f21b2 🚀

@@
 that exclude the mechanism of interest (Fortuna and Bascompte 2006;
-Delmas et al. 2019). The difference between the model and empirical data
+Delmas et al. 2019). The deviation between the model and empirical data
 helps us evaluate the effect of this ecological process in nature. On
@@
 the largest network (\(S = 714\)) for computational efficiency reasons.
-Then, to this set we added food webs from two different sources: the
-New-Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday
-Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
-complete dataset thus contained a total of \(257\) food webs. Of the
-New-Zealand and Tuesday Lake datasets, \(19\) networks had data on
-species' relative abundances that were used in the neutral model (Box
-2). These \(19\) food webs constitute our \emph{abundance dataset},
-which is a subset of our complete dataset.
+Then, to this set we added food webs from two different sources: the New
+Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday Lake
+dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete
+dataset thus contained a total of \(257\) food webs. Of the New Zealand
+and Tuesday Lake datasets, \(19\) networks had data on species' relative
+abundances that were used in the neutral model (Box 2). These \(19\)
+food webs constitute our \emph{abundance dataset}, which is a subset of
+our complete dataset.
 
@@
 
+Our analytical approach is the most common way to use and develop
+maximum entropy models. As shown above, starting from a defined set of
+constraints, a mathematical expression for the target distribution is
+derived using the method of Lagrange multipliers. The derived
+distribution of maximum entropy is unique and least biased given the
+constraints used. Although we refer to this approach as analytic,
+finding the values of the Lagrange multipliers usually requires the use
+of numerical methods. Here we use MaxEnt to derive two species level
+properties in food webs: the joint degree distribution and the degree
+distribution. The degree distribution has driven the attention of
+ecologists because of its role in determining the assembly of ecological
+networks (Vázquez 2005), shaping their emerging structure (Fortuna et
+al. 2010), and understanding interaction mechanisms (Williams 2011). As
+noted above, although the degree distribution of maximum entropy has
+already been derived in bipartite networks (Williams 2011), we show in
+much greater details its mathematical derivation in food webs. But
+first, we derive the joint degree distribution, a related property that
+holds significantly more ecological information than the degree
+distribution.
+
 \hypertarget{joint-degree-distribution}{%
@@
 
-The joint degree distribution \(p(k_{in},k_{out})\) is a joint discrete
-probability distribution describing the probability that a species has
-\(k_{in}\) predators and \(k_{out}\) preys, with \(k_{in}\) and
-\(k_{out}\) \(\epsilon\) \([0, S]\). Basal species (e.g., plants) have a
-\(k_{out}\) of \(0\), whereas top predators have a \(k_{in}\) of \(0\).
-In contrast, the maximum number of preys and predators a species can
-have is set by the number of species \(S\) in the food web. Here we show
-how the joint degree distribution of maximum entropy can be obtained
-given knowledge of \(S\) and \(L\).
+The joint degree distribution \(p(k_{in},k_{out})\) of a food web with
+\(S\) species is a joint discrete probability distribution describing
+the probability that a species has \(k_{in}\) predators and \(k_{out}\)
+preys, with \(k_{in}\) and \(k_{out}\) \(\epsilon\) \([0, S]\). Basal
+species (e.g., plants) have a \(k_{out}\) of \(0\), whereas top
+predators have a \(k_{in}\) of \(0\). In contrast, the maximum number of
+preys and predators a species can have is set by the number of species
+in the food web. Here we show how the joint degree distribution of
+maximum entropy can be obtained given knowledge of the number of species
+\(S\) and the number of interactions \(L\).
 
@@
 \(\langle k_{in} \rangle = \langle k_{out} \rangle\) because every edge
-is associated to a predator and a prey. Therefore, without any further
-constraints, we expect the joint degree distribution of maximum entropy
-to be a symmetric probability distribution with regards to \(k_{in}\)
-and \(k_{out}\). However, this does not mean that the joint degree
-\emph{sequence} will be symmetric, since the joint degree sequence is
-essentially a random realization of its probabilistic counterpart.
+is associated to a predator and a prey. Therefore, without using any
+further constraints, we would expect the joint degree distribution of
+maximum entropy to be a symmetric probability distribution with regards
+to \(k_{in}\) and \(k_{out}\). However, this does not mean that the
+joint degree \emph{sequence} will be symmetric, since the joint degree
+sequence is essentially a random realization of its probabilistic
+counterpart.
 
@@
 multipliers. The probability distribution that maximizes entropy is
-obtained by finding these values. As pointed out in the introduction,
-\(F\) is just Shannon's entropy to which we added terms that each sums
-to zero (our constraints). \(F\) is maximized by setting to 0 its
-partial derivative with respect to \(p(k_{in},k_{out})\). Because the
-derivative of a constant is zero, this gives us:
+obtained by finding these values. As pointed out above, \(F\) is just
+Shannon's entropy to which we added terms that each sums to zero (our
+constraints). \(F\) is maximized by setting to 0 its partial derivative
+with respect to \(p(k_{in},k_{out})\). Because the derivative of a
+constant is zero, this gives us:
 
@@
 
-We predicted the joint degree distribution of maximum entropy for each
-food web in our complete dataset, i.e.~using their numbers of species
-and numbers of interactions as state variables. We then sampled one
-realization of the degree sequence for each network using the
-probabilities given by the joint degree distribution. In
-fig.~\ref{fig:joint_dd} (left panels), we show the relationship between
-\(k_{out}\) and \(k_{in}\) standardized by the number of species in
-their networks, for empirical and maximum entropy joint degree
-distributions. We see that our model predicts higher values of
-generality and vulnerability compared to empirical food webs
-(i.e.~relative values of \(k_{out}\) and \(k_{in}\) closer to \(1\)).
-However, plotting the difference between predicted and empirical values
-for each species gives a different perspective. The right panel of
-fig.~\ref{fig:joint_dd} presents these differences when species are
-ordered by their total degree in their network (i.e.~by the sum of their
-in and out-degrees). Indeed, our predicted joint degree sequences have
-the same number of species as their empirical counterparts, but they are
-species agnostic; in other words, instead of predicting a value for each
-species directly, we predicted the entire joint degree sequence without
-taking into account species' identity. The challenge is to associate
-predictions with empirical data. If we associate each node according to
-their rank in total degrees, we see that species predicted to be have a
-higher generality (number of preys) generally have a lower vulnerability
-(number of preys) than what is observed (and conversely). This is
-because the difference in total degree (\(k_{out} + k_{in}\)) is
-minimized when species are ordered by their total degrees in both
-predicted and empirical sets (i.e.~the average sum of relative
+We assessed the empirical support of this expression using all food webs
+in our complete dataset. First, we predicted the joint degree
+distribution of maximum entropy for each of these food webs, i.e.~using
+their number of species and number of interactions as state variables.
+Then, we sampled one realization of the joint degree sequence for each
+network using the probabilities given by the joint degree distribution
+of maximum entropy, while fixing the total number of interactions. This
+gave us a random realization of the number of prey and predators for
+each species in each network. We standardized the predicted \(k_{out}\)
+and \(k_{in}\) of each species by the total number of species in their
+network to generate relative values, which can be compared across
+networks. In fig.~\ref{fig:joint_dd} (left panels), we show the
+relationship between these relative \(k_{out}\) and \(k_{in}\) obtained
+from the joint degree distributions of maximum entropy (bottom panel)
+and this relationship using empirical values (top panel). We observe
+that our model predicts higher values of generality and vulnerability
+compared to empirical food webs (i.e.~relative values of \(k_{out}\) and
+\(k_{in}\) both closer to \(1\)) for many species. In other words, our
+model predicts that species that have many predators also have more prey
+than what is observed empirically (and conversely). This is not
+surprising, given that our model did not include biological factors
+preventing generalist predators from having many prey. Nevertheless,
+with the exception of these generalist species, MaxEnt adequately
+predicts that most species have low generality and vulnerability values.
+
+Examining the difference between predicted and empirical values for each
+species gives a slightly different perspective (right panel of
+fig.~\ref{fig:joint_dd}). To do so, we must first associate each of our
+predictions to a specific species in a network in order to make that
+comparison. Indeed, our predicted joint degree sequences have the same
+number of species (elements) as their empirical counterparts, but they
+are species agnostic. In other words, instead of predicting a value for
+each species directly, we predicted the entire joint degree sequence
+without taking into account species' identity. The challenge is thus to
+adequately associate predictions with empirical data. In
+fig.~\ref{fig:joint_dd}, we present these differences when species are
+ordered by their total degree in their respective networks (i.e.~by the
+sum of their in and out-degrees). This means that the species with the
+highest total degree in its network will be associated with the highest
+prediction, and so forth. Doing so, we see that species predicted to
+have a higher number of predators than what is observed generally have a
+lower number of prey than what is observed (and conversely). This is
+because the difference in total degree (\(k_{out} + k_{in}\)) between
+predictions and empirical data is minimized when species are ranked by
+their total degree (i.e.~the average deviation of the sum of relative
 \(k_{out}\) and \(k_{in}\) is close to \(0\) across all species). This
-result shows that the differences between predicted and empirical
-degrees (in and out) are low for most species when ordered by their
-total degrees and that there is no apparent biases towards in or out
-degrees. In fig.~S1, we show how these differences change when species
-are instead ordered by their out-degrees (left panel) and in-degrees
-(right panel), respectively.
+result thus shows that the difference between predicted and empirical
+total degrees is low for most species when ordered by their total
+degrees. There are no apparent biases towards in or out degrees. In
+fig.~S1, we show how these differences change when species are instead
+ordered by their out-degrees (left panel) and in-degrees (right panel),
+respectively.
 
@@
 \includegraphics{figures/joint_degree_dist.png}
-\caption{Relative number of predators (\(k_{in}\)) as a function of the
-relative number of preys (\(k_{out}\)) for each species in (a) empirical
-and (b) maximum entropy joint degree sequences. Empirical networks
-include all food webs archived on Mangal, as well as the New-Zealand and
-Tuesday Lake datasets (our complete dataset). The predicted joint degree
-sequences were obtained after sampling one realization of the joint
-degree distribution of maximum entropy for each network. (c) Difference
-between predicted and empirical values when species are ordered
-according to their total degrees. In all panels, each dot corresponds to
-a single species in one of the network of our complete
-dataset.}\label{fig:joint_dd}
+\caption{Relative number of predators (\(k_{in}\)) as a function of
+their relative number of prey (\(k_{out}\)) for each species in (a)
+empirical and (b) joint degree sequences obtained from the analytical
+MaxEnt model. Empirical networks include most food webs archived on
+Mangal, as well as the New Zealand and Tuesday Lake datasets (our
+complete dataset). The predicted joint degree sequences were obtained
+after sampling one realization of the joint degree distribution of
+maximum entropy for each network, while keeping the total number of
+interactions constant. (c) Difference between predicted and empirical
+values when species are ordered according to their total degrees. In
+each panel, each dot corresponds to a single species in one of the
+networks.}\label{fig:joint_dd}
 }
@@
 
-We can describe the shape of the joint degree distribution by comparing
-its marginal distributions with one another. Doing so allows us to
-evaluate how much the shape of the joint degree distribution of maximum
-entropy differs from empirical data. We plotted the Kullback--Leibler
-(KL) divergence between in and out-degree sequences to compare the
-symmetry of empirical and maximum entropy joint degree sequences
-fig.~\ref{fig:kl_diverg}. As we expected, our model predicted more
-similar in-degree and out-degree distributions than empirical data
-(shown by a lower KL divergence). However, this difference decreased
-with connectance right panel of \ref{fig:kl_diverg}. This suggests that
-other ecological constraints might be needed to account for the
-asymmetry of the joint degree distribution, especially for networks with
-a lower connectance. This might be due to the fact that food webs with
-low connectance are typically harder to predict than food webs with high
-connectance. Indeed, in low connectance systems, what makes two species
-interact might be more important for predictions than in high
-connectance systems, in which what prevents species from interacting
-might be more meaningful. This implies that more ecological information
-might be needed in food webs with fewer numbers of interactions because
-more nuances determine interactions compared to non-interactions.
+Another way to evaluate the empirical support of the predicted joint
+degree sequences is to compare their shape with the ones of empirical
+food webs. We can describe the shape of a joint degree sequence by
+comparing its marginal distributions with one another. To do so, we
+calculated the Kullback--Leibler (KL) divergence between the in and
+out-degree sequences sampled from the joint degree distribution of
+maximum entropy. Similarly, we calculated the divergence between the in
+and out-degree sequences obtained empirically. This allows us to compare
+the symmetry of empirical and predicted joint degree sequences (left
+panel of fig.~\ref{fig:kl_diverg}). As we expected, our model predicts
+more similar in-degree and out-degree sequences than empirical data
+(shown by lower KL divergence values). However, this difference
+decreases with connectance (right panel of fig.~\ref{fig:kl_diverg}).
+This might be due to the fact that food webs with a low connectance are
+harder to predict than food webs with a high connectance. Indeed, in low
+connectance systems, what makes two species interact might be more
+important for prediction than in high connectance systems, in which what
+prevents species from interacting might be more meaningful. This implies
+that more ecological information might be needed in food webs with a low
+connectance because more ecological processes determine interactions
+compared to non-interactions. Therefore, other ecological constraints
+might be needed to account for the asymmetry of the joint degree
+distribution, especially for networks with a lower connectance. However,
+our MaxEnt model was able to capture quite well the shape of the joint
+degree sequence for networks having a high connectance.
 
@@
 \caption{(a) Probability density of KL divergence between in and
-out-degree sequences of empirical and maximum entropy joint degree
-sequences. (b) Difference between the KL divergence of empirical and
-predicted networks as a function of connectance. In both panels,
-empirical networks include all food webs archived on Mangal, as well as
-the New-Zealand and Tuesday Lake datasets (our complete dataset). The
-joint degree sequence of simulated networks was obtained after sampling
-one realization of the joint degree distribution of maximum entropy for
-each network.}\label{fig:kl_diverg}
+out-degree sequences of empirical and predicted joint degree sequences.
+(b) Difference between the KL divergence of empirical and predicted
+joint degree sequences as a function of connectance. In both panels,
+empirical networks include most food webs archived on Mangal, as well as
+the New Zealand and Tuesday Lake datasets (our complete dataset). The
+predicted joint degree sequences were obtained after sampling one
+realization of the joint degree distribution of maximum entropy for each
+network, while keeping the total number of interactions
+constant.}\label{fig:kl_diverg}
 }
@@
 (\(n = 257\)). Positive (negative) values indicate that the measure is
-overestimated (underestimated) on average. Empirical networks are all
+overestimated (underestimated) on average. Empirical networks are most
 food webs archived on Mangal, as well as the New Zealand and Tuesday
@@
 \caption{Motifs profile of empirical, maximum entropy, and null food
-webs. Empirical networks include all food webs archived on Mangal, as
-well as the New-Zealand and Tuesday Lake datasets (our complete
+webs. Empirical networks include most food webs archived on Mangal, as
+well as the New Zealand and Tuesday Lake datasets (our complete
 dataset). Maximum entropy networks were estimated using a simulated
@@
 \caption{Relationship between motifs proportions of empirical, maximum
-entropy, and null food webs. Empirical networks include all food webs
-archived on Mangal, as well as the New-Zealand and Tuesday Lake datasets
+entropy, and null food webs. Empirical networks include most food webs
+archived on Mangal, as well as the New Zealand and Tuesday Lake datasets
 (our complete dataset). Maximum entropy networks were estimated using a
@@
 \caption{Relationship between the structure of empirical and maximum
-entropy food webs. Empirical networks include all food webs archived on
-Mangal, as well as the New-Zealand and Tuesday Lake datasets (our
+entropy food webs. Empirical networks include most food webs archived on
+Mangal, as well as the New Zealand and Tuesday Lake datasets (our
 complete dataset). Maximum entropy networks were estimated using a
@@
 
+\leavevmode\hypertarget{ref-Vazquez2005DegDisa}{}%
+Vázquez, Diego P. 2005. {``Degree Distribution in Plantanimal
+Mutualistic Networks: Forbidden Links or Random Interactions?''}
+\emph{Oikos} 108 (2): 421--26.
+\url{https://doi.org/10.1111/j.0030-1299.2005.13619.x}.
+
 \leavevmode\hypertarget{ref-Williams2011BioMet}{}%

### 2022-08-11 suppmat.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@481219c9ee03a62380abfc0b350bf453c57f21b2 🚀

@@
 \caption{Difference between predicted and empirical values for the
-number of predators \(k_{in}\) and the number of preys \(k_{out}\) when
-species are ordered according to (a) their out-degree and (b) their
-in-degree. Empirical networks include all food webs archived on Mangal,
-as well as the New-Zealand and Tuesday lake datasets (our complete
-dataset). The joint degree sequences for simulated networks were
+relative number of predators \(k_{in}\) and the relative number of prey
+\(k_{out}\) when species are ordered according to (a) their out-degree
+and (b) their in-degree. Empirical networks include most food webs
+archived on Mangal, as well as the New Zealand and Tuesday lake datasets
+(our complete dataset). The predicted joint degree sequences were
 obtained after sampling one realization of the joint degree distribution
-of maximum entropy for each network. Each dot corresponds to a single
-species in one of the network of our dataset.}\label{fig:kin_kout_diff}
+of maximum entropy for each network, while keeping the total number of
+interactions constant. In each panel, each dot corresponds to a single
+species in one of the networks.}\label{fig:kin_kout_diff}
 }
@@
 II MaxEnt network model) and of empirical food webs. Empirical networks
-include all food webs archived on Mangal, as well as the New Zealand and
-Tuesday lake datasets (our complete dataset). Maximum entropy networks
-were derived using a simulating annealing algorithm to find the network
-of maximum SVD entropy while maintaining the joint degree sequence. (b)
-Distribution of z-scores of the SVD entropy of all empirical food webs.
-Z-scores were computed using the mean and standard deviation of the
-distribution of SVD entropy of MaxEnt food webs (type II MaxEnt network
-model). The dash line corresponds to the median
+include most food webs archived on Mangal, as well as the New Zealand
+and Tuesday lake datasets (our complete dataset). Maximum entropy
+networks were derived using a simulating annealing algorithm to find the
+network of maximum SVD entropy while maintaining the joint degree
+sequence. (b) Distribution of z-scores of the SVD entropy of all
+empirical food webs. Z-scores were computed using the mean and standard
+deviation of the distribution of SVD entropy of MaxEnt food webs (type
+II MaxEnt network model). The dash line corresponds to the median
 z-score.}\label{fig:entropy_dist}
@@
 richness, (b) the number of links, and (c) connectance. Empirical
-networks include all food webs archived on Mangal, as well as the New
+networks include most food webs archived on Mangal, as well as the New
 Zealand and Tuesday lake datasets (our complete dataset). Maximum
@@
 MaxEnt network model) as a function of species richness. Empirical
-networks include all food webs archived on Mangal, as well as the New
+networks include most food webs archived on Mangal, as well as the New
 Zealand and Tuesday lake datasets (our complete dataset). Maximum

### 2022-08-11 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@b53464e1b97a8e6f7721c7453fb60da2b54765ad 🚀

@@
 
-The degree distribution \(p(k)\) represents the probability that a
-species has \(k\) interactions in a food web, with
-\(k = k_{in} + k_{out}\). It can thus be directly obtained from the
-joint degree distribution:
+The joint degree distribution derived using MaxEnt
+(eq.~\ref{eq:lagrange3_jdd}) can be used to obtain the degree
+distribution of maximum entropy. Indeed, the degree distribution
+\(p(k)\) represents the probability that a species has \(k\)
+interactions in its food web, with \(k = k_{in} + k_{out}\). It can thus
+be obtained from the joint degree distribution as follows:
 
@@
 
-In fig.~S2, we show that the degree distribution of maximum entropy,
-given \(S\) and \(L\), predicts very low probabilities that a species
-will be isolated in its food web (i.e.~having \(k = 0\)). As MacDonald,
+The degree distribution could have also been obtained directly using the
+principle of maximum entropy, as discussed in Williams (2011). This
+gives the following distribution:
+
+\begin{equation}\protect\hypertarget{eq:lagrange_dd}{}{p(k) = \frac{e^{-\lambda_2k}}{Z},}\label{eq:lagrange_dd}\end{equation}
+
+with \(Z = \sum_{k=0}^S e^{-\lambda_2k}.\)
+
+This can be solved numerically using the constraint of the average
+degree \(\langle k \rangle = \frac{2L}{S}\) of a species, yielding an
+identical solution to the one obtained using the joint degree
+distribution as an intermediate. Note that the mean degree is twice the
+value of the linkage density, because every link must be counted twice
+when we add in and out-degrees together.
+
+\begin{equation}\protect\hypertarget{eq:lagrange2_dd}{}{\frac{1}{Z}\sum_{k=0}^S k e^{-\lambda_2k} = \frac{2L}{S}}\label{eq:lagrange2_dd}\end{equation}
+
+One aspect of the degree distribution that informs us of its ecological
+realism is the number of isolated species it predicts. As MacDonald,
 Banville, and Poisot (2020) pointed out, the size of food webs should at
 least be of \(S-1\) interactions, since a lower number would yield
-isolated species, i.e.~species without any predators or preys. Our
-results show that, under our purely information-theoretic model, the
-probability that a species is isolated is quite high when the total
-number of interactions is below this threshold. The expected proportion
-of isolated species rapidly declines by orders of magnitude with
-increasing numbers of species and interactions.
-
-The degree distribution could also have been obtained directly using the
-principle of maximum entropy, as discussed in Williams (2011). This
-gives the following distribution:
-
-\begin{equation}\protect\hypertarget{eq:lagrange_dd}{}{p(k) = \frac{e^{-\lambda_2k}}{Z},}\label{eq:lagrange_dd}\end{equation}
-
-with \(Z = \sum_{k=0}^S e^{-\lambda_2k}.\)
-
-This can be solved numerically using the constraint of the average
-degree \(\langle k \rangle = \frac{2L}{S}\) of a species. Note that the
-mean degree is twice the value of the linkage density, because every
-link must be counted twice when we add in and out-degrees together.
-
-\begin{equation}\protect\hypertarget{eq:lagrange2_dd}{}{\frac{1}{Z}\sum_{k=0}^S k e^{-\lambda_2k} = \frac{2L}{S}}\label{eq:lagrange2_dd}\end{equation}
-
-The numerical solution is identical to the one we obtained using the
-joint degree distribution as an intermediate. Ecologists wanting to
-model a system without considering isolated species could simply change
-the lower limit of \(k\) to \(1\) and solve the resulting equation
-numerically.
-
-In this section, we showed how important measures of food-web structure,
-namely the degree distribution and the joint degree distribution, could
-be derived with the principle of maximum entropy using minimal knowledge
-on a biological community. This type of models, although useful to make
-least-biased predictions on many network properties, can be hard to
-apply for other measures. Indeed, there are dozens of measures of
-network structure (Delmas et al. 2019) and many are not calculated with
-mathematical equations, but with algorithms. Moreover, the applicability
-of this method to empirical systems is limited by the state variables we
-can actually measure and use. In the next section, we propose a more
-flexible method to predict many measures of network structure
-simultaneously, i.e.~by finding heuristically the configuration of the
-network having maximum entropy given partial knowledge of its emerging
-structure.
+isolated species, i.e.~species without any predators or preys. Because
+non-basal species must eat to survive, isolated species could indicate
+that other species are missing or they could simply be removed from the
+food web. In fig.~S2, we show that the degree distribution of maximum
+entropy, given \(S\) and \(L\), gives very low probabilities that a
+species will be isolated in its food web (i.e.~having \(k = 0\)) above
+the \(S-1\) threshold. However, under our purely information-theoretic
+model, the probability that a species is isolated is quite high when the
+total number of interactions is below \(S-1\). Moreover, the expected
+proportion of isolated species rapidly declines by orders of magnitude
+with increasing numbers of species and interactions (fig.~S2). This
+supports the ecological realism of the degree distribution of maximum
+entropy derived above. Nevertheless, ecologists wanting to model a
+system without allowing isolated species could simply change the lower
+limit of \(k\) to \(1\) in eq.~\ref{eq:lagrange2_dd} and solve the
+resulting equation numerically.
+
+Using the analytical approach, we showed how important measures of
+food-web structure, namely the degree distribution and the joint degree
+distribution, could be derived with the principle of maximum entropy
+given minimal knowledge about a biological community. This type of
+models, although useful to make least-biased predictions on many network
+properties, can be hard to apply for other measures. Indeed, there are
+dozens of measures of network structure (Delmas et al. 2019) and many
+are not calculated with mathematical equations, but with algorithms.
+Moreover, the applicability of this method to empirical systems is
+limited by the state variables we can actually measure and use. In the
+next section, we propose a more flexible method to predict many measures
+of network structure simultaneously, i.e.~by finding heuristically the
+network configuration having maximum entropy given partial knowledge of
+its emerging structure.
 

### 2022-08-11 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@3311c55a5d9572bc262b0c1d5d170f7d02afe1ed 🚀

@@
 
-Our models need information on the number of species and the number of
-interactions. However, since the later is rarely estimated empirically,
-ecologists might need to use predictive methods to estimate the total
-number of interactions in a food web.
+Our analytical MaxEnt models require information on the number of
+species and the number of interactions. However, since the later is
+rarely measured empirically, ecologists might need to use predictive
+models to estimate the total number of interactions in a food web before
+using MaxEnt. Here we illustrate how this can be done by combining both
+models sequentially.
 
@@
 distribution of our model parameters (eq.~\ref{eq:BBpost}), and
-calculated the mean degree for each simulated values. The resulting
+calculated the mean degree for each simulated value. The resulting
 distributions are shown in the left panel of
@@
 \includegraphics{figures/maxent_degree_dist_fl.png}
-\caption{(a) Probability density of the predicted mean degree for
-different values of species richness. The number of interactions was
-predicted using the flexible links model fitted to all empirical
-networks in our complete dataset (eq.~\ref{eq:BBpost}). (b) Degree
-distributions of maximum entropy for a network of \(S=27\) species and
-different numbers of interactions \(L\). The numbers of interactions
-correspond to the lower and upper bounds of the \(67%
+\caption{(a) Probability density of the mean degree of a food web
+obtained using different values of species richness \(S\). The number of
+interactions \(L\) was simulated \(1000\) times using the flexible links
+model fitted to all empirical networks in our complete dataset
+(eq.~\ref{eq:BBpost}). The mean degrees \(2L/S\) were then obtained from
+these simulated values. (b) Degree distributions of maximum entropy for
+a network of \(S=27\) species and different numbers of interactions. The
+numbers of interactions correspond to the lower and upper bounds of the
+\(67%
 \), \(89%
@@
 \) percentile intervals (PI), as well as the median (in blue), of the
-counterfactuals of the flexible links model. Each degree distribution
-was obtained using eq.~\ref{eq:lagrange_dd} after solving numerically
-eq.~\ref{eq:lagrange2_dd} using different values for the mean degree
-constraint \(2L/S\).}\label{fig:degree_dist_fl}
+counterfactuals of the flexible links model. Each degree distribution of
+maximum entropy was obtained using eq.~\ref{eq:lagrange_dd} after
+solving numerically eq.~\ref{eq:lagrange2_dd} using different values of
+the mean degree constraint.}\label{fig:degree_dist_fl}
 }
@@
 
-Our heuristic MaxEnt models, predicting adjacency matrices, were
-compared with two topological null models. The first is the type I null
-model of Fortuna and Bascompte (2006), in which the probability that a
-species \(i\) predates on another species \(j\) is given by
+The predictions of our heuristic maximum entropy models were compared
+against two topological null models. These null models use the same
+ecological information as our heuristic models and thus constitute an
+adequate baseline for comparison. The first is the type I null model of
+Fortuna and Bascompte (2006), in which the probability that a species
+\(i\) predates on another species \(j\) is given by
 
@@
 
-where \(k_{in}\) and \(k_{out}\) are the in and out-degrees,
-respectively. The type I null model is based on connectance, whereas the
-type II null model is based on the joint degree sequence. Therefore, the
-type I and II topological null models correspond with our type I and II
-MaxEnt network models, respectively, since they use similar constraints.
-
-We predicted both types of null networks for all empirical networks in
-our complete dataset (\(n = 257\)). We converted all probabilistic
-networks to Boolean networks by generating \(100\) random Boolean
-networks for each of these probabilistic webs. Then, we counted the
-number of times each interaction was sampled, and kept the \(L\) entries
-that were drawn the most amount of time, with \(L\) given by the number
-of interactions in each food web. This ensured that the resulting null
-networks had the same number of interactions as their empirical
-counterparts.
+where \(k_{in}(j)\) and \(k_{out}(i)\) are the in and out-degrees of
+species \(j\) and \(i\), respectively. The type I null model is based on
+connectance, whereas the type II null model is based on the joint degree
+sequence. Therefore, the type I and II topological null models
+correspond to our type I and II heuristic MaxEnt models, respectively,
+since they use similar constraints.
+
+We generated probabilistic networks using both types of null models for
+all empirical food webs in our complete dataset. Then, we converted
+these networks to adjacency matrices of Boolean values by generating
+\(100\) random networks for each of these probabilistic webs. Finally,
+we kept the \(L\) entries that were sampled the most amount of times,
+with \(L\) given by the number of interactions in each food web. This
+ensured that the resulting null networks had the same number of
+interactions as their empirical counterparts. Thus, for each null model,
+we ended up with one null adjacency matrix for each empirical network.
 
@@
 
-We also compared our MaxEnt network models with a neutral model of
+We also compared our heuristic MaxEnt models with a neutral model of
 relative abundances, in which the probabilities of interaction are given
@@
 the network. We predicted neutral abundance matrices for all empirical
-networks in our abundance dataset (\(n = 19\)), and converted these
-weighted matrices to Boolean networks using an approach analogue to the
-one we used for our null models.
+food webs in our abundance dataset, and converted these weighted
+networks to adjacency matrices of Boolean values using the same method
+as the one we used for our null models.
 

### 2022-08-11 suppmat.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@3311c55a5d9572bc262b0c1d5d170f7d02afe1ed 🚀

@@
 \caption{Probability that a species is isolated in its food web
-according to the degree distribution of maximum entropy. We derived
+according to the degree distribution of maximum entropy. We derived many
 degree distributions of maximum entropy given a range of values of \(S\)
@@
 \(5\) and \(100\) species, by increment of \(5\) species. For each level
-of species richness, the numbers of links correspond to all 20-quantiles
-of the interval between \(0\) and \(S^2\). The black line marks the
-\(S-1\) minimum numbers of links required to have no isolated
-species.}\label{fig:heatmap}
+of species richness, the numbers of interactions correspond to all
+20-quantiles of the interval between \(0\) and \(S^2\). The black line
+marks the \(S-1\) minimum numbers of interactions required to have no
+isolated species.}\label{fig:heatmap}
 }
@@
 network model) and empirical food webs as a function of (a) species
-richness, (b) the number of links, and (c) connectance. Empirical
+richness, (b) the number of interactions, and (c) connectance. Empirical
 networks include most food webs archived on Mangal, as well as the New

### 2022-08-11 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@65f93fed31c2742c2277d044da061b61287dd750 🚀

@@
 among species, as well as the distribution of species' vulnerability
-(number of predators) and generality (number of preys). By contrast, at
+(number of predators) and generality (number of prey). By contrast, at
 the interaction level, predictions can be made on the distribution of
@@
 the probability that a species has \(k_{in}\) predators and \(k_{out}\)
-preys, with \(k_{in}\) and \(k_{out}\) \(\epsilon\) \([0, S]\). Basal
+prey, with \(k_{in}\) and \(k_{out}\) \(\epsilon\) \([0, S]\). Basal
 species (e.g., plants) have a \(k_{out}\) of \(0\), whereas top
 predators have a \(k_{in}\) of \(0\). In contrast, the maximum number of
-preys and predators a species can have is set by the number of species
-in the food web. Here we show how the joint degree distribution of
-maximum entropy can be obtained given knowledge of the number of species
-\(S\) and the number of interactions \(L\).
+prey and predators a species can have is set by the number of species in
+the food web. Here we show how the joint degree distribution of maximum
+entropy can be obtained given knowledge of the number of species \(S\)
+and the number of interactions \(L\).
 
@@
 least be of \(S-1\) interactions, since a lower number would yield
-isolated species, i.e.~species without any predators or preys. Because
+isolated species, i.e.~species without any predators or prey. Because
 non-basal species must eat to survive, isolated species could indicate
@@
 proportion of isolated species rapidly declines by orders of magnitude
-with increasing numbers of species and interactions (fig.~S2). This
-supports the ecological realism of the degree distribution of maximum
-entropy derived above. Nevertheless, ecologists wanting to model a
-system without allowing isolated species could simply change the lower
-limit of \(k\) to \(1\) in eq.~\ref{eq:lagrange2_dd} and solve the
-resulting equation numerically.
-
-Using the analytical approach, we showed how important measures of
-food-web structure, namely the degree distribution and the joint degree
-distribution, could be derived with the principle of maximum entropy
-given minimal knowledge about a biological community. This type of
-models, although useful to make least-biased predictions on many network
-properties, can be hard to apply for other measures. Indeed, there are
-dozens of measures of network structure (Delmas et al. 2019) and many
-are not calculated with mathematical equations, but with algorithms.
-Moreover, the applicability of this method to empirical systems is
-limited by the state variables we can actually measure and use. In the
-next section, we propose a more flexible method to predict many measures
-of network structure simultaneously, i.e.~by finding heuristically the
-network configuration having maximum entropy given partial knowledge of
-its emerging structure.
+with increasing numbers of species and interactions. This supports the
+ecological realism of the degree distribution of maximum entropy derived
+above. Nevertheless, ecologists wanting to model a system without
+allowing isolated species could simply change the lower limit of \(k\)
+to \(1\) in eq.~\ref{eq:lagrange2_dd} and solve the resulting equation
+numerically.
 
@@
 these networks to adjacency matrices of Boolean values by generating
-\(100\) random networks for each of these probabilistic webs. Finally,
-we kept the \(L\) entries that were sampled the most amount of times,
-with \(L\) given by the number of interactions in each food web. This
-ensured that the resulting null networks had the same number of
-interactions as their empirical counterparts. Thus, for each null model,
-we ended up with one null adjacency matrix for each empirical network.
+\(100\) random networks for each of these probabilistic webs, and kept
+the \(L\) entries that were sampled the most amount of times, with \(L\)
+given by the number of interactions in each food web. This ensured that
+the resulting null networks had the same number of interactions as their
+empirical counterparts. Thus, for each null model, we ended up with one
+null adjacency matrix for each empirical network.
 
@@
 species, and \(N\) is the total abundance (or biomass) of all species in
-the network. We predicted neutral abundance matrices for all empirical
+the network. We generated neutral abundance matrices for all empirical
 food webs in our abundance dataset, and converted these weighted
@@
 models}\label{heuristic-maximum-entropy-models}}
+
+With the analytical approach, we showed how important measures of
+food-web structure (e.g., the degree distribution and the joint degree
+distribution) can be derived with the principle of maximum entropy using
+minimal knowledge about a biological community. This type of models,
+although useful to make least-biased predictions of many network
+properties, can be hard to apply for other measures. Indeed, there are
+dozens of measures of network structure (Delmas et al. 2019) and many
+are not directly calculated with mathematical equations, but are instead
+estimated algorithmically. Moreover, the applicability of this method to
+empirical systems is limited by the state variables we can actually
+measure and use. Here, we propose a more flexible method to predict many
+measures of network structure simultaneously, i.e.~by finding
+heuristically the network configuration having maximum entropy given
+partial knowledge of its emerging structure.
 
@@
 given the ecological constraints used. This configuration is the least
-biased one considering the information we had. In other words, the most
-we can say about a network's adjacency matrix, without making more
+biased one considering the information available. In other words, the
+most we can say about a network's adjacency matrix, without making more
 assumptions than the ones given by our incomplete knowledge of its
@@
 distributions such as the joint degree distribution. This is an issue
-when working with ecological networks, which are represented as
-adjacency matrices. For this reason, we need to use another measure of
-entropy if we want to predict the network itself. We used the SVD
-entropy as our measure of entropy, which is an application of Shannon's
-entropy to the relative non-zero singular values of a truncated singular
-value decomposition (t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a
-food web's Boolean adjacency matrix. We measured SVD entropy as follows:
+when working with the adjacency matrix of ecological networks. For this
+reason, we need to use another measure of entropy if we want to predict
+a network's configuration directly using MaxEnt. We used the SVD entropy
+as our measure of entropy, which is an application of Shannon's entropy
+to the relative non-zero singular values of a truncated singular value
+decomposition (t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a food
+web's Boolean adjacency matrix. We measured SVD entropy as follows:
 
@@
 values is analogous to a probability distribution, with \(0 < s_i < 1\)
-and \(\sum s_i = 1\). Thus, this measure also satisfies all four
-properties of an appropriate entropy measure (see introduction), while
-being a proper measure of the internal complexity of food webs (Strydom,
-Dalla Riva, and Poisot 2021). Following Strydom, Dalla Riva, and Poisot
-(2021), we standardized this measure with the rank \(R\) of the matrix
+and \(\sum s_i = 1\). This measure also satisfies all four properties of
+an appropriate entropy measure above-mentioned, while being a proper
+measure of the internal complexity of food webs (Strydom, Dalla Riva,
+and Poisot 2021). Following Strydom, Dalla Riva, and Poisot (2021), we
+standardized this measure with the rank \(R\) of the matrix
 (i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
-between networks (\emph{sensu} Pielou's evenness; Pielou 1975). We used
-SVD entropy to predict the network configuration of maximum entropy
-(i.e.~of maximum complexity) heuristically given a predetermined set of
-constraints.
-
-\hypertarget{maxent-network-models-types-i-and-ii}{%
-\subsection{MaxEnt network models (types I and
-II)}\label{maxent-network-models-types-i-and-ii}}
-
-We built two types of MaxEnt network models: one based on connectance
-(type I MaxEnt network model) and the other based on the joint degree
-sequence (type II MaxEnt network model). They are based on the same
-constraints as the types I and II null models presented above. For both
-models, we used a simulated annealing algorithm with \(4\) chains,
-\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
-first generated one random Boolean matrix with the same order (number of
-species) as empirical webs, while maintaining the total number of
-interactions (type I MaxEnt network model) or rows and columns sums
-(type II MaxEnt network model). These are our initial configurations.
-Then, we swapped interactions sequentially while maintaining the
-original connectance or the joint degree sequence for types I and II
-MaxEnt network model, respectively. Configurations with a higher SVD
-entropy than the previous one in the chain were always accepted, whereas
-they were accepted with a probability conditional to a decreasing
-temperature when lower. The final configuration with the highest SVD
-entropy among the four chains constitute our estimated MaxEnt network.
-Even though we decided to work with point estimates, it is possible to
-have a (non MaxEnt) probability distribution of networks when working
-with the entire chains (or part of the chains). For each network in our
-complete and abundance datasets, we estimated their configuration with
-maximum entropy using both types of MaxEnt network models.
+between networks (\emph{sensu} Pielou's evenness; Pielou 1975).
+
+\hypertarget{types-i-and-ii-heuristic-maxent-models}{%
+\subsection{Types I and II heuristic MaxEnt
+models}\label{types-i-and-ii-heuristic-maxent-models}}
+
+We used SVD entropy to predict the network configuration of maximum
+entropy (i.e.~of maximum complexity) heuristically given different
+constraints for all food webs in our complete dataset. We built two
+types of heuristic MaxEnt models that differ on the constraint used. The
+type I heuristic MaxEnt model is based on connectance, whereas the type
+II heuristic MaxEnt model is based on the joint degree sequence. These
+models are thus based on the same constraints as the types I (Fortuna
+and Bascompte 2006) and II (Bascompte et al. 2003) null models (Box 2)
+frequently used to generate random networks topologically. This allows
+direct comparison of the performance of null and heuristic MaxEnt models
+in reproducing the emerging structure of empirical food webs.
+
+For each network in our complete dataset, we estimated their
+configuration of maximum entropy given each of these constraints. For
+both types of heuristic MaxEnt models, we used a simulated annealing
+algorithm with \(4\) chains, \(2000\) steps and an initial temperature
+of \(0.2\). For each food web, we first generated one random Boolean
+matrix per chain while fixing the number of species. We also maintained
+the total number of interactions (i.e.~the sum of all elements in the
+matrix) in the type I MaxEnt model and the joint degree sequence
+(i.e.~the rows and columns sums) in the type II MaxEnt model. These were
+our initial configurations. Then, we swapped interactions sequentially
+while maintaining the original connectance or joint degree sequence.
+Configurations with a higher SVD entropy than the previous one in the
+chain were always accepted, whereas they were accepted with a
+probability conditional to a decreasing temperature when lower. The
+final configuration with the highest SVD entropy among the four chains
+constitute the estimated maximum entropy configuration of a food web
+given the constraint used.
 
@@
 self loops), and the proportion of omnivorous species \emph{Omniv}
-(i.e.~species whose preys are of different trophic levels).
-\emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply defined in
-Williams and Martinez (2000).
+(i.e.~species whose prey are of different trophic levels). \emph{MxSim},
+\emph{Cannib}, and \emph{Omniv} are more deeply defined in Williams and
+Martinez (2000).
 
@@
 aspects of network structure found in nature. However, a model based on
-the entire joint degree sequence, i.e.~on the number of preys and
+the entire joint degree sequence, i.e.~on the number of prey and
 predators for each species, gave more convincing results. Our type II
@@
 food-web topology is configured almost entirely randomly around these
-marginal numbers of predators and preys.
+marginal numbers of predators and prey.
 
@@
 
+Even though we decided to work with point estimates, it is possible to
+have a (non MaxEnt) probability distribution of networks when working
+with the entire chains (or part of the chains).
+
 \hypertarget{applications}{%
@@
 is to identify the set of state variables that best reproduce empirical
-data. We found that the numbers of preys and predators for each species
+data. We found that the numbers of prey and predators for each species
 are important state variables for the prediction of maximum entropy

### 2022-08-12 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@fa572cda4b73fa82f7f24e762cbdf281e286e661 🚀

@@
 Then, to this set we added food webs from two different sources: the New
-Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday Lake
-dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete
-dataset thus contained a total of \(257\) food webs. Of the New Zealand
-and Tuesday Lake datasets, \(19\) networks had data on species' relative
-abundances that were used in the neutral model (Box 2). These \(19\)
-food webs constitute our \emph{abundance dataset}, which is a subset of
-our complete dataset.
+Zealand dataset (\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday
+Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
+complete dataset thus contained a total of \(257\) food webs. Of the New
+Zealand and Tuesday Lake datasets, \(19\) networks had data on species'
+relative abundances that were used in the neutral model (Box 2). These
+\(19\) food webs constitute our \emph{abundance dataset}, which is a
+subset of our complete dataset.
 
@@
 
-\hypertarget{structure-of-maxent-networks}{%
-\subsection{Structure of MaxEnt
-networks}\label{structure-of-maxent-networks}}
-
-We compared different measures of the structure of empirical food webs
-with the ones of null, neutral, and MaxEnt networks
-tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
-we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
-Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
-Nestedness indicates how much the diet of specialist species is a subset
-of the one of generalists, and is strongly associated with network
-modularity (Fortuna et al. 2010). We also measured the maximum trophic
-level \emph{maxtl}, network diameter \emph{diam} (i.e.~the longest of
-the shortest paths between all species pairs; Albert and Barabasi 2002),
-the average maximum similarity between species pairs \emph{MxSim}, the
-proportion of cannibal species \emph{Cannib} (i.e.~the proportion of
-self loops), and the proportion of omnivorous species \emph{Omniv}
-(i.e.~species whose prey are of different trophic levels). \emph{MxSim},
-\emph{Cannib}, and \emph{Omniv} are more deeply defined in Williams and
-Martinez (2000).
-
-Overall, we found that models based on the joint degree sequence (type
-II null and MaxEnt network models) reproduced the structure of empirical
-networks much better than the ones based on connectance and the neutral
-model of relative abundances. On average, the type II MaxEnt network
-model was better at predicting nestedness (\(0.62 \pm 0.08\)) than its
-corresponding null model (\(0.73 \pm 0.05\); empirical networks:
-\(0.63 \pm 0.09\)), as well as the proportion of cannibal species. This
-might in part be due because nestedness was calculated using the
-spectral radius of the adjacency matrix, which leverages information on
-the network itself just like the MaxEnt network models. However, the
-type II null model was better at predicting network diameter and average
-maximum similarity between species pairs. Predictions were similar
-between both type II models for the maximum trophic level and the
-proportion of omnivorous species.
+\hypertarget{structure-of-maxent-food-webs}{%
+\subsection{Structure of MaxEnt food
+webs}\label{structure-of-maxent-food-webs}}
+
+We measured various properties of these configurations of maximum
+entropy to evaluate how well they reproduce the structure of sampled
+food webs. Specifically, we evaluated their nestedness \(\rho\), their
+maximum trophic level \(maxtl\), their network diameter \(diam\), their
+average maximum similarity between species pairs \(MxSim\) (Williams and
+Martinez 2000), their proportion of cannibal species \(Cannib\), their
+proportion of omnivorous species \(Omniv\), their SVD entropy, and their
+motifs profile. Nestedness indicates how much the diet of specialist
+species is a subset of the one of generalists (Delmas et al. 2019) and
+was measured using the spectral radius of the adjacency matrix
+(Staniczenko, Kopp, and Allesina 2013). In turn, the network diameter
+represents the longest of the shortest paths between all species pair
+(Albert and Barabasi 2002). Further, cannibal species are species that
+can eat members of their own species (i.e.~species having self loops),
+whereas omnivorous species can prey on different trophic levels
+(Williams and Martinez 2000). Finally, a motifs profile represents the
+proportion of three-species motifs (subnetworks), which can be
+considered as simple building blocks of ecological networks (Milo et al.
+2002; Stouffer et al. 2007). All of these properties are relatively easy
+to measure and were chosen based on their ecological importance and
+prevalent use in network ecology (McCann 2011; Delmas et al. 2019).
+
+We compared the performance of both heuristic MaxEnt models in
+predicting these measures to the one of the null
+(tbl.~\ref{tbl:measures_all}) and neutral models
+(tbl.~\ref{tbl:measures_abund}). Overall, we found that the models based
+on the joint degree sequence (i.e.~the type II null and heuristic MaxEnt
+models) reproduced the structure of empirical food webs much better than
+the ones based on connectance (i.e.~the type I null and heuristic MaxEnt
+models). This suggests that the predictive power of connectance might be
+more limited than what was previously suggested (Timothée Poisot and
+Gravel 2014). On the other hand, the neutral model of relative
+abundances was surprisingly good at predicting the maximum trophic level
+and the network diameter. However, with the exception of the network
+diameter, the type II heuristic MaxEnt model was better at predicting
+network structure than the neutral model for all measures considered.
+This might be because, although neutral processes are important, they
+act in concert with niche processes in determining species interactions
+Canard et al. (2014). The joint degree sequence encodes information on
+both neutral and niche processes because the number of prey and
+predators a species has is determined by its relative abundance and
+biological traits. Therefore, these results show that having information
+on the number of prey and predators for each species substantially
+improves the prediction of food-web structure, both compared to models
+solely based on connectance and the ones solely based on species
+relative abundances.
+
+Next, the predictions of the type II heuristic MaxEnt model can be
+compared to its null model counterpart. On average, the type II
+heuristic MaxEnt model was better at predicting nestedness
+(\(0.62 \pm 0.08\)) than its corresponding null model
+(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for all
+networks in our complete dataset (tbl.~\ref{tbl:measures_all}). This
+might in part be due to the fact that nestedness was calculated using
+the spectral radius of the adjacency matrix, which directly leverages
+information on the network itself just like the heuristic MaxEnt models.
+The proportion of self-loops (cannibal species) was also better
+predicted by the type II heuristic MaxEnt model in comparison to the
+type II null model. However, the type II null model was better at
+predicting network diameter and average maximum similarity between
+species pairs, and predictions of the maximum trophic level and the
+proportion of omnivorous species were similar between both types of
+models. This suggests that, although the joint degree sequence is an
+important driver of food-web structure, other ecological constraints
+might need to be added to our MaxEnt model to account for some emerging
+food-web properties, especially the ones regarding food-chain lengths.
 
@@
 \begin{longtable}[]{@{}rrrrrrrr@{}}
-\caption{\label{tbl:measures_all}Standardized mean differences of
-predicted network measures with empirical data for all networks
-(\(n = 257\)). Positive (negative) values indicate that the measure is
-overestimated (underestimated) on average. Empirical networks are most
-food webs archived on Mangal, as well as the New Zealand and Tuesday
-Lake food webs (our complete dataset). Null 1: Type I null model based
-on connectance. MaxEnt 1: Type I MaxEnt network model based on
+\caption{\label{tbl:measures_all}Standardized mean difference between
+predicted network measures and empirical data for all food webs in our
+complete dataset (\(N = 257\)). Positive (negative) values indicate that
+the measure is overestimated (underestimated) on average. Empirical
+networks include most food webs archived on Mangal, as well as the New
+Zealand and Tuesday Lake food webs. Null 1: Type I null model based on
+connectance. MaxEnt 1: Type I heuristic MaxEnt model based on
 connectance. Null 2: Type II null model based on the joint degree
-sequence. MaxEnt 2: Type II MaxEnt network model based on the joint
+sequence. MaxEnt 2: Type II heuristic MaxEnt model based on the joint
 degree sequence. \(\rho\): nestedness measured by the spectral radius of
-the adjacency matrix. \emph{maxtl}: maximum trophic level. \emph{diam}:
-network diameter. \emph{MxSim}: average maximum similarity between
-species pairs. \emph{Cannib}: proportion of cannibal species (self
-loops). \emph{Omniv}: proportion of omnivorous species. \emph{entropy}:
-SVD entropy.}\tabularnewline
+the adjacency matrix. \(maxtl\): maximum trophic level. \(diam\):
+network diameter. \(MxSim\): average maximum similarity between species
+pairs. \(Cannib\): proportion of cannibal species (self loops).
+\(Omniv\): proportion of omnivorous species. \emph{entropy}: SVD
+entropy.}\tabularnewline
 \toprule
@@
 \begin{longtable}[]{@{}rrrrrrrr@{}}
-\caption{\label{tbl:measures_abund}Standardized mean differences of
-predicted network measures with empirical data for networks having
-abundance data (\(n = 19\)). Positive (negative) values indicate that
+\caption{\label{tbl:measures_abund}Standardized mean difference between
+predicted network measures and empirical data for all food webs in our
+abundance dataset (\(N = 19\)). Positive (negative) values indicate that
 the measure is overestimated (underestimated) on average. Empirical
-networks are all New Zealand and Tuesday Lake food webs with abundance
-data (our abundance dataset). Null 1: Type I null model based on
-connectance. MaxEnt 1: Type I MaxEnt network model based on connectance.
-Null 2: Type II null model based on the joint degree sequence. MaxEnt 2:
-Type II MaxEnt network model based on the joint degree sequence.
-\(\rho\): nestedness measured by the spectral radius of the adjacency
-matrix. \emph{maxtl}: maximum trophic level. \emph{diam}: network
-diameter. \emph{MxSim}: average maximum similarity between species
-pairs. \emph{Cannib}: proportion of cannibal species (self loops).
-\emph{Omniv}: proportion of omnivorous species. \emph{entropy}: SVD
+networks include the New Zealand and Tuesday Lake food webs having
+abundance data. Neutral: Neutral model of relative abundances. Null 1:
+Type I null model based on connectance. MaxEnt 1: Type I heuristic
+MaxEnt model based on connectance. Null 2: Type II null model based on
+the joint degree sequence. MaxEnt 2: Type II heuristic MaxEnt model
+based on the joint degree sequence. \(\rho\): nestedness measured by the
+spectral radius of the adjacency matrix. \(maxtl\): maximum trophic
+level. \(diam\): network diameter. \(MxSim\): average maximum similarity
+between species pairs. \(Cannib\): proportion of cannibal species (self
+loops). \(Omniv\): proportion of omnivorous species. \emph{entropy}: SVD
 entropy.}\tabularnewline
@@
 
+Finally, in fig.~\ref{fig:measures} we also show how well empirical
+measures are predicted using the type II MaxEnt network model. In
+accordance with our previous results, we found that nestedness was very
+well predicted by our model. However, the type II MaxEnt network model
+overestimated the maximum trophic level and network diameter, especially
+for networks with high empirical values. In fig.~S3, we show that the
+pairwise relationships between these four measures and species richness
+in empirical food webs are similar (in magnitude and sign) to the ones
+in type II MaxEnt networks.
+
+\begin{figure}
+\hypertarget{fig:measures}{%
+\centering
+\includegraphics{figures/measures_emp_maxent.png}
+\caption{Relationship between the structure of empirical and maximum
+entropy food webs. Empirical networks include most food webs archived on
+Mangal, as well as the New Zealand and Tuesday Lake datasets (our
+complete dataset). Maximum entropy networks were estimated using a
+simulated annealing algorithm to find the network of maximum SVD entropy
+while maintaining the joint degree sequence (type II MaxEnt network
+model). (a) Nestedness (estimated using the spectral radius of the
+adjacency matrix), (b) the maximum trophic level, (c) the network
+diameter, and (d) the SVD entropy were measured on these empirical and
+maximum entropy food webs. The identity line is plotted in each
+panel.}\label{fig:measures}
+}
+\end{figure}
+
+Moreover, we found that empirical networks are close to their maximum
+entropy considering a fixed joint degree sequence (fig.~S4). Empirical
+networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
+MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
+between the SVD entropy of empirical food webs and their maximum entropy
+configuration is plotted in the last panel of fig.~\ref{fig:measures}.
+We also found no relationship between the difference in SVD entropy
+between empirical and type II MaxEnt networks and species richness, the
+number of interactions, or connectance (fig.~S5). Similarly, we found no
+correlation between the difference in SVD entropy and the Jaccard
+distance of empirical and type II MaxEnt networks.
+
 The picture slightly changes when we consider another important property
@@
 which network structure is built upon and play a crucial role in
-community dynamics and assembly (Bastolla et al. 2009; Stouffer and
-Bascompte 2011). In fig.~\ref{fig:motifs}, we show that the motifs
-profile of networks generated using the type II MaxEnt network model was
-very close to the one of empirical data, and that they made better
-predictions than the type II null model. This is also shown in
-fig.~\ref{fig:motifs_rel}, where we see that relationships between
-motifs pairs of empirical food webs are very similar to the ones of the
-type II MaxEnt networks. Overall, these results suggest that the type II
-MaxEnt network model can reproduce many aspects of network structure.
-This highlights the importance of the joint degree sequence in shaping
-ecological networks.
+community dynamics and assembly (Stouffer and Bascompte 2011). In
+fig.~\ref{fig:motifs}, we show that the motifs profile of networks
+generated using the type II MaxEnt network model was very close to the
+one of empirical data, and that they made better predictions than the
+type II null model. This is also shown in fig.~\ref{fig:motifs_rel},
+where we see that relationships between motifs pairs of empirical food
+webs are very similar to the ones of the type II MaxEnt networks.
+Overall, these results suggest that the type II MaxEnt network model can
+reproduce many aspects of network structure. This highlights the
+importance of the joint degree sequence in shaping ecological networks.
 
@@
 (2007).}\label{fig:motifs_rel}
-}
-\end{figure}
-
-Moreover, we found that empirical networks are close to their maximum
-entropy considering a fixed joint degree sequence (fig.~S3). Empirical
-networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
-MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
-between the SVD entropy of empirical food webs and their maximum entropy
-configuration is plotted in the last panel of fig.~\ref{fig:measures}.
-We also found no relationship between the difference in SVD entropy
-between empirical and type II MaxEnt networks and species richness, the
-number of interactions, or connectance (fig.~S4). Similarly, we found no
-correlation between the difference in SVD entropy and the Jaccard
-distance of empirical and type II MaxEnt networks.
-
-Finally, in fig.~\ref{fig:measures} we also show how well empirical
-measures are predicted using the type II MaxEnt network model. In
-accordance with our previous results, we found that nestedness was very
-well predicted by our model. However, the type II MaxEnt network model
-overestimated the maximum trophic level and network diameter, especially
-for networks with high empirical values. In fig.~S5, we show that the
-pairwise relationships between these four measures and species richness
-in empirical food webs are similar (in magnitude and sign) to the ones
-in type II MaxEnt networks.
-
-\begin{figure}
-\hypertarget{fig:measures}{%
-\centering
-\includegraphics{figures/measures_emp_maxent.png}
-\caption{Relationship between the structure of empirical and maximum
-entropy food webs. Empirical networks include most food webs archived on
-Mangal, as well as the New Zealand and Tuesday Lake datasets (our
-complete dataset). Maximum entropy networks were estimated using a
-simulated annealing algorithm to find the network of maximum SVD entropy
-while maintaining the joint degree sequence (type II MaxEnt network
-model). (a) Nestedness (estimated using the spectral radius of the
-adjacency matrix), (b) the maximum trophic level, (c) the network
-diameter, and (d) the SVD entropy were measured on these empirical and
-maximum entropy food webs. The identity line is plotted in each
-panel.}\label{fig:measures}
 }
@@
 
+\leavevmode\hypertarget{ref-Bartomeus2016ComFra}{}%
+Bartomeus, Ignasi, Dominique Gravel, Jason M. Tylianakis, Marcelo A.
+Aizen, Ian A. Dickie, and Maud Bernard-Verdier. 2016. {``A Common
+Framework for Identifying Linkage Rules Across Different Types of
+Interactions.''} \emph{Functional Ecology} 30 (12): 1894--1903.
+\url{https://doi.org/10.1111/1365-2435.12666}.
+
 \leavevmode\hypertarget{ref-Bascompte2003NesAssa}{}%
@@
 
-\leavevmode\hypertarget{ref-Bastolla2009ArcMut}{}%
-Bastolla, Ugo, Miguel A. Fortuna, Alberto Pascual-Garcia, Antonio
-Ferrera, Bartolo Luque, and Jordi Bascompte. 2009. {``The Architecture
-of Mutualistic Networks Minimizes Competition and Increases
-Biodiversity.''} \emph{Nature} 458 (7241): 1018--U91.
-\url{https://doi.org/10.1038/nature07950}.
-
 \leavevmode\hypertarget{ref-Beck2009GenInf}{}%
@@
 \url{https://doi.org/10.1080/00107510902823517}.
+
+\leavevmode\hypertarget{ref-Canard2014EmpEva}{}%
+Canard, E. F., N. Mouquet, D. Mouillot, M. Stanko, D. Miklisova, and D.
+Gravel. 2014. {``Empirical Evaluation of Neutral Interactions in
+Host-Parasite Networks.''} \emph{The American Naturalist} 183 (4):
+468--79. \url{https://doi.org/10.1086/675363}.
 
@@
 
+\leavevmode\hypertarget{ref-McCann2011FooWeb}{}%
+McCann, Kevin S. 2011. \emph{Food Webs (MPB-50)}. \emph{Food Webs
+(MPB-50)}. Princeton University Press.
+\url{https://doi.org/10.1515/9781400840687}.
+
 \leavevmode\hypertarget{ref-Milo2002NetMot}{}%
@@
 \url{https://doi.org/10.7717/peerj.251}.
+
+\leavevmode\hypertarget{ref-Poisot2015SpeWhya}{}%
+Poisot, Timothée, Daniel B. Stouffer, and Dominique Gravel. 2015.
+{``Beyond Species: Why Ecological Interaction Networks Vary Through
+Space and Time.''} \emph{Oikos} 124 (3): 243--51.
+\url{https://doi.org/10.1111/oik.01719}.
+
+\leavevmode\hypertarget{ref-Pomeranz2019InfPrea}{}%
+Pomeranz, Justin P. F., Ross M. Thompson, Timothée Poisot, and Jon S.
+Harding. 2019. {``Inferring Predatorprey Interactions in Food Webs.''}
+\emph{Methods in Ecology and Evolution} 10 (3): 356--67.
+\url{https://doi.org/10.1111/2041-210X.13125}.
 

### 2022-08-12 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@e6d22e1ebae71c288345b90ff61ad0c0da5a02be 🚀

@@
 (Albert and Barabasi 2002). Further, cannibal species are species that
-can eat members of their own species (i.e.~species having self loops),
-whereas omnivorous species can prey on different trophic levels
+can eat individuals of their own species (i.e.~species having self
+loops), whereas omnivorous species can prey on different trophic levels
 (Williams and Martinez 2000). Finally, a motifs profile represents the
@@
 diameter, the type II heuristic MaxEnt model was better at predicting
-network structure than the neutral model for all measures considered.
+network structure than the neutral model for most measures considered.
 This might be because, although neutral processes are important, they
@@
 predators a species has is determined by its relative abundance and
-biological traits. Therefore, these results show that having information
-on the number of prey and predators for each species substantially
-improves the prediction of food-web structure, both compared to models
-solely based on connectance and the ones solely based on species
-relative abundances.
+biological traits. These results thus show that having information on
+the number of prey and predators for each species substantially improves
+the prediction of food-web structure, both compared to models solely
+based on connectance and to the ones solely based on species relative
+abundances.
 
@@
 (\(0.62 \pm 0.08\)) than its corresponding null model
-(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for all
-networks in our complete dataset (tbl.~\ref{tbl:measures_all}). This
-might in part be due to the fact that nestedness was calculated using
-the spectral radius of the adjacency matrix, which directly leverages
-information on the network itself just like the heuristic MaxEnt models.
+(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for networks
+in our complete dataset (tbl.~\ref{tbl:measures_all}). This might in
+part be due to the fact that nestedness was calculated using the
+spectral radius of the adjacency matrix, which directly leverages
+information on the network itself just like the heuristic MaxEnt model.
 The proportion of self-loops (cannibal species) was also better
@@
 proportion of omnivorous species were similar between both types of
-models. This suggests that, although the joint degree sequence is an
-important driver of food-web structure, other ecological constraints
-might need to be added to our MaxEnt model to account for some emerging
-food-web properties, especially the ones regarding food-chain lengths.
+models. We believe that this is because increasing the complexity of a
+food web might increase its average and maximum food-chain lengths. In
+comparison, the null model was more stochastic and does not necessarily
+produce more complex food webs with longer food-chain lengths.
 
@@
 
-Finally, in fig.~\ref{fig:measures} we also show how well empirical
-measures are predicted using the type II MaxEnt network model. In
-accordance with our previous results, we found that nestedness was very
-well predicted by our model. However, the type II MaxEnt network model
-overestimated the maximum trophic level and network diameter, especially
-for networks with high empirical values. In fig.~S3, we show that the
-pairwise relationships between these four measures and species richness
-in empirical food webs are similar (in magnitude and sign) to the ones
-in type II MaxEnt networks.
+Despite this increase in maximum trophic level and network diameter in
+MaxEnt food webs, we found that empirical food webs are close to their
+maximum entropy given a fixed joint degree sequence (fig.~S3). Empirical
+food webs in the complete dataset had an SVD entropy of
+\(0.89 \pm 0.04\), compared to an SVD entropy of \(0.94 \pm 0.03\) for
+networks generated using the type II heuristic MaxEnt model. The
+relationship between the SVD entropy of empirical food webs and their
+maximum entropy is plotted in the last panel of fig.~\ref{fig:measures}.
+As expected, the SVD entropy of maximum entropy food webs was higher
+than that of empirical food webs for almost all networks, confirming
+that our method indeed generated more complex networks. Moreover, we
+found no to a weak relationship between the increase in SVD entropy and
+the number of species, the number of interactions, and connectance
+(fig.~S4). This suggests that the slight increase in entropy between
+empirical food webs and their maximum entropy configuration was
+sufficient to modify some of their properties, regardless of their
+number of species and their number of interactions.
+
+A direct comparison of the structure of maximum entropy food webs,
+constrained by the joint degree sequence, with empirical data also
+supports the results depicted in tbl.~\ref{tbl:measures_all}. Indeed, in
+fig.~\ref{fig:measures} we show how well empirical measures are
+predicted by the type II heuristic MaxEnt model. In accordance with our
+previous results, we found that nestedness was very well predicted by
+our model. However, the model overestimated the maximum trophic level
+and network diameter, especially when the sampled food web had
+intermediate values of these measures. In fig.~S5, we show that the
+pairwise relationships between the four measures in
+fig.~\ref{fig:measures} and species richness in empirical food webs are
+similar (in magnitude and sign) to the ones found in food webs generated
+using the type II heuristic MaxEnt model. This indicates that the number
+of species in the network does not seem to impact the ability of the
+model to reproduce food-web structure.
 
@@
 entropy food webs. Empirical networks include most food webs archived on
-Mangal, as well as the New Zealand and Tuesday Lake datasets (our
-complete dataset). Maximum entropy networks were estimated using a
-simulated annealing algorithm to find the network of maximum SVD entropy
-while maintaining the joint degree sequence (type II MaxEnt network
-model). (a) Nestedness (estimated using the spectral radius of the
-adjacency matrix), (b) the maximum trophic level, (c) the network
-diameter, and (d) the SVD entropy were measured on these empirical and
-maximum entropy food webs. The identity line is plotted in each
+Mangal, as well as the New Zealand and Tuesday Lake food webs (our
+complete dataset). Maximum entropy networks were obtained using the type
+II heuristic MaxEnt model based on the joint degree sequence. (a)
+Nestedness (estimated using the spectral radius of the adjacency
+matrix), (b) the maximum trophic level, (c) the network diameter, and
+(d) the SVD entropy were measured on these empirical and maximum entropy
+food webs. The identity line is plotted in each
 panel.}\label{fig:measures}
@@
 
-Moreover, we found that empirical networks are close to their maximum
-entropy considering a fixed joint degree sequence (fig.~S4). Empirical
-networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
-MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
-between the SVD entropy of empirical food webs and their maximum entropy
-configuration is plotted in the last panel of fig.~\ref{fig:measures}.
-We also found no relationship between the difference in SVD entropy
-between empirical and type II MaxEnt networks and species richness, the
-number of interactions, or connectance (fig.~S5). Similarly, we found no
-correlation between the difference in SVD entropy and the Jaccard
-distance of empirical and type II MaxEnt networks.
-
-The picture slightly changes when we consider another important property
-of ecological networks, i.e.~their motifs profile. We measured the
-proportion of three-species motifs, which can be considered as simple
-building blocks of ecological networks (Milo et al. 2002; Stouffer et
-al. 2007). Motifs are the backbone of complex ecological networks from
-which network structure is built upon and play a crucial role in
-community dynamics and assembly (Stouffer and Bascompte 2011). In
+Notwithstanding its difficulties in reproducing adequate measures of
+food-chain lengths, the type II heuristic MaxEnt model can predict
+surprisingly well the motifs profile. Motifs are the backbone of complex
+ecological networks from which network structure is built upon and play
+a crucial role in community dynamics and assembly (Stouffer and
+Bascompte 2011). For this reason, the motifs profile can act as an
+effective ecological constraint shaping species interactions networks,
+and thus constitute a substantial source of ecological information. In
 fig.~\ref{fig:motifs}, we show that the motifs profile of networks
-generated using the type II MaxEnt network model was very close to the
-one of empirical data, and that they made better predictions than the
-type II null model. This is also shown in fig.~\ref{fig:motifs_rel},
-where we see that relationships between motifs pairs of empirical food
-webs are very similar to the ones of the type II MaxEnt networks.
-Overall, these results suggest that the type II MaxEnt network model can
-reproduce many aspects of network structure. This highlights the
-importance of the joint degree sequence in shaping ecological networks.
+generated using the type II heuristic MaxEnt model accurately reproduced
+the one of empirical data. This model made significantly better
+predictions than the ones based on connectance and the type II null
+model based on the joint degree sequence. This is also shown in
+fig.~\ref{fig:motifs_rel}, where we see that the relationships between
+motifs proportions in empirical food webs are very similar to the ones
+in networks generated using the type II heuristic MaxEnt model. This is
+in contrast with the type I null and MaxEnt models based on connectance,
+which produced opposite relationships than what was observed
+empirically. Our findings thus suggest that increasing food-web
+complexity within a maximum entropy framework based on the joint degree
+sequence does not alter the proportion of three-species motifs, but
+might alter their position relative to one another.
 
@@
 \includegraphics{figures/motifs_distribution.png}
-\caption{Motifs profile of empirical, maximum entropy, and null food
-webs. Empirical networks include most food webs archived on Mangal, as
-well as the New Zealand and Tuesday Lake datasets (our complete
-dataset). Maximum entropy networks were estimated using a simulated
-annealing algorithm to find the network of maximum SVD entropy while
-maintaining the connectance (type I MaxEnt network model) or the joint
-degree sequence (type II MaxEnt network model). The predictions of the
-types I and II null models are also plotted. Boxplots display the median
-proportions of each motif (middle horizontal lines), as well as the
-first (bottom horizontal lines) and third (top horizontal lines)
-quartiles. Vertical lines encompass all data points that fall within
-\(1.5\) times the interquartile range from both quartiles, and dots are
-data points that fall outside this range. Motifs names are from Stouffer
-et al. (2007).}\label{fig:motifs}
+\caption{Motifs profile of empirical and predicted food webs. Empirical
+networks include most food webs archived on Mangal, as well as the New
+Zealand and Tuesday Lake datasets (our complete dataset). Null 1: Type I
+null model based on connectance. MaxEnt 1: Type I heuristic MaxEnt model
+based on connectance. Null 2: Type II null model based on the joint
+degree sequence. MaxEnt 2: Type II heuristic MaxEnt model based on the
+joint degree sequence. Boxplots display the median proportion of each
+motif (middle horizontal lines), as well as the first (bottom horizontal
+lines) and third (top horizontal lines) quartiles. Vertical lines
+encompass all data points that fall within \(1.5\) times the
+interquartile range from both quartiles, and dots are data points that
+fall outside this range. Motifs names are from Stouffer et al.
+(2007).}\label{fig:motifs}
 }
@@
 \includegraphics{figures/motifs_relations.png}
-\caption{Relationship between motifs proportions of empirical, maximum
-entropy, and null food webs. Empirical networks include most food webs
+\caption{Pairwise relationships between motifs proportions of empirical
+and predicted food webs. Empirical networks include most food webs
 archived on Mangal, as well as the New Zealand and Tuesday Lake datasets
-(our complete dataset). Maximum entropy networks were estimated using a
-simulated annealing algorithm to find the network of maximum SVD entropy
-while maintaining the connectance (type I MaxEnt network model) or the
-joint degree sequence (type II MaxEnt network model). The predictions of
-the types I and II null models are also plotted. Regression lines are
-plotted in each panel. Motifs names are from Stouffer et al.
+(our complete dataset). Null 1: Type I null model based on connectance.
+MaxEnt 1: Type I heuristic MaxEnt model based on connectance. Null 2:
+Type II null model based on the joint degree sequence. MaxEnt 2: Type II
+heuristic MaxEnt model based on the joint degree sequence. Regression
+lines are plotted in each panel. Motifs names are from Stouffer et al.
 (2007).}\label{fig:motifs_rel}
@@
 \end{figure}
+
+One of the challenges in implementing and validating a maximum entropy
+model is to discover where its predictions break down. The results
+depicted in tbl.~\ref{tbl:measures_all} and fig.~\ref{fig:measures} show
+that our type II heuristic MaxEnt model can capture many high-level
+properties of food webs, but does a poor job of capturing others. This
+suggests that, although the joint degree sequence is an important driver
+of food-web structure, other ecological constraints might be needed to
+account for some emerging food-web properties, especially the ones
+regarding food-chain lengths. Nevertheless, fig.~\ref{fig:motifs} and
+fig.~\ref{fig:motifs_rel} show that the model can reproduce surprisingly
+well the motifs profile, one of the most ecologically informative
+properties of food webs. This suggests that the emerging structure of
+food webs is mainly driven by their joint degree sequence, although
+higher-level properties might be needed to ensure that food-chain
+lengths fall within realistic values.
 

### 2022-08-12 suppmat.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@fa572cda4b73fa82f7f24e762cbdf281e286e661 🚀
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@e6d22e1ebae71c288345b90ff61ad0c0da5a02be 🚀

@@
 \includegraphics{figures/entropy_distribution.png}
-\caption{(a) Distribution of the SVD entropy of MaxEnt food webs (type
-II MaxEnt network model) and of empirical food webs. Empirical networks
-include most food webs archived on Mangal, as well as the New Zealand
-and Tuesday lake datasets (our complete dataset). Maximum entropy
-networks were derived using a simulating annealing algorithm to find the
-network of maximum SVD entropy while maintaining the joint degree
-sequence. (b) Distribution of z-scores of the SVD entropy of all
-empirical food webs. Z-scores were computed using the mean and standard
-deviation of the distribution of SVD entropy of MaxEnt food webs (type
-II MaxEnt network model). The dash line corresponds to the median
+\caption{(a) Distribution of the SVD entropy of empirical and maximum
+entropy food webs. Empirical networks include most food webs archived on
+Mangal, as well as the New Zealand and Tuesday Lake food webs (our
+complete dataset). Maximum entropy networks were obtained using the type
+II heuristic MaxEnt model based on the joint degree sequence. (b)
+Distribution of z-scores of the SVD entropy of all empirical food webs.
+Z-scores were computed using the mean and standard deviation of the
+distribution of SVD entropy of MaxEnt food webs (type II heuristic
+MaxEnt model). The dash line corresponds to the median
 z-score.}\label{fig:entropy_dist}
@@
 \includegraphics{figures/difference_entropy.png}
-\caption{Difference in SVD entropy between MaxEnt (type II MaxEnt
-network model) and empirical food webs as a function of (a) species
-richness, (b) the number of interactions, and (c) connectance. Empirical
-networks include most food webs archived on Mangal, as well as the New
-Zealand and Tuesday lake datasets (our complete dataset). Maximum
-entropy networks were derived using a simulating annealing algorithm to
-find the network of maximum SVD entropy while maintaining the joint
-degree sequence. Regression lines are plotted in each
+\caption{Difference in SVD entropy between maximum entropy and empirical
+food webs as a function of (a) species richness, (b) the number of
+interactions, and (c) connectance. Empirical networks include most food
+webs archived on Mangal, as well as the New Zealand and Tuesday Lake
+food webs (our complete dataset). Maximum entropy networks were obtained
+using the type II heuristic MaxEnt model based on the joint degree
+sequence. Regression lines are plotted in each
 panel.}\label{fig:entropy_size}
@@
 \includegraphics{figures/measures_richness.png}
-\caption{Structure of empirical and maximum entropy food webs (type II
-MaxEnt network model) as a function of species richness. Empirical
-networks include most food webs archived on Mangal, as well as the New
-Zealand and Tuesday lake datasets (our complete dataset). Maximum
-entropy networks were derived using a simulating annealing algorithm to
-find the network of maximum SVD entropy while maintaining the joint
-degree sequence. (a) Nestedness (estimated with the spectral radius of
-the adjacency matrix), (b) the maximum trophic level, (c) the network
-diameter (i.e., the longest shortest path between all species pairs),
-and (d) the SVD entropy were measured on these empirical and predicted
-food webs and plotted against species richness. Regression lines are
-plotted in each panel.}\label{fig:measures_richness}
+\caption{Structure of empirical and maximum entropy food webs as a
+function of species richness. Empirical networks include most food webs
+archived on Mangal, as well as the New Zealand and Tuesday Lake food
+webs (our complete dataset). Maximum entropy networks were obtained
+using the type II heuristic MaxEnt model based on the joint degree
+sequence. (a) Nestedness (estimated using the spectral radius of the
+adjacency matrix), (b) the maximum trophic level, (c) the network
+diameter, and (d) the SVD entropy were measured on these empirical and
+maximum entropy food webs and plotted against species richness.
+Regression lines are plotted in each
+panel.}\label{fig:measures_richness}
 }

### 2022-08-13 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@647f4ae1f1a69dd22182f068828fcd142e1f4afe 🚀

@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
 The principle of maximum entropy is a robust mathematical method of
-finding least-biased probability distributions that have some desired
-properties given by prior knowledge on a system. We first used MaxEnt to
-predict the joint degree distribution and the degree distribution of
-maximum entropy given known numbers of species and of interactions. We
-found that the resulting joint degree distributions were more symmetric
-than the ones of empirical food webs, which suggests that other
-constraints might be needed to improve those predictions. We also used
-MaxEnt to predict networks of maximum entropy with a specified
-structure. These networks are the most complex, or random, given the
-constraints used. Likewise, we found that knowledge of species richness
-and of the number of interactions were not sufficient to reproduce many
-aspects of network structure found in nature. However, a model based on
-the entire joint degree sequence, i.e.~on the number of prey and
-predators for each species, gave more convincing results. Our type II
-MaxEnt network model yielded better or similar predictions than the type
-II null model, also based on the joint degree sequence, for most
-measures considered including the motifs profile. This suggests that the
-joint degree sequence drives many aspects of network structure. Indeed,
-considering our findings that empirical networks are close to maximum
-entropy for a given joint degree sequence, our results suggest that
-food-web topology is configured almost entirely randomly around these
-marginal numbers of predators and prey.
+finding least-biased probability distributions that have some specified
+properties given by prior knowledge about a system. We first applied
+this conventional MaxEnt approach on food webs to predict species level
+properties, namely the joint degree distribution and the degree
+distribution of maximum entropy given known numbers of species and
+interactions. We found that the joint degree distributions of maximum
+entropy had a similar shape to the ones of empirical food webs in
+high-connectance systems. However, these MaxEnt distributions were more
+symmetric than the ones of empirical food webs when connectance was low,
+which suggests that other constraints might be needed to improve these
+predictions in low-connectance systems. Then, we used a slightly
+different approach that aimed at finding heuristically the network
+configuration with the highest SVD entropy, i.e.~whose vector of
+relative singular values has maximum entropy. This network of maximum
+entropy is the most complex, or random, given the specified structure.
+We found that the heuristic maximum entropy model based on connectance
+did not predict the structure of sampled food webs very well. However,
+the heuristic maximum entropy model based on the entire joint degree
+sequence, i.e.~on the number of prey and predators for each species,
+gave more convincing results. Indeed, this model reproduced food-web
+structure surprisingly well, including the highly informative motifs
+profile. Nevertheless, it was not able to predict realistic food-chain
+lengths.
 
@@
 distribution in shaping food-web structure. This echoes the work of
-Williams (2011), which shows using MaxEnt how the degree distribution
-can drive the structure of bipartite networks. Fortuna et al. (2010)
-also found that the degree distribution of ecological networks is
-associated with some of their emerging properties such as nestedness and
-modularity. Network ecologists tend to focus on several emerging
-properties of food webs (Delmas et al. 2019), although we show that
-food-web structure is very well predicted by the joint degree
-distribution. In the light of our results, there is a lot more
-ecological information in the difference between properties of empirical
-networks and MaxEnt predictions than in these properties themselves.
+Fortuna et al. (2010), who found that the degree distribution of
+ecological networks drive their emerging structure such as their
+nestedness and modularity. Network ecologists tend to focus on several
+measures of food webs when studying the ecological consequences of their
+structure (McCann 2011; Delmas et al. 2019). In fact, following Williams
+(2011), we believe there is a lot more ecological information in the
+deviation between these properties in empirical systems and in their
+maximum entropy configuration given a fixed joint degree sequence.
 
@@
 
-There are different ways to generate food webs using MaxEnt. In this
-contribution, we used a method based on simulated annealing to find the
-network configuration with the highest SVD entropy while fixing some
-aspects of its structure. Another technique, also based on simulated
-annealing, could start by generating a food web randomly with fixed
-numbers of species and interactions and calculating its joint degree
-distribution. Pairs of interactions could be swapped sequentially until
-we minimize the divergence between the calculated joint degree
-distribution and the one of maximum entropy obtained analytically. In
-that case, this is the entropy of the joint degree distribution that
-would be maximized, not the one of the network's topology. To a certain
-extent, this method would thus bridge the gap between the two approaches
-presented in this article (i.e.~analytical and heuristic MaxEnt models).
-More research is needed to compare the quality of different methods
-generating adjacency matrices of food webs using MaxEnt.
-
-Another method generating adjacency matrices are maximum entropy graph
-models, which predict a probability distribution of networks under soft
-or hard constraints (e.g., Park and Newman 2004; Cimini et al. 2019).
-Under hard constraints, every network with a non-zero probability
-exactly satisfies the constraints on its structure. This is in contrast
-with soft constraints, which require that networks satisfy them on
-average (i.e.~many networks with a non-zero probability do not have the
-exact structure set by the constraints). Maximum entropy graph models
-are helpful because they can provide probability distributions for many
-network properties, by measuring the structure of all adjacency matrices
-with a non-zero probability. However, we consider that our approach
-based on simulated annealing is more flexible and more computationally
-efficient. Indeed, many measures of food-web structure are hard to
-translate into mathematical constraints. Moreover, because food webs are
-directed networks with self-loops, it makes the mathematical derivation
+In this contribution, we used a method based on simulated annealing to
+find the network configuration with the highest SVD entropy while fixing
+some aspects of its structure. However, there are different ways to
+generate adjacency matrices using MaxEnt. Another technique, also based
+on simulated annealing, could begin by generating a food web randomly
+with fixed numbers of species and interactions and calculating its joint
+degree distribution. Pairs of interactions could then be swapped
+sequentially until we minimize the divergence between the calculated
+joint degree distribution and the one of maximum entropy obtained
+analytically. In that case, this is the entropy of the joint degree
+distribution that would be maximized, not the one of the network's
+topology. To a certain extent, this method would thus bridge the gap
+between the analytical and heuristic approaches presented in this
+article. More research is needed to compare the quality of different
+methods generating adjacency matrices of food webs using MaxEnt.
+
+Maximum entropy graph models are another type of methods that predict a
+distribution of adjacency matrices under soft or hard constraints (e.g.,
+Park and Newman 2004; Cimini et al. 2019). Under hard constraints, every
+network with a non-zero probability exactly satisfies the constraints on
+its structure. This is in contrast with soft constraints, which require
+that networks satisfy them on average (i.e.~many networks with a
+non-zero probability do not have the exact structure set by the
+constraints). Maximum entropy graph models are helpful because they can
+provide probability distributions for many network properties by
+measuring the structure of all adjacency matrices with a non-zero
+probability. However, we consider that our approach based on simulated
+annealing is more flexible and more computationally efficient. Indeed,
+many measures of food-web structure are hard to translate into
+mathematical constraints. Moreover, because food webs are directed
+networks that can have self-loops, it makes the mathematical derivation
 of maximum entropy graph models difficult. We believe that identifying
@@
 useful first step before attempting to derive the mathematical
-formulation of a maximum entropy graph model adapted to food webs.
-
-Even though we decided to work with point estimates, it is possible to
-have a (non MaxEnt) probability distribution of networks when working
-with the entire chains (or part of the chains).
+formulation of a maximum entropy graph model for food webs.
 
@@
 
-Our method and results could be used for different purposes. First, they
-could be used as first-order approximations of network structure when
-only state variables are known. This could prove useful when predicting
-network structure at large spatial scales, where few ecological
-information is known at that scale. Second, they could be used as
-informative priors in Bayesian analyses of the structure of ecological
-networks (e.g., Cirtwill et al. 2019). Third, they could be used to make
-better predictions of pairwise species interactions by constraining the
-space of feasible networks, as discussed in Strydom et al. (2021).
-Finally, they could be used as alternative null models of ecological
-networks to better understand ecological mechanisms driving food-web
-structure. In that case, our model might need to be slightly adapted to
-give a probability distribution of Boolean networks (in contrast with
-point estimates of maximum entropy networks).
-
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
+Our analytical and heuristic models can be applied for different
+purposes. First, they could be used to generate first-order
+approximations of a network's properties when state variables are known
+empirically. For example, knowing the number of species in an ecological
+community, we can predict its number of interactions using the flexible
+links model and then predict its joint degree distribution with minimal
+biases using the principle of maximum entropy. This could prove
+particularly useful when predicting network structure at large spatial
+scales, subdividing the study area into smaller communities (e.g., grid
+cells). Indeed, because species richness and other ecological data are
+increasingly abundant (e.g., Dickinson, Zuckerberg, and Bonter 2010),
+validated MaxEnt models can be used to respond to a wider range of
+macroecological questions regarding food webs.
+
+Second, our analytical model can be used to generate informative priors
+in Bayesian analyses of the structure of ecological networks (e.g.,
+Cirtwill et al. 2019). Indeed, the probability distribution of maximum
+entropy derived using MaxEnt can be used as a prior that can be
+constantly updated with novel data. For instance, if we know the number
+of species and the number of interactions, we can get the degree
+distribution of maximum entropy, as shown in this contribution. The
+degree distribution represents the probability that a species can
+interact (as a predator or a prey) with a number of other species. Data
+on species interactions can be used to update the prior degree
+distribution to generate a more accurate posterior distribution, thus
+improving our description and understanding of the system.
+
+Third, our analytical and heuristic models can be used to make better
+predictions of pairwise species interactions by constraining the space
+of feasible networks, as discussed in Strydom et al. (2021). In other
+words, we can use the network configuration and/or specific measures of
+food-web structure derived using MaxEnt to ensure that our predictions
+of interspecific interactions form feasible networks. This means that
+the probability that two species interact can be conditional on the
+network structure and on the probability of interactions of all other
+species pairs. For that purpose, MaxEnt can be used to predict network
+structure when other data is lacking.
+
+Finally, our analytical and heuristic models can be used as alternative
+null models of ecological networks to better understand and identify the
+ecological processes driving food-web structure. Indeed, these
+mechanisms can be better described when analyzing the deviation of
+empirical data from MaxEnt predictions. A strong deviation would
+indicate that ecological mechanisms not encodes in the statistical
+constraints is at play for the system at hand. If deviations are
+systematic, the maximum entropy model might need to be revised to
+include appropriate ecological constraints. This revision process helps
+us reflect on and identify what really constrains food-web structure.
+
+The principle of maximum entropy can thus be applied for both the
+prediction and understanding of natural systems. Therefore, the model
+interpretation depends on how we use it. It can be used as a baseline
+distribution to identify the ecological constraints organizing natural
+systems. It can also be used as predictions of ecological systems. This
+distinction between understanding and predicting is essential when using
+and interpreting MaxEnt models.
+
+\hypertarget{final-remarks}{%
+\subsection{Final remarks}\label{final-remarks}}
 
@@
 is to identify the set of state variables that best reproduce empirical
-data. We found that the numbers of prey and predators for each species
-are important state variables for the prediction of maximum entropy
-networks. However, our predictions overestimated some measures of
-network structure, especially the maximum trophic level and network
-diameter. Therefore, we should continue playing the ecological detective
-to find these other topological constraints that would improve the
-predictions of our MaxEnt network models.
+data. We found that the number of species and the number of interactions
+are important state variables for the prediction of the joint degree
+distribution. Similarly, we found that the numbers of prey and predators
+for each species in a food web are important state variables for the
+prediction of the network configuration. However, our predictions
+overestimated the symmetry of the joint degree distribution for our
+analytical model and the maximum trophic level and network diameter for
+our heuristic model. We should thus continue to play the ecological
+detective to find these other topological constraints that would improve
+the predictions of MaxEnt models and help us understand better what
+really drives food-web structure.
 
@@
 
+\leavevmode\hypertarget{ref-Dickinson2010CitSci}{}%
+Dickinson, Janis L., Benjamin Zuckerberg, and David N. Bonter. 2010.
+{``Citizen Science as an Ecological Research Tool: Challenges and
+Benefits.''} In \emph{Annual Review of Ecology, Evolution, and
+Systematics, Vol 41}, edited by D. J. Futuyma, H. B. Shafer, and D.
+Simberloff, 41:149--72. Palo Alto: Annual Reviews.
+\url{https://doi.org/10.1146/annurev-ecolsys-102209-144636}.
+
 \leavevmode\hypertarget{ref-Dunne2002NetStrb}{}%

### 2022-08-13 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@fe5f2b2b1a820012f671a6765f220776b569f59a 🚀

@@
 al. 2010), and network connectance has been shown to be an important
-driver of many other emerging network properties (Timothée Poisot and
-Gravel 2014). In light of these observations, it is difficult to assess
-whether attributed effects of given properties are the artifacts of
-other, perhaps simpler, measures.
+driver of many other emerging network properties (Poisot and Gravel
+2014). In light of these observations, it is difficult to assess whether
+attributed effects of given properties are the artifacts of other,
+perhaps simpler, measures.
 
@@
 possible species pairs (MacDonald, Banville, and Poisot 2020). As shown
-by Timothée Poisot and Gravel (2014), connectance itself can constrain
-different aspects of network structure such as the degree distribution
-(i.e.~the probability distribution of the number of interspecific
-interactions realized by a species). Other measures, such as the maximum
-trophic level, can also constrain the space of feasible networks.
+by Poisot and Gravel (2014), connectance itself can constrain different
+aspects of network structure such as the degree distribution (i.e.~the
+probability distribution of the number of interspecific interactions
+realized by a species). Other measures, such as the maximum trophic
+level, can also constrain the space of feasible networks.
 
@@
 without data on pairwise species interactions, a prevailing situation
-worldwide (Timothée Poisot et al. 2021).
+worldwide (Poisot et al. 2021).
 
@@
 United States). First, all food webs archived on \texttt{mangal.io}
-(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
-directly queried from the database (\(N = 235\)). Most ecological
-networks archived on Mangal are multilayer networks, i.e.~networks that
-describe different types of interactions. We kept all networks whose
-interactions were mainly of predation and herbivory types, and removed
-the largest network (\(S = 714\)) for computational efficiency reasons.
-Then, to this set we added food webs from two different sources: the New
-Zealand dataset (\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday
-Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
-complete dataset thus contained a total of \(257\) food webs. Of the New
-Zealand and Tuesday Lake datasets, \(19\) networks had data on species'
-relative abundances that were used in the neutral model (Box 2). These
-\(19\) food webs constitute our \emph{abundance dataset}, which is a
-subset of our complete dataset.
+(Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were directly
+queried from the database (\(N = 235\)). Most ecological networks
+archived on Mangal are multilayer networks, i.e.~networks that describe
+different types of interactions. We kept all networks whose interactions
+were mainly of predation and herbivory types, and removed the largest
+network (\(S = 714\)) for computational efficiency reasons. Then, to
+this set we added food webs from two different sources: the New Zealand
+dataset (\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday Lake
+dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete
+dataset thus contained a total of \(257\) food webs. Of the New Zealand
+and Tuesday Lake datasets, \(19\) networks had data on species' relative
+abundances that were used in the neutral model (Box 2). These \(19\)
+food webs constitute our \emph{abundance dataset}, which is a subset of
+our complete dataset.
 
@@
 models). This suggests that the predictive power of connectance might be
-more limited than what was previously suggested (Timothée Poisot and
-Gravel 2014). On the other hand, the neutral model of relative
-abundances was surprisingly good at predicting the maximum trophic level
-and the network diameter. However, with the exception of the network
-diameter, the type II heuristic MaxEnt model was better at predicting
-network structure than the neutral model for most measures considered.
-This might be because, although neutral processes are important, they
-act in concert with niche processes in determining species interactions
-Canard et al. (2014). The joint degree sequence encodes information on
-both neutral and niche processes because the number of prey and
-predators a species has is determined by its relative abundance and
-biological traits. These results thus show that having information on
-the number of prey and predators for each species substantially improves
-the prediction of food-web structure, both compared to models solely
-based on connectance and to the ones solely based on species relative
+more limited than what was previously suggested (Poisot and Gravel
+2014). On the other hand, the neutral model of relative abundances was
+surprisingly good at predicting the maximum trophic level and the
+network diameter. However, with the exception of the network diameter,
+the type II heuristic MaxEnt model was better at predicting network
+structure than the neutral model for most measures considered. This
+might be because, although neutral processes are important, they act in
+concert with niche processes in determining species interactions Canard
+et al. (2014). The joint degree sequence encodes information on both
+neutral and niche processes because the number of prey and predators a
+species has is determined by its relative abundance and biological
+traits. These results thus show that having information on the number of
+prey and predators for each species substantially improves the
+prediction of food-web structure, both compared to models solely based
+on connectance and to the ones solely based on species relative
 abundances.
@@
 \leavevmode\hypertarget{ref-Poisot2016ManMakb}{}%
-Poisot, Timothee, Benjamin Baiser, Jennifer A. Dunne, Sonia Kefi,
+Poisot, Timothée, Benjamin Baiser, Jennifer A. Dunne, Sonia Kefi,
 Francois Massol, Nicolas Mouquet, Tamara N. Romanuk, Daniel B. Stouffer,

### 2022-09-06 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@ffc342bcba49ab4c5e81a9637768acf724056a12 🚀

@@
 
-\hypertarget{data-and-code}{%
-\section{Data and code}\label{data-and-code}}
-
-We tested our analytical and heuristic MaxEnt models against open
-food-web data queried from three different sources and integrated into
-what we call our \emph{complete dataset}. These sources include (1)
-terrestrial and aquatic food webs sampled globally and archived on the
-ecological interactions database Mangal; (2) freshwater stream food webs
-from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
-United States). First, all food webs archived on \texttt{mangal.io}
-(Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were directly
-queried from the database (\(N = 235\)). Most ecological networks
-archived on Mangal are multilayer networks, i.e.~networks that describe
-different types of interactions. We kept all networks whose interactions
-were mainly of predation and herbivory types, and removed the largest
-network (\(S = 714\)) for computational efficiency reasons. Then, to
-this set we added food webs from two different sources: the New Zealand
-dataset (\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday Lake
-dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete
-dataset thus contained a total of \(257\) food webs. Of the New Zealand
-and Tuesday Lake datasets, \(19\) networks had data on species' relative
-abundances that were used in the neutral model (Box 2). These \(19\)
-food webs constitute our \emph{abundance dataset}, which is a subset of
-our complete dataset.
-
-All code and data to reproduce this article are available at the Open
-Science Framework. Data cleaning, simulations and analyses were
-conducted in Julia v1.6.1.
-
 \hypertarget{analytical-maximum-entropy-models}{%
@@
 distribution.
+
+We tested our analytical MaxEnt model against open food-web data queried
+from three different sources and integrated into what we call our
+\emph{complete dataset}. These sources include (1) terrestrial and
+aquatic food webs sampled globally and archived on the ecological
+interactions database Mangal; (2) freshwater stream food webs from New
+Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan, United
+States). First, all food webs archived on \texttt{mangal.io} (Poisot et
+al. 2016; Banville, Vissault, and Poisot 2021) were directly queried
+from the database (\(N = 235\)). Most ecological networks archived on
+Mangal are multilayer networks, i.e.~networks that describe different
+types of interactions. We kept all networks whose interactions were
+mainly of predation and herbivory types, and removed the largest network
+(\(S = 714\)) for computational efficiency reasons. Then, to this set we
+added food webs from two different sources: the New Zealand dataset
+(\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday Lake dataset
+(\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete dataset
+thus contained a total of \(257\) food webs. All code and data to
+reproduce this article are available at the Open Science Framework. Data
+cleaning, simulations and analyses were conducted in Julia v1.6.1.
 
@@
 We compared the performance of both heuristic MaxEnt models in
-predicting these measures to the one of the null
-(tbl.~\ref{tbl:measures_all}) and neutral models
-(tbl.~\ref{tbl:measures_abund}). Overall, we found that the models based
-on the joint degree sequence (i.e.~the type II null and heuristic MaxEnt
-models) reproduced the structure of empirical food webs much better than
-the ones based on connectance (i.e.~the type I null and heuristic MaxEnt
-models). This suggests that the predictive power of connectance might be
-more limited than what was previously suggested (Poisot and Gravel
-2014). On the other hand, the neutral model of relative abundances was
-surprisingly good at predicting the maximum trophic level and the
-network diameter. However, with the exception of the network diameter,
-the type II heuristic MaxEnt model was better at predicting network
-structure than the neutral model for most measures considered. This
-might be because, although neutral processes are important, they act in
-concert with niche processes in determining species interactions Canard
-et al. (2014). The joint degree sequence encodes information on both
-neutral and niche processes because the number of prey and predators a
-species has is determined by its relative abundance and biological
-traits. These results thus show that having information on the number of
-prey and predators for each species substantially improves the
-prediction of food-web structure, both compared to models solely based
-on connectance and to the ones solely based on species relative
-abundances.
+predicting these measures to the one of the null and neutral models (Box
+2). We conducted these comparisons using two different datasets: (1) our
+complete dataset including most food webs archived on Mangal, as well as
+all food webs in the New Zealand and Tuesday Lake datasets, and (2) our
+\emph{abundance dataset}, a subset of the complete dataset comprising
+all food webs having data on their species' relative abundances
+(\(N=19\)). Indeed, of the New Zealand and Tuesday Lake datasets, \(19\)
+networks had data on species' relative abundances that were used in the
+neutral model to better assess the performance of our heuristic models.
+We compared our models' predictions using these two datasets separately
+to minimize biases and to better represent food webs with abundance data
+(tbl.~\ref{tbl:measures_all} and tbl.~\ref{tbl:measures_abund}).
+
+Overall, we found that the models based on the joint degree sequence
+(i.e.~the type II null and heuristic MaxEnt models) reproduced the
+structure of empirical food webs much better than the ones based on
+connectance (i.e.~the type I null and heuristic MaxEnt models). This
+suggests that the predictive power of connectance might be more limited
+than what was previously suggested (Poisot and Gravel 2014). On the
+other hand, the neutral model of relative abundances was surprisingly
+good at predicting the maximum trophic level and the network diameter.
+However, with the exception of the network diameter, the type II
+heuristic MaxEnt model was better at predicting network structure than
+the neutral model for most measures considered. This might be because,
+although neutral processes are important, they act in concert with niche
+processes in determining species interactions Canard et al. (2014). The
+joint degree sequence encodes information on both neutral and niche
+processes because the number of prey and predators a species has is
+determined by its relative abundance and biological traits. These
+results thus show that having information on the number of prey and
+predators for each species substantially improves the prediction of
+food-web structure, both compared to models solely based on connectance
+and to the ones solely based on species relative abundances.
 

### 2022-10-05 code/02_clean_data.jl
✨ remove Mangal food webs with missing interactions

@@ -5,6 +5,11 @@
 # food webs archived on mangal.io (generated from 01_import_mangal_metadata.jl) 
 mangal_foodwebs = DataFrame(CSV.File(joinpath("data", "raw", "mangal", "mangal_foodwebs.csv")))
 N_mangal = network.(mangal_foodwebs.id)
+
+# remove food webs with missing values of interaction strength
+N_mangal = N_mangal[[!ismissing(Mangal.get_all_interactions(N_mangal[i])[1].strength) for i in 1:length(N_mangal)]]
+
+# convert to unipartite networks
 N_mangal = convert.(UnipartiteNetwork, N_mangal)
 N_mangal = UnipartiteNetwork.(N_mangal[i].edges for i in 1:length(N_mangal))
 
@@ -12,14 +17,14 @@ N_mangal = UnipartiteNetwork.(N_mangal[i].edges for i in 1:length(N_mangal))
 NZ_foodwebs = DataFrame.(CSV.File.(glob("*.csv", 
                                         joinpath("data", "raw", "new_zealand", "adjacency_matrices")),
                                         drop=[1])) # the first column is row names
-N_NZ = convert.(Matrix{Bool}, NZ_foodwebs)
+N_NZ = Matrix{Bool}.(NZ_foodwebs)
 N_NZ = UnipartiteNetwork.(N_NZ, names.(NZ_foodwebs))
 
 # Tuesday lake food webs
 tuesday_foodwebs = DataFrame.(CSV.File.(glob("*.csv", 
                                         joinpath("data", "raw", "tuesday_lake", "adjacency_matrices")),
                                         header=false)) # no column names here
-N_tuesday = convert.(Matrix{Bool}, tuesday_foodwebs)
+N_tuesday = Matrix{Bool}.(tuesday_foodwebs)
 N_tuesday = UnipartiteNetwork.(N_tuesday)
 
 # simplify all food webs
@@ -50,7 +55,7 @@ abundance_data_NZ(fw_name::String)
     fw_name: name of the food web (file name)
 Returns the density (no/m2) of all species in the simplified food web with abundance data
 """
-function abundance_data_NZ(fw_name::String)
+function abundance_data_NZ(fw_name)
   # filter abundance data for this food web and get species names
   abund_fw = abund_NZ[abund_NZ[!, :food_web] .== fw_name, :]
   abund_taxa = abund_fw[!, :taxa]
[…]

### 2022-10-05 code/01_import_mangal_metadata.jl
⬆️ adapt to latest version of julia

@@ -17,10 +17,10 @@ mangal_networks = DataFrame(fill(0, (number_of_networks, 5)),
                  [:id, :S, :L, :P, :H])
 
 global cursor = 1
-@progress "Paging networks" for page in 1:number_of_pages
+@showprogress "Paging networks" for page in 1:number_of_pages
     global cursor
     networks_in_page = Mangal.networks("count" => count_per_page, "page" => page-1)
-    @progress "Counting items" for current_network in networks_in_page
+    @showprogress "Counting items" for current_network in networks_in_page
         S = count(MangalNode, current_network)
         L = count(MangalInteraction, current_network)
         P = count(MangalInteraction, current_network, "type" => "predation")

### 2022-10-05 code/functions/MxSim.jl

@@ -6,26 +6,28 @@ Returns the average of all species’ largest similarity index
 function MxSim(N::UnipartiteNetwork)
       AJS_N = AJS(N) # Additive Jaccard similarity between all species pairs
       length_AJS_N = length(AJS_N)
-      max_AJS = []
-      try
-            # find the maximum similarity for every species
-            for i in 1:richness(N) 
-                  spi = []
-                  for j in 1:length_AJS_N
-                        if in(AJS_N[j][1])(species(N)[i])
-                              push!(spi, AJS_N[j][2])
+      if length_AJS_N == 0
+            return missing
+      else 
+            max_AJS = []
+            try
+                  # find the maximum similarity for every species
+                  for i in 1:richness(N) 
+                    spi = []
+                        for j in 1:length_AJS_N
+                              if in(AJS_N[j][1])(species(N)[i])
+                               push!(spi, AJS_N[j][2])
+                              end
+                        end
+                        if length(spi) > 0
+                         push!(max_AJS, maximum(spi)[1])
                         end
                   end
-                  if length(spi) > 0
-                        push!(max_AJS, maximum(spi)[1])
-                  end
+            catch
+                  return missing
             end
-      catch
-            return missing
+            return mean(max_AJS) # return average similarity index 
       end
-      return mean(max_AJS) # return average similarity index 
 end
-
-
 
 

### 2022-10-05 main.jl

@@ -55,6 +55,3 @@ include(joinpath("code", "02_clean_data.jl"))
 include(joinpath("code", "03_predict_networks.jl"))
 include(joinpath("code", "04_compute_measures.jl"))
 include(joinpath("code", "05_make_figures.jl"))
-
-
-Pkg.add(Pkg.PackageSpec(;name="EcologicalNetworks", version="0.5.2"))

### 2022-10-05 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@c82bda2d4fbf4257add2e0471edd1854dc7d2abc 🚀
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@5b2b1da7cfed18eccd4e53e14d47ee8a4f61b83e 🚀

@@
 thus contained a total of \(257\) food webs. All code and data to
-reproduce this article are available at the Open Science Framework. Data
-cleaning, simulations and analyses were conducted in Julia v1.6.1.
+reproduce this article are available at the Open Science Framework
+(OSF.IO/KT4GS). Data cleaning, simulations and analyses were conducted
+in Julia v1.8.0.
 

### 2022-12-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@88e24362304b6916bc5f1dd8a2e5a49d3b4d0d7c 🚀

@@
 nature of ecological networks, it is surprising that the principle of
-maximum entropy, a rigorous mathematical method designed for both the
-analysis and prediction of constrained systems, has been barely used in
-network ecology.
+maximum entropy, a mathematical method designed for both the analysis
+and prediction of constrained systems, has been barely used in network
+ecology.
 
@@
 of data, as well as identify the most important ecological processes
-shaping that system. In other words, MaxEnt is a valuable and
-underexploited tool in network ecology that allows us to better
-understand and predict the structure of ecological networks.
+shaping that system. In other words, MaxEnt can help us better
+understand and predict the structure of ecological networks worldwide.
 
@@
 
-The principle of maximum entropy is a robust mathematical method of
-finding least-biased probability distributions that have some specified
+The principle of maximum entropy is a mathematical method of finding
+least-biased probability distributions that have some specified
 properties given by prior knowledge about a system. We first applied

### 2022-12-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@5714511bbb7c5379a9a4f4146d00f8a17640f682 🚀

@@
 chain were always accepted, whereas they were accepted with a
-probability conditional to a decreasing temperature when lower. The
-final configuration with the highest SVD entropy among the four chains
-constitute the estimated maximum entropy configuration of a food web
-given the constraint used.
+probability conditional to a decreasing temperature and the difference
+in SVD entropy when lower. The final configuration with the highest SVD
+entropy among the four chains constitute the estimated maximum entropy
+configuration of a food web given the constraint used.
 

### 2022-12-15 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1910093eefdd152d601a6c5d0ae3a79e33e933bc 🚀

@@
 ecological processes driving food-web structure. Indeed, these
-mechanisms can be better described when analyzing the deviation of
+mechanisms can be better described when analysing the deviation of
 empirical data from MaxEnt predictions. A strong deviation would
-indicate that ecological mechanisms not encodes in the statistical
-constraints is at play for the system at hand. If deviations are
+indicate that ecological mechanisms not encoded in the statistical
+constraints are at play for the system at hand. If deviations are
 systematic, the maximum entropy model might need to be revised to
@@
 us reflect on and identify what really constrains food-web structure.
+However, it is important to note here that tangible ecological
+mechanisms cannot be directly inferred from statistical distributions
+(Warren II, Costa, and Bradford 2022). Instead, by identifying the
+constraints of a system and by analysing empirical deviations from
+maximum entropy predictions, MaxEnt can only help us redirect research
+efforts towards understanding the biological mechanisms behind these
+constraints.
 
@@
 
+\leavevmode\hypertarget{ref-WarrenII2022SeeSha}{}%
+Warren II, Robert J., James T. Costa, and Mark A. Bradford. 2022.
+{``Seeing Shapes in Clouds: The Fallacy of Deriving Ecological
+Hypotheses from Statistical Distributions.''} \emph{Oikos} 2022 (11):
+e09315. \url{https://doi.org/10.1111/oik.09315}.
+
 \leavevmode\hypertarget{ref-Williams2011BioMet}{}%

### 2023-01-13 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1fd7f0c201ea0f03820cb55e1349980d2ccbea5b 🚀

@@
 the principle of maximum entropy (MaxEnt) in network ecology is still
-rare. This is surprising considering that MaxEnt is a renowned and
-rigorous statistical tool precisely designed for understanding and
-predicting many different types of constrained systems. Precisely, this
-principle asserts that the least-biased probability distribution of a
-system's property, constrained by prior knowledge about that system, is
-the one with maximum information entropy. MaxEnt has been proven useful
-in many ecological modelling problems, but its application in food webs
-and other ecological networks is limited. Here we show how MaxEnt can be
+rare. This is surprising considering that MaxEnt is a statistical tool
+precisely designed for understanding and predicting many different types
+of constrained systems. Precisely, this principle asserts that the
+least-biased probability distribution of a system's property,
+constrained by prior knowledge about that system, is the one with
+maximum information entropy. MaxEnt has been proven useful in many
+ecological modelling problems, but its application in food webs and
+other ecological networks is limited. Here we show how MaxEnt can be
 used to derive many food-web properties both analytically and
